# Cartels and Collusion

## Learning goals
- Detect and quantify cartel behavior using screens, structural breaks, and damages models.
- Combine empirical indicators with documentary and whistleblower evidence.
- Estimate overcharge and pass-through with appropriate uncertainty.

## Core topics
- Screens: price variance/compression, Benford, procurement anomalies, rotation patterns.
- Break tests and structural changes around cartel start/stop or dawn raids.
- Damages: overcharge estimation, pass-through to intermediaries/consumers, interest.
- Leniency and cooperation evidence; how to align with data patterns.
- Workflow: (1) timeline and conduct map; (2) descriptive stats; (3) screens; (4) event/break tests; (5) overcharge and pass-through; (6) sensitivity and uncertainty; (7) reconciliation with documents and testimony.

### Timeline and conduct map
- Build a dated timeline from complaints, internal messages, and interviews; align suspected coordination start/stop and enforcement events (e.g., dawn raids, leniency filings).
- Trace conduct hypotheses (bid rotation, list price coordination, geographic allocation) to specific products/customers and periods.
- This anchors all empirical work and prevents fishing; every regression or screen should map back to a conduct hypothesis.

### Descriptives and screens
- Plot prices/quantities over time with costs; look for compression, regime shifts, and unnatural stability in volatile cost environments.
- Procurement: rank bids over time by vendor; count rotations and clustering around “convenient” numbers; map geography to detect allocation.
- Use screens as triage, not proof. They motivate deeper causal tests and inform which periods/firms to emphasize.
- Document any screens in line with accepted practice [@oecd_cartel_screens_2013] and flag data limitations (e.g., missing bidders, list vs. net prices).

### Event and break tests
- If there is an enforcement or whistleblower event, run event studies around the announcement/raid. Test both price levels and spreads.
- Structural break tests (e.g., Bai–Perron) on prices/margins can flag regime changes; pair with documents to argue plausibility.
- In markets with cost shocks, include costs to avoid attributing cost-driven breaks to cartel conduct.

### Overcharge and pass-through
- Baseline: compare cartel period to pre/post or to a yardstick market; include FE for products/customers and cost controls.
- Robustness: alternative functional forms, trimming outliers, placebo periods. Cluster appropriately (customer/product/firm) and consider randomization inference for small N.
- Pass-through: regress downstream prices on upstream prices/costs with lags; compute confidence intervals. Distinguish partial vs. full pass-through when sizing harm.
- Calibrate damages with transparency on assumptions; relate magnitudes to benchmarks from known cartels (e.g., lysine/DRAM) when persuasive.

### Leniency, documents, and triangulation
- Leniency statements and chats/emails can pinpoint start/stop dates and mechanisms (allocation rules, triggers), guiding event windows and sample selection.
- Reconcile any “clean” quantitative results with qualitative evidence: if models show no effect where documents admit coordination, revisit data coverage, product mapping, or timing.
- Conversely, if screens suggest collusion but evidence is weak, highlight data issues or alternative explanations (common costs, regulation).
- Cite leniency frameworks when discussing incentives and timing [@oecd_leniency_2015]; be explicit about what is corroborated by statements vs. inferred from data.

::: {.callout-note title="Method box"}
- Event study around raid/announcement.
- Regression-based overcharge with controls and FE; robustness to functional form.
- Pass-through estimation using cost shifters and dynamic lags.
:::

::: {.callout-note title="Method box: screens and variance"}
- Distribution screens for bids/prices/quantities (variance compression, spikes, Benford where appropriate).
- Procurement rotation checks (bid ordering, geographic allocation).
- Cost-pass-through logic as a diagnostic for coordinated pricing.
:::

::: {.callout-note title="Qualitative evidence"}
- Bid sheets, meeting minutes, messaging apps; corroboration with timing in data.
- Procurement narratives from buyers on rotation and cover bids.
- Expert synthesis: building a timeline that ties conduct to measurable outcomes.
:::

::: {.callout-note title="Case box: Illustrative matters"}
- Apple eBooks (US) — hub-and-spoke coordination and price fixing in digital distribution.
- Lysine and DRAM cartels — classic overcharge estimation and dawn-raid event analysis.
- Bid-rigging and geographic allocation cases (construction, public procurement; include JFTC/SAMR examples for contrast).
:::

::: {.callout-note title="Case box: South African enforcement highlights"}
- **Bread cartel (Competition Commission v. Pioneer Foods, Tiger Brands, and Premier, Tribunal case 15/CR/Feb07).** Investigators combined mill-level cost data with Stats SA micro-CPI series for standard loaves to show synchronized 30–40 cent price jumps following coordination meetings. Break tests flagged the January 2007 spike that misaligned with wheat cost trends, supporting an estimated overcharge of 7–10%. Settlement conditions paired a R250 million administrative penalty with a R150 million consumer price relief program and a R250 million agro-processing fund targeted at small bakeries.
- **Construction fast-track settlement (2013).** Ahead of the 2010 FIFA World Cup, 15 firms self-reported 300 rigged projects, enabling the Commission to reconstruct bid-rotation matrices from SANRAL and stadia tenders. Event windows around dawn raids showed 8–12% drops in winning bid values relative to engineers’ estimates, and the Tribunal confirmed R1.46 billion in penalties while preserving the data archive for future screening exercises in provincial procurement.
- **Fertilizer and ammonia markets (Competition Commission v. Sasol Chemical Industries, Yara, and Omnia).** Using export-parity benchmarks and plant-level variable cost data, the Commission demonstrated that domestic ammonia and urea prices averaged 25–30% above competitive levels between 2005 and 2009 despite spare capacity. The Tribunal accepted the overcharge estimates and the settlement imposed a R250 million penalty plus ring-fenced divestitures and supply commitments that opened the market to smaller blenders in South Africa and neighboring SACU countries.
:::

::: {.callout-note title="Code box: raid event study scaffold"}
```{r}
#| label: cartel-raid-event
#| message: false
#| warning: false
# df needs columns: date, price, cost, treated (0/1), firm, product
# library(fixest)
# df <- df |> mutate(post = date >= as.Date("YYYY-MM-DD"))
# feols(log(price) ~ post * treated + cost + month(date) | firm + product + year(date), cluster = ~firm, data = df)
```
:::

## Code box: simple break test on public price series
```{r}
#| label: cartel-break-gasoline
#| message: false
#| warning: false
#| fig-cap: "Structural break detection in gasoline prices"
#| fig-width: 6
#| fig-height: 4
library(fredr)
library(dplyr)
library(strucchange)
library(ggplot2)
source("../R/helpers.R")

fredr_set_key(Sys.getenv("FRED_API_KEY"))
gas <- fredr(series_id = "GASREGW", observation_start = as.Date("2015-01-01")) |>
  rename(price = value) |>
  mutate(week = row_number())

bp <- breakpoints(price ~ week, data = gas, h = 26)
gas$regime <- factor(cut(gas$week, c(0, bp$breakpoints, Inf)))

ggplot(gas, aes(date, price, color = regime)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Regular Gasoline Prices: Structural Break Illustration",
    subtitle = "FRED series GASREGW (weekly, US average)",
    x = NULL,
    y = "USD per gallon",
    color = "Regime"
  ) +
  scale_color_antitrust() +
  theme_antitrust() +
  guides(color = guide_legend(nrow = 1))
```
This uses a public FRED series to illustrate break detection; replace with case data (e.g., product-level transactions) for real matters.


::: {.callout-note title="Citations and comparative note"}
- Cite cartel overcharge methods and benchmarks (e.g., OECD reports, reference guides) and US/EU leniency frameworks (US DOJ leniency policy; EC leniency notice).
- When referencing case examples (e.g., lysine, auto parts, LIBOR), include case numbers and agency decisions; for bid-rigging, add JFTC/SAMR examples where relevant.
- For screens, cite empirical papers validating specific indicators in procurement or commodities markets.
:::
