# Cartels and Collusion

## Learning goals
Cartel enforcement hinges on weaving together quantitative screens, econometric estimates, and documentary evidence. This chapter walks through that workflow using established methodologies from the OECD and academic literature on cartel detection.

By the end you should be able to:

- Spot suspicious conduct using dashboards, variance/rotation screens, and early break tests.
- Tie raid/leniency events to structural changes in prices, margins, and quantities.
- Estimate overcharge and pass-through with transparent assumptions and uncertainty analysis.
- Triangulate econometrics with leniency statements, chats, and procurement narratives to build court-ready stories.

## Workflow roadmap
1. **Timeline & conduct map.** Build a dated chronology from complaints, internal messages, procurement milestones, and raids. Align suspected mechanisms (bid rotation, geographic allocation, price floors) with specific products and periods.  
2. **Descriptive diagnostics.** Plot prices, costs, and quantities; look for unnatural stability or compression.  
3. **Screens.** Run variance, digit, and rotation screens to triage markets. Treat them as prioritization tools, not dispositive proof.  
4. **Event/break tests.** Evaluate how prices or spreads change around raids, leniency announcements, or policy shifts.  
5. **Overcharge & pass-through.** Estimate counterfactual prices using panel regressions, yardsticks, or synthetic controls; propagate uncertainty.  
6. **Triangulation.** Reconcile quantitative results with leniency statements, buyer interviews, and messaging logs.  
7. **Damages & remedies.** Translate overcharge estimates into monetary relief or structural commitments, respecting jurisdictional rules (e.g., Competition Tribunal SA public-interest remedies).

## Timeline and conduct map
Document every known communication, meeting, or enforcement action in a single table (date, event, evidence source, hypothesis). Update it as new statements arrive. Every regression or screen should cite the row(s) it tests—this discipline prevents “data dredging” and makes testimony easier.

## Descriptives and screens
Start with `ggplot2` dashboards that overlay prices with cost indices, demand proxies, and competitor prices. Flag regimes where prices remain static despite volatile costs or where margins converge across firms.

- **Variance/dispersion screens:** Check price spreads, standard deviations, and coefficents of variation across firms or regions [@abrantes_mello_2010].  
- **Procurement rotation:** Rank bids chronologically to flag turn-taking, convenient price endings, or geographic allocations [@porter_zona_1993; @conley_decarolis_2016].  
- **Digit/Benford checks:** Use sparingly and only when invoice conventions support the assumptions.  
- **Correlation screens:** High correlations in supposedly independent bids can justify deeper probes [@harrington_2008].

Document screen logic following OECD guidance on cartel screens [@oecd_cartel_screens_2013] and note data limitations (missing bidders, net vs. list prices).

### Bid-rotation scaffold
```{r}
#| label: cartel-rotation
#| eval: false
library(dplyr)
library(ggplot2)

rotation <- bids_df |>
  group_by(bidder) |>
  summarise(
    wins = sum(rank == 1),
    avg_rank = mean(rank, na.rm = TRUE),
    rotation_sd = sd(rank, na.rm = TRUE)
  ) |>
  arrange(desc(wins))

ggplot(rotation, aes(x = reorder(bidder, wins), y = wins)) +
  geom_col(fill = "#0072B2") +
  coord_flip() +
  labs(title = "Bid wins by vendor", x = NULL, y = "Wins") +
  theme_antitrust()
```
Pair charts with tables showing consecutive wins, geographic patterns, or unexplained bid withdrawals.

### Bid-rotation network graph
Network graphs help visualize suspected coordination patterns by showing which bidders consistently "yield" to others. This is particularly useful for identifying systematic rotation schemes or hub-and-spoke arrangements.

```{r}
#| label: cartel-rotation-network
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8
#| fig-format: png
library(dplyr)
library(tidyr)
library(ggplot2)
library(igraph)
library(ggraph)

# Simulated bid data: tender_id, bidder, rank, bid_amount
# In practice, replace with actual procurement data
set.seed(123)
n_tenders <- 50
bidders <- c("Firm A", "Firm B", "Firm C", "Firm D", "Firm E")

bids <- expand.grid(
  tender_id = 1:n_tenders,
  bidder = bidders
) |>
  group_by(tender_id) |>
  mutate(
    # Simulate rotation: certain patterns are more likely
    base_bid = runif(n(), 100, 200),
    rotation_boost = case_when(
      tender_id %% 5 == 0 & bidder == "Firm A" ~ -20,
      tender_id %% 5 == 1 & bidder == "Firm B" ~ -20,
      tender_id %% 5 == 2 & bidder == "Firm C" ~ -20,
      tender_id %% 5 == 3 & bidder == "Firm D" ~ -20,
      tender_id %% 5 == 4 & bidder == "Firm E" ~ -20,
      TRUE ~ 0
    ),
    bid = base_bid + rotation_boost,
    rank = rank(bid)
  ) |>
  ungroup()

# Identify winners and runner-ups
winners <- bids |>
  filter(rank == 1) |>
  select(tender_id, winner = bidder)

runner_ups <- bids |>
  filter(rank == 2) |>
  select(tender_id, runner_up = bidder)

# Create edge list: who "yielded" to whom (runner-up -> winner)
edges <- inner_join(winners, runner_ups, by = "tender_id") |>
  count(runner_up, winner, name = "weight") |>
  filter(runner_up != winner)

# Create network graph
g <- graph_from_data_frame(edges, directed = TRUE)

# Plot using ggraph
ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = weight, alpha = weight),
                 arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                 end_cap = circle(3, 'mm'),
                 color = "#666666") +
  geom_node_point(size = 12, color = "#0072B2", alpha = 0.8) +
  geom_node_text(
    aes(label = name),
    color = "white",
    fontface = "bold",
    size = 4,
    family = "Arial"
  ) +
  scale_edge_width(range = c(0.5, 3)) +
  scale_edge_alpha(range = c(0.3, 0.9)) +
  labs(
    title = "Bid Rotation Network Graph",
    subtitle = "Arrows show which firms consistently submit 2nd-place 'cover bids' when others win",
    caption = "Edge width = frequency of runner-up pattern. Systematic rotation creates hub-spoke or circular patterns."
  ) +
  theme_graph() +
  theme(
    text = element_text(family = "Arial"),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "none"
  )

# Supplementary table: win rates and rotation metrics
rotation_metrics <- bids |>
  group_by(bidder) |>
  summarise(
    wins = sum(rank == 1),
    second_place = sum(rank == 2),
    avg_rank = mean(rank),
    rotation_index = wins / n_tenders
  ) |>
  arrange(desc(wins))

cat("\nRotation metrics by bidder:\n")
print(rotation_metrics, n = Inf)
```

**How to interpret this graph:**
- **Hub-spoke patterns**: If one firm is at the center with many arrows pointing to it, that firm may be the "ringleader" coordinating bids.
- **Circular patterns**: Arrows forming a circle (A→B→C→D→A) suggest systematic rotation where firms take turns winning.
- **Asymmetric flows**: If Firm A frequently comes second when Firm B wins, but not vice versa, this may indicate a hierarchical arrangement.
- **Edge weights**: Thicker arrows indicate more frequent patterns, which are more suspicious than occasional occurrences.

**Practical applications:**
- Combine with document evidence (emails, meeting logs) to validate rotation schemes.
- Compare network patterns across product lines or regions to identify geographic allocation.
- Use as demonstrative exhibit in tribunal presentations or expert reports.

Replace simulated data with actual procurement records (World Bank, TED, US DOT, Stats SA infrastructure tenders). Document bid amounts, dates, and project characteristics in `data/raw/procurement/` with provenance notes.

## Event and break tests
When raids or leniency filings occur, run event studies on prices, spreads, or quantities. Check both cartel participants and outsiders; in public procurement you can also evaluate engineers’ estimates or cost indices.

Structural break tests (Bai–Perron, CUSUM) highlight regime shifts. Always incorporate costs so you do not misattribute cost-driven changes to conduct. Present break plots next to relevant timeline entries (e.g., WhatsApp thread confirming a January 2019 meeting).

::: {.callout-note title="Code box: raid event study scaffold"}
```{r}
#| label: cartel-raid-event
#| message: false
#| warning: false
# df must include date, price, cost, treated (0/1), firm, product
# library(fixest)
# df <- df |> mutate(post = date >= as.Date("YYYY-MM-DD"))
# feols(log(price) ~ post * treated + cost + month(date) | firm + product + year(date),
#       cluster = ~firm, data = df)
```
:::

## Overcharge and pass-through
Pick a counterfactual that matches data availability:

- **Before/after:** Include product, customer, and time fixed effects plus cost controls.  
- **Difference-in-differences:** Compare cartel markets to unaffected regions or product classes; test pre-trends.  
- **Yardsticks:** Use similar markets (neighboring countries, regulated benchmarks) when local data are thin.  
- **Synthetic controls:** Helpful for single-market cartels; combine unaffected markets into a donor pool.

Robustness: alternative functional forms (log vs. level), trimming outliers, placebo periods, heteroskedasticity checks, and randomization inference when N is small. For pass-through, regress downstream prices on upstream costs with lags to quantify harm on intermediaries vs. end-users.

## Leniency, documents, and triangulation
Leniency statements, chats, and board minutes pin down conduct mechanisms (rotation order, price floors, trigger strategies). Use them to:

- Set regression windows and sample selections.  
- Validate that estimated start/stop dates match qualitative evidence.  
- Explain residuals—if econometrics show little effect where documents admit coordination, revisit product mapping or data coverage.  
- Clarify what is fact (documented) vs. inference (econometric patterns), citing [@oecd_leniency_2015] or agency policies.

::: {.callout-note title="Method box: econometric toolkit"}
**Event studies:** Evaluate raid/leniency impact on prices or spreads.  
**Overcharge regressions:** `feols(log(price) ~ cartel_period + cost + demand | product + customer + time)` with clustered SEs and placebo checks.  
**Pass-through:** Dynamic regressions or VECMs linking upstream and downstream prices to trace harm along the chain.
:::

::: {.callout-note title="Method box: screens & variance"}
**Distribution screens:** Variance compression, digit spikes, Benford where appropriate.  
**Rotation checks:** Consecutive wins, geographic allocation, or sequential ordering.  
**Cost-pass-through diagnostics:** Compare pass-through before/during suspected collusion; suppressed pass-through can corroborate coordination.
:::

::: {.callout-note title="Qualitative evidence"}
Treat qualitative materials as structured data:

- **Bid sheets & messaging apps:** Extract explicit allocation rules or fallback prices.  
- **Buyer interviews:** Document procurement timelines, awareness of cover bids, or complaints.  
- **Expert narratives:** Craft timelines tying every empirical result to specific documents.
:::

::: {.callout-note title="Case box: Illustrative matters"}
**Apple eBooks (US).** Hub-and-spoke coordination with agency-model contracts and SSNDQ logic.  
**Lysine & DRAM (US/EU).** Classic dawn-raid analyses (see Connor, 2007, "Price-Fixing Overcharges," American Antitrust Institute Working Paper; Levenstein & Suslow, 2006, "What Determines Cartel Success?," *Journal of Economic Literature* 44(1): 43-95).  
**Construction bid-rigging (EU, SA, JFTC).** Rotation matrices and dawn raids aligned with procurement archives.  
**Digital advertising (SAMR).** Algorithms and platform policies as qualitative evidence.
:::

::: {.callout-note title="Case box: South African enforcement highlights"}
- **Bread cartel (Tribunal case 15/CR/Feb07).** Stats SA CPI data plus mill-level costs revealed synchronized 30–40 cent jumps; overcharge estimated at 7–10% with R250 million penalties and consumer relief funds.  
- **Construction fast-track settlement (2013).** Self-reported matrices across 300 tenders showed rotation; event windows around raids captured 8–12% price drops relative to engineers’ estimates.  
- **Fertilizer/ammonia (Sasol/Yara/Omnia).** Export-parity benchmarks and plant-level variable-cost data showed 25–30% overcharges despite spare capacity, leading to penalties and divestitures.
:::

## Code box: structural break illustration
```{r}
#| label: cartel-break-gasoline
#| message: false
#| warning: false
#| fig-cap: "Structural break detection in gasoline prices"
#| fig-width: 6
#| fig-height: 4
library(fredr)
library(dplyr)
library(strucchange)
library(ggplot2)
source("../program/R/helpers.R")

fredr_set_key(Sys.getenv("FRED_API_KEY"))
gas <- fredr(series_id = "GASREGW", observation_start = as.Date("2015-01-01")) |>
  rename(price = value) |>
  mutate(week = row_number())

bp <- breakpoints(price ~ week, data = gas, h = 26)
gas$regime <- factor(cut(gas$week, c(0, bp$breakpoints, Inf)))

ggplot(gas, aes(date, price, color = regime)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Regular Gasoline Prices: Structural Break Illustration",
    subtitle = "FRED series GASREGW (weekly, US average)",
    x = NULL,
    y = "USD per gallon",
    color = "Regime"
  ) +
  scale_color_viridis_d() +
  theme_antitrust() +
  guides(color = guide_legend(nrow = 1))
```
Replace the public FRED data with product-level transactions to present in litigation.

::: {.callout-note title="Citations and comparative note"}
- Cite OECD screen guidance, FTC/DOJ econometrics speeches, and EC/JFTC leniency notices when defending methodology.  
- Reference cases (lysine, DRAM, auto parts, LIBOR) with docket numbers.  
- Flag jurisdictional differences: e.g., SAMR often requires data-driven screens, while US courts scrutinize documentary corroboration.
:::

## Looking ahead
Archive every screen, regression, and figure in `data/derived` with READMEs so the damages and litigation chapters can reuse them. Push generic templates (rotation indices, break plots, overcharge tables) into `chapters/13-empirical-appendix.qmd`. When you shift to the mergers chapter, carry over demand estimates, diversion ratios, or pass-through results—post-cartel mergers frequently face heightened scrutiny, and reusing these diagnostics speeds up the next phase.
