---
title: "Cartels and Collusion"
---

## Learning goals
- Detect and quantify cartel behavior using screens, structural breaks, and damages models.
- Combine empirical indicators with documentary and whistleblower evidence.
- Estimate overcharge and pass-through with appropriate uncertainty.

## Core topics
- Screens: price variance/compression, Benford, procurement anomalies, rotation patterns.
- Break tests and structural changes around cartel start/stop or dawn raids.
- Damages: overcharge estimation, pass-through to intermediaries/consumers, interest.
- Leniency and cooperation evidence; how to align with data patterns.
- Workflow: (1) timeline and conduct map; (2) descriptive stats; (3) screens; (4) event/break tests; (5) overcharge and pass-through; (6) sensitivity and uncertainty; (7) reconciliation with documents and testimony.

### Timeline and conduct map
- Build a dated timeline from complaints, internal messages, and interviews; align suspected coordination start/stop and enforcement events (e.g., dawn raids, leniency filings).
- Trace conduct hypotheses (bid rotation, list price coordination, geographic allocation) to specific products/customers and periods.
- This anchors all empirical work and prevents fishing; every regression or screen should map back to a conduct hypothesis.

### Descriptives and screens
- Plot prices/quantities over time with costs; look for compression, regime shifts, and unnatural stability in volatile cost environments.
- Procurement: rank bids over time by vendor; count rotations and clustering around “convenient” numbers; map geography to detect allocation.
- Use screens as triage, not proof. They motivate deeper causal tests and inform which periods/firms to emphasize.
- Document any screens in line with accepted practice [@oecd_cartel_screens_2013] and flag data limitations (e.g., missing bidders, list vs. net prices).

### Event and break tests
- If there is an enforcement or whistleblower event, run event studies around the announcement/raid. Test both price levels and spreads.
- Structural break tests (e.g., Bai–Perron) on prices/margins can flag regime changes; pair with documents to argue plausibility.
- In markets with cost shocks, include costs to avoid attributing cost-driven breaks to cartel conduct.

### Overcharge and pass-through
- Baseline: compare cartel period to pre/post or to a yardstick market; include FE for products/customers and cost controls.
- Robustness: alternative functional forms, trimming outliers, placebo periods. Cluster appropriately (customer/product/firm) and consider randomization inference for small N.
- Pass-through: regress downstream prices on upstream prices/costs with lags; compute confidence intervals. Distinguish partial vs. full pass-through when sizing harm.
- Calibrate damages with transparency on assumptions; relate magnitudes to benchmarks from known cartels (e.g., lysine/DRAM) when persuasive.

### Leniency, documents, and triangulation
- Leniency statements and chats/emails can pinpoint start/stop dates and mechanisms (allocation rules, triggers), guiding event windows and sample selection.
- Reconcile any “clean” quantitative results with qualitative evidence: if models show no effect where documents admit coordination, revisit data coverage, product mapping, or timing.
- Conversely, if screens suggest collusion but evidence is weak, highlight data issues or alternative explanations (common costs, regulation).
- Cite leniency frameworks when discussing incentives and timing [@oecd_leniency_2015]; be explicit about what is corroborated by statements vs. inferred from data.

::: {.callout-note title="Method box"}
- Event study around raid/announcement.
- Regression-based overcharge with controls and FE; robustness to functional form.
- Pass-through estimation using cost shifters and dynamic lags.
:::

::: {.callout-note title="Method box: screens and variance"}
- Distribution screens for bids/prices/quantities (variance compression, spikes, Benford where appropriate).
- Procurement rotation checks (bid ordering, geographic allocation).
- Cost-pass-through logic as a diagnostic for coordinated pricing.
:::

::: {.callout-note title="Qualitative evidence"}
- Bid sheets, meeting minutes, messaging apps; corroboration with timing in data.
- Procurement narratives from buyers on rotation and cover bids.
- Expert synthesis: building a timeline that ties conduct to measurable outcomes.
:::

::: {.callout-note title="Case box: Illustrative matters"}
- Apple eBooks (US) — hub-and-spoke coordination and price fixing in digital distribution.
- Lysine and DRAM cartels — classic overcharge estimation and dawn-raid event analysis.
- Bid-rigging and geographic allocation cases (construction, public procurement; include JFTC/SAMR examples for contrast).
:::

::: {.callout-note title="Code box: raid event study scaffold"}
```{r}
#| label: cartel-raid-event
#| message: false
#| warning: false
# df needs columns: date, price, cost, treated (0/1), firm, product
# library(fixest)
# df <- df |> mutate(post = date >= as.Date("YYYY-MM-DD"))
# feols(log(price) ~ post * treated + cost + month(date) | firm + product + year(date), cluster = ~firm, data = df)
```
:::

## Code box: simple break test on public price series
```{r}
#| label: cartel-break-gasoline
#| message: false
#| warning: false
library(fredr)
library(dplyr)
library(strucchange)
library(ggplot2)

fredr_set_key(Sys.getenv("FRED_API_KEY"))
gas <- fredr(series_id = "GASREGW", observation_start = as.Date("2015-01-01")) |>
  rename(price = value) |>
  mutate(week = row_number())

bp <- breakpoints(price ~ week, data = gas, h = 26)
gas$regime <- factor(cut(gas$week, c(0, bp$breakpoints, Inf)))

ggplot(gas, aes(date, price, color = regime)) +
  geom_line() +
  labs(title = "Regular Gasoline Prices: Structural Break Illustration",
       subtitle = "FRED series GASREGW (weekly, US average)",
       x = NULL, y = "USD per gallon") +
  theme_minimal()
```
This uses a public FRED series to illustrate break detection; replace with case data (e.g., product-level transactions) for real matters.


::: {.callout-note title="Citations and comparative note"}
- Cite cartel overcharge methods and benchmarks (e.g., OECD reports, reference guides) and US/EU leniency frameworks (US DOJ leniency policy; EC leniency notice).
- When referencing case examples (e.g., lysine, auto parts, LIBOR), include case numbers and agency decisions; for bid-rigging, add JFTC/SAMR examples where relevant.
- For screens, cite empirical papers validating specific indicators in procurement or commodities markets.
:::
