# Mergers {#sec-mergers}

A merger review pulls on most of the tools developed so far. Market definition (@sec-market-definition) sets the competitive boundaries. The IO toolkit (@sec-io-toolkit) supplies the simulation machinery. The coordinated effects framework from the cartel chapter (@sec-cartels) informs whether a merger facilitates tacit collusion. The qualitative evidence protocols from @sec-research-design guide how internal documents, customer testimony, and efficiency claims fit together.

This chapter walks through a merger review end to end: transaction overview, shares and concentration metrics, unilateral and coordinated effects, vertical theories, efficiency evaluation, and remedy design. Legal standards and analytical approaches differ across the major jurisdictions (US, EU, UK, and South Africa), and merger practitioners increasingly work in more than one of them at once.

## Learning goals
Merger reviews draw on market definition, IO modeling, qualitative evidence, and remedies. This chapter provides a practical workflow for assessing unilateral, coordinated, vertical, and public-interest effects across US, EU/UK, and South African jurisdictions, drawing on agency guidance [@doj_ftc_hmg_2023; @ec_hmg_2004; @cma_merger_assessment_2021].

By the end you should be able to:

- Assess unilateral and coordinated effects with descriptive and structural tools (UPP/GUPPI, logit simulations, retrospectives).
- Evaluate vertical/mixed theories (elimination of double marginalization (EDM) vs. foreclosure) and quantify vUPP.
- Integrate efficiencies and public-interest claims with evidence.
- Propose remedies tied explicitly to diagnosed harms.

## Core topics

This section covers the theories of harm and the workflow for evaluating each.

::: {.callout-note title="Merger Review Workflow"}
```
PHASE 1: SCREENING              PHASE 2: DEEP DIVE              PHASE 3: DECISION
       |                              |                              |
       v                              v                              v
+----------------+            +----------------+            +----------------+
| Transaction    |            | Unilateral     |            | Efficiency     |
| overview &     |            | effects:       |            | evaluation &   |
| market mapping |----------->| UPP/GUPPI +    |----------->| Remedy design  |
|                |            | Simulation     |            |                |
+----------------+            +----------------+            +----------------+
       |                              |                              |
       v                              v                              v
+----------------+            +----------------+            +----------------+
| Shares, HHI &  |            | Coordinated    |            | Decision:      |
| diversion      |            | effects:       |            | Clear / Block  |
| calculation    |            | Maverick +     |            | / Remedies     |
|                |            | symmetry       |            |                |
+----------------+            +----------------+            +----------------+
       |                              |
       v                              v
+----------------+            +----------------+
| Flag: Above    |            | Vertical/mixed |
| 1,800 HHI +    |            | vUPP vs. EDM   |
| 100 delta?     |            | analysis       |
+----------------+            +----------------+
```
**Agency timelines:** US HSR (30 days initial, 30+ days Phase II) | EU EUMR (25+90 working days) | UK CMA (40+24 weeks) | SA (60 business days)
:::

**Horizontal mergers** eliminate competition between direct rivals. The key questions are:

- **Unilateral effects**: Will the merged firm raise prices by internalizing competition between the merging parties? (Measured via diversion ratios, UPP/GUPPI, simulation)
- **Coordinated effects**: Will the merger make tacit coordination easier among remaining competitors? (Measured via maverick analysis, symmetry, transparency)

**Vertical and mixed mergers** combine firms at different supply chain levels or across related markets:

- **Efficiency gains**: Elimination of double marginalization (EDM) can lower prices
- **Foreclosure risks**: The merged firm may raise rivals' costs or deny access to inputs/customers

**The merger review workflow** proceeds through seven stages:

1. Transaction and market overview
2. Shares, diversion, and HHI calculations
3. UPP/GUPPI screens with margin data
4. Unilateral and coordinated effects analysis
5. Vertical/mixed theories (EDM vs. foreclosure)
6. Efficiencies evidence
7. Remedies and retrospectives

### Transaction overview and market structure
- Summarize products, geographies, overlaps, and timing; map to candidate relevant markets.
- Compute shares/HHI and diversion using shares, switching, or survey-based measures. Flag mavericks and fringe.
- Upstream/downstream definitions depend on the theory of harm: be explicit about the vertical chain and platform sides.
- Ground definitions in agency guidance (US Merger Guidelines [@doj_ftc_hmg_2023; @doj_ftc_hmg_2010], EC/CMA guidance [@ec_hmg_2004; @cma_merger_assessment_2021]) and adjust for platform/digital contexts.

**Practical tips:** keep a data inventory (@sec-empirical-appendix template) noting which datasets inform shares (transaction data, Nielsen panels, loyalty cards). Align product labels with the IO models you plan to run later so you avoid remapping midstream.

### Unilateral effects: UPP/GUPPI and sim
- Use UPP/GUPPI as a screen [@farrell_shapiro_2010_merger; @jaffe_weyl_2013]; report inputs (diversion, margins) transparently. Sensitivity to margin measurement and diversion estimates should be shown.
- For differentiated products, simulate with logit/nested-logit or BLP-lite [@nevo_2000; @berry_levinsohn_pakes_1995]; emphasize calibration choices (outside share, margin sources).
- Convey uncertainty: confidence intervals on diversion, ranges on margins, and alternative ownership assumptions [@weinberg_hosken_2013].
- For readers, tie back to case examples in your slides to show how margins and diversion were evidenced (transaction data, surveys, switching analyses).

#### Simple UPP/GUPPI calculation

**What UPP measures:** Upward Pricing Pressure (UPP) quantifies the merged firm's incentive to raise prices. When Product A raises its price post-merger, some customers switch to Product B (now owned by the same firm). Pre-merger, those customers were "lost" to a competitor. Post-merger, the merged firm captures that profit—the "diverted profit"—giving it an incentive to raise prices.

**The formula:** $\text{UPP}_A = D_{A \to B} \times (P_B - C_B) - E_A$, where $D_{A \to B}$ is the diversion ratio from product A to its merger partner B, $P_B - C_B$ is the *partner's* absolute margin (the profit earned on each diverted sale), and $E_A$ is the per-unit efficiency credit. UPP is in currency units. GUPPI is the index form, $\text{GUPPI}_A = D_{A \to B} \times M_B \times P_B / P_A$: the same diverted profit expressed as a share of A's own price. @sec-io-toolkit states the formulas in that index form; the two agree once you divide by $P_A$.

**What to look for:** UPP > 0 means the merger creates upward pricing pressure. GUPPI values above 5-10% typically trigger enhanced scrutiny, though thresholds vary by jurisdiction and context.

```{r}
#| label: merger-upp-example
#| message: false
#| warning: false
library(dplyr)

#' Calculate Upward Pricing Pressure (UPP) for a merger
#' UPP_A = D(A->B) x (P_B - C_B) - E_A: the diverted profit is valued at the
#' MERGER PARTNER's price and margin, not those of the product losing sales.
#' @param diversion Fraction of product A's lost sales diverted to partner B
#' @param partner_price Merger partner's price (P_B)
#' @param partner_margin Merger partner's Lerner margin (P_B - C_B)/P_B
#' @param efficiency Per-unit cost savings (currency units) credited to A
upp <- function(diversion, partner_price, partner_margin, efficiency = 0) {
  # partner_price * partner_margin = P_B - C_B = partner's profit per unit
  diversion * partner_price * partner_margin - efficiency
}

inputs <- tibble::tribble(
  ~pair, ~diversion, ~partner_price, ~partner_margin, ~efficiency,
  "A->B", 0.35, 11.50, 0.45, 0.50,
  "B->A", 0.28, 10.00, 0.40, 0.30
) |>
  mutate(upp = upp(diversion, partner_price, partner_margin, efficiency))

inputs
```
Highlight the evidence sources for diversion (switching matrices, conjoint surveys, clickstream data) and margins (cost accounting, P&L, expert testimony). Note that negative efficiencies shift UPP upward, so document synergy assumptions thoroughly.

#### Logit simulation: executable example

**Why simulation matters:** UPP/GUPPI screens provide first-order approximations, but they don't account for competitive response or demand curvature. Merger simulation uses a calibrated demand model to predict post-merger equilibrium prices, incorporating how rivals will react and how consumers substitute across products.

**What this example shows:** Using a simple 3-product market, we demonstrate how to:

1. Calibrate price sensitivity from pre-merger shares and margins
2. Compute diversion ratios implied by the demand model
3. Calculate UPP/GUPPI for the merging products
4. Predict post-merger price increases

**Key assumptions to scrutinize:** Results depend heavily on the outside share (customers who leave the market entirely), margin data quality, and the logit functional form. Always show sensitivity analysis. To turn UPP into a predicted price change, the example uses the standard first-order approximation: the price rise is roughly pass-through times UPP, and pass-through is about one-half under linear demand.

This example demonstrates merger simulation using a 3-product differentiated market. The logit demand model allows us to compute diversion ratios, own-price elasticities, and post-merger price effects.

```{r}
#| label: merger-sim-example
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6
library(dplyr)
library(tidyr)
library(ggplot2)
source("program/R/helpers.R")

# Toy market: 3 products, 2 firms (Firm A owns products 1 & 2, Firm B owns product 3)
# After merger: Combined firm owns all three products
products <- tibble::tribble(
  ~product,    ~firm,    ~price,  ~share,  ~margin,
  "Product 1", "Firm A",  10.00,   0.30,    0.40,

  "Product 2", "Firm A",  12.00,   0.25,    0.35,
  "Product 3", "Firm B",  11.00,   0.20,    0.38
) |>
  mutate(
    mc = price * (1 - margin),  # Marginal cost from margin
    outside_share = 1 - sum(share)
  )

# Step 1: Calibrate logit parameters
# Mean utility delta from shares: delta_j = log(s_j) - log(s_0)
s0 <- 1 - sum(products$share)  # Outside good share = 0.25
products <- products |>
  mutate(
    delta = log(share) - log(s0),
    # Calibrate alpha (price sensitivity) from FOC: margin = 1 / (alpha * price * (1 - s_j))
    # Rearranging: alpha = 1 / (margin * price * (1 - s_j))
    alpha_implied = 1 / (margin * price * (1 - share))
  )

# Use average alpha for simulation
alpha <- mean(products$alpha_implied)
cat("Calibrated price sensitivity (alpha):", round(alpha, 3), "\n\n")

# Step 2: Compute own-price elasticities and diversion ratios
products <- products |>
  mutate(
    own_elasticity = -alpha * price * (1 - share),
    # Diversion ratio from j to k: D_jk = s_k / (1 - s_j)
    diversion_to_1 = products$share[1] / (1 - share),
    diversion_to_2 = products$share[2] / (1 - share),
    diversion_to_3 = products$share[3] / (1 - share)
  )

# Display pre-merger diagnostics
cat("Pre-merger elasticities and diversion:\n")
products |>
  select(product, firm, price, share, margin, own_elasticity) |>
  mutate(across(c(share, margin), ~scales::percent(., accuracy = 0.1)),
         own_elasticity = round(own_elasticity, 2)) |>
  print()

# Diversion matrix
cat("\nDiversion matrix (row = source product, column = destination):\n")
diversion_matrix <- products |>
  select(product, diversion_to_1, diversion_to_2, diversion_to_3) |>
  mutate(across(starts_with("diversion"), ~round(., 3)))
print(diversion_matrix)

# Step 3: Compute UPP for merger of Firm A and Firm B
# UPP_j = sum over rival products k: D_jk * (P_k - MC_k)
# After merger, Product 3 internalizes diversion to Products 1 & 2

upp_results <- tibble(
  product = products$product,
  firm = products$firm,
  price = products$price,
  # UPP calculation
  upp = c(
    # Product 1: already owned with Product 2, gains internalization of Product 3
    products$share[3] / (1 - products$share[1]) * (products$price[3] - products$mc[3]),
    # Product 2: already owned with Product 1, gains internalization of Product 3
    products$share[3] / (1 - products$share[2]) * (products$price[3] - products$mc[3]),
    # Product 3: gains internalization of Products 1 & 2
    products$share[1] / (1 - products$share[3]) * (products$price[1] - products$mc[1]) +
    products$share[2] / (1 - products$share[3]) * (products$price[2] - products$mc[2])
  )
) |>
  mutate(
    guppi = upp / price,  # GUPPI as % of own price
    # First-order price effect: delta-P ~ pass-through x UPP.
    # Pass-through is ~0.5 under linear demand; state and vary this assumption.
    pass_through = 0.5,
    price_effect = pass_through * upp,
    price_effect_pct = price_effect / price
  )

cat("\nMerger price pressure (UPP/GUPPI):\n")
upp_results |>
  select(-pass_through) |>
  mutate(
    upp = scales::dollar(upp, accuracy = 0.01),
    guppi = scales::percent(guppi, accuracy = 0.1),
    price_effect = scales::dollar(price_effect, accuracy = 0.01),
    price_effect_pct = scales::percent(price_effect_pct, accuracy = 0.1)
  ) |>
  print()

# Step 4: Visualize results
# Waterfall showing price effect decomposition for Product 3 (acquired product).
# Each step is a computed UPP component times the 0.5 linear-demand pass-through;
# total bars (pre- and post-merger price) are drawn from zero, not stacked.
pass_through <- 0.5
step_p1 <- pass_through * products$share[1] / (1 - products$share[3]) *
  (products$price[1] - products$mc[1])
step_p2 <- pass_through * products$share[2] / (1 - products$share[3]) *
  (products$price[2] - products$mc[2])
pre_price3 <- products$price[3]
post_price3 <- pre_price3 + step_p1 + step_p2

waterfall_data <- tibble(
  component = c("Pre-merger price", "Diversion to Product 1",
                "Diversion to Product 2", "Post-merger price"),
  value = c(pre_price3, step_p1, step_p2, post_price3),
  start = c(0, pre_price3, pre_price3 + step_p1, 0),
  end   = c(pre_price3, pre_price3 + step_p1, post_price3, post_price3),
  type  = c("total", "increase", "increase", "total")
) |>
  mutate(component = factor(component, levels = component))

ggplot(waterfall_data) +
  geom_rect(aes(xmin = as.numeric(component) - 0.4,
                xmax = as.numeric(component) + 0.4,
                ymin = start, ymax = end, fill = type),
            color = "black", linewidth = 0.5) +
  geom_text(aes(x = as.numeric(component),
                y = (start + end) / 2,
                label = scales::dollar(value, accuracy = 0.01)),
            size = 4, fontface = "bold") +
  scale_fill_manual(values = c("total" = "#0072B2", "increase" = "#D55E00")) +
  scale_x_continuous(breaks = 1:4, labels = levels(waterfall_data$component)) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Merger Simulation: Price Effect Decomposition (Product 3)",
    subtitle = "First-order approximation: price rise = 0.5 x UPP (linear-demand pass-through)",
    x = NULL, y = "Price ($)",
    caption = "Based on logit demand calibration. No rival repricing is modeled; a full equilibrium simulation would add it."
  ) +
  theme_antitrust() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")

# Summary statistics
cat("\n--- MERGER SIMULATION SUMMARY ---\n")
cat(paste0("Market: 3 products, outside share = ", scales::percent(s0, accuracy = 0.1), "\n"))
cat(paste0("Calibrated alpha: ", round(alpha, 3), "\n"))
cat(paste0("Product 3 predicted price increase: ",
           scales::percent(upp_results$price_effect_pct[3], accuracy = 0.1), "\n"))
cat(paste0("Combined firm post-merger share: ",
           scales::percent(sum(products$share), accuracy = 0.1), "\n"))
```

**Key takeaways from this simulation:**

1. **Diversion ratios** determine how much pricing pressure the merger creates. Higher diversion between merging products = higher UPP.
2. **GUPPI** (Gross Upward Pricing Pressure Index) expresses UPP as a percentage of price—values above 5-10% typically warrant further scrutiny.
3. **Price effects** here use the first-order rule: price rise = pass-through x UPP, with pass-through of 0.5 implied by linear demand. Convex demand raises pass-through, and rival repricing feeds back; a full equilibrium simulation handles both.
4. **Calibration matters**: Results are sensitive to the outside share assumption and margin data quality.

Before presenting results, show calibration diagnostics: how well the model reproduces pre-merger shares, whether price elasticities fall in plausible ranges, and how sensitive predictions are to alternative marginal-cost assumptions.

::: {.callout-tip title="Running example: Simulating the AA/US Airways merger"}
The AA/US Airways merger (announced February 2013, completed December 2013) lets us calibrate the simulation framework above to a real transaction. The demand estimates from @sec-io-toolkit and the route-level market definitions from @sec-market-definition supply the key inputs.

**Calibration.** To simulate the merger on a given route, we need three ingredients: (1) carrier market shares (from DOT T-100 or DB1B data, as in the route-share heatmap), (2) an outside share representing passengers who would choose alternative transport (driving, rail) or forgo travel entirely---typically 20--40% on short-haul routes like DCA--LGA but much higher on long-haul routes with no driving substitute, and (3) carrier-level margins derived from DOT Form 41 cost reports, which provide operating expenses by carrier and stage length. Legacy carriers like AA and US Airways typically reported operating margins of 8--15% in this period, while low-cost carriers like Southwest operated at 12--18%.

**Where the DOJ focused.** The government's challenge centered on routes where the merging carriers had combined nonstop shares exceeding 50%, particularly at hub airports: DCA (where AA and US together held the vast majority of slots), LGA (a slot-controlled airport with significant overlap), ORD (an AA hub where US Airways offered connecting competition), and PHL (the primary US Airways hub). On routes like DCA--LGA, the merger would have combined the two largest carriers, raising HHI well above the 2,500 threshold of the structural presumption (under the 2010 Guidelines then in force; the 2023 Guidelines later lowered the presumption to 1,800 with a delta above 100).

**The remedy.** Rather than blocking the merger outright, the DOJ required divestitures of 104 slots (52 slot pairs) at DCA and 34 slots (17 slot pairs) at LGA to low-cost carriers, primarily Southwest Airlines, to preserve competitive entry on the most concentrated routes. The logic: if the harm came from eliminating head-to-head overlap at slot-constrained airports, transferring slots to an aggressive competitor could restore the lost competition. Whether the remedy succeeded is an empirical question that the retrospective tools from @sec-research-design can answer, and remedy design is taken up in @sec-regulation-remedies.
:::

### Coordinated effects
- Look for increased symmetry, transparency, or capacity alignment post-merger; include bidding/auction context where relevant.
- Event studies around merger announcements can show rivals' stock price reactions (coordination signal) but should be paired with real-world capacity/contract evidence.
- Document maverick roles and whether the transaction removes or disciplines them.

Useful evidence: internal documents describing "price umbrella" logic, third-party contracts showing increased transparency, and supply/demand data indicating higher capacity utilization or inventory visibility. Build a maverick profile (pricing aggressiveness, innovation track record) to show whether elimination meaningfully raises coordination risk.

#### Detecting mavericks: quantitative screen
A "maverick" is a firm whose pricing or capacity behavior disrupts tacit coordination. Quantitative screens can identify mavericks by measuring pricing volatility, capacity utilization patterns, and deviations from industry norms. The coefficient of variation (CV) in prices is a useful starting metric.

```{r}
#| label: maverick-detection
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
source("program/R/helpers.R")

# Simulated pricing data for 4 firms over 36 months
# Replace with actual transaction data, Nielsen panels, or industry pricing indices
set.seed(42)

pricing_data <- expand.grid(
  firm = c("Industry Avg", "Firm A (Target)", "Firm B", "Firm C"),
  month = seq.Date(as.Date("2022-01-01"), as.Date("2024-12-01"), by = "month")
) |>
  as_tibble() |>
  mutate(
    # Base price trend
    base_price = 100 + as.numeric(month - min(month)) / 30 * 5,
    # Firm-specific behavior
    price = case_when(
      # Target firm: aggressive pricing, high volatility, frequent promotions
      firm == "Firm A (Target)" ~ base_price * (1 + rnorm(n(), -0.05, 0.08)),
      # Industry average: stable pricing
      firm == "Industry Avg" ~ base_price * (1 + rnorm(n(), 0, 0.02)),
      # Firm B: follows industry
      firm == "Firm B" ~ base_price * (1 + rnorm(n(), 0.02, 0.03)),
      # Firm C: premium pricing, stable
      firm == "Firm C" ~ base_price * (1 + rnorm(n(), 0.08, 0.025))
    ),
    # Capacity utilization (mavericks often run fuller capacity)
    capacity_util = case_when(
      firm == "Firm A (Target)" ~ pmin(1, rnorm(n(), 0.92, 0.05)),
      firm == "Industry Avg" ~ rnorm(n(), 0.78, 0.04),
      firm == "Firm B" ~ rnorm(n(), 0.75, 0.05),
      firm == "Firm C" ~ rnorm(n(), 0.72, 0.06)
    )
  )

# Calculate coefficient of variation by firm
maverick_metrics <- pricing_data |>
  group_by(firm) |>
  summarize(
    mean_price = mean(price),
    sd_price = sd(price),
    cv_price = sd_price / mean_price * 100,  # CV as percentage
    mean_capacity = mean(capacity_util) * 100,
    price_vs_industry = mean(price) / mean(pricing_data$price[pricing_data$firm == "Industry Avg"]) - 1,
    .groups = "drop"
  ) |>
  mutate(
    maverick_score = cv_price * (1 - price_vs_industry) * (mean_capacity / 80),
    maverick_flag = maverick_score > quantile(maverick_score, 0.75)
  )

# Print summary
cat("Maverick Detection Metrics:\n")
cat("Higher CV + Lower prices + Higher capacity = Maverick signal\n\n")
maverick_metrics |>
  mutate(
    cv_price = paste0(round(cv_price, 1), "%"),
    mean_capacity = paste0(round(mean_capacity, 0), "%"),
    price_vs_industry = scales::percent(price_vs_industry, accuracy = 0.1),
    maverick_score = round(maverick_score, 2)
  ) |>
  select(firm, cv_price, price_vs_industry, mean_capacity, maverick_score, maverick_flag) |>
  print()

# Plot 1: Price time series
p1 <- ggplot(pricing_data, aes(x = month, y = price, color = firm)) +
  geom_line(aes(linewidth = firm == "Firm A (Target)"), alpha = 0.8) +
  scale_linewidth_manual(values = c("TRUE" = 1.2, "FALSE" = 0.6), guide = "none") +
  scale_color_manual(
    values = c("Industry Avg" = "#999999", "Firm A (Target)" = "#D55E00",
               "Firm B" = "#0072B2", "Firm C" = "#009E73")
  ) +
  labs(
    title = "Price Trends by Firm",
    subtitle = "Target firm (orange) shows higher volatility and lower average price",
    x = NULL, y = "Price Index", color = NULL
  ) +
  theme_antitrust() +
  theme(legend.position = "bottom")

# Plot 2: Price volatility (CV) comparison
p2 <- maverick_metrics |>
  mutate(highlight = firm == "Firm A (Target)") |>
  ggplot(aes(x = reorder(firm, cv_price), y = as.numeric(gsub("%", "", cv_price)),
             fill = highlight)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "gray40") +
  annotate("text", x = 0.5, y = 5, label = "Industry norm (~5%)",
           hjust = 0, vjust = -0.5, size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#D55E00", "FALSE" = "#0072B2"),
                    guide = "none") +
  labs(
    title = "Price Volatility (Coefficient of Variation)",
    subtitle = "Mavericks typically show CV > industry average",
    x = NULL, y = "CV (%)"
  ) +
  theme_antitrust()

# Plot 3: Capacity utilization distribution
p3 <- pricing_data |>
  ggplot(aes(x = capacity_util * 100, fill = firm)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(
    values = c("Industry Avg" = "#999999", "Firm A (Target)" = "#D55E00",
               "Firm B" = "#0072B2", "Firm C" = "#009E73")
  ) +
  labs(
    title = "Capacity Utilization Distribution",
    subtitle = "Mavericks often run higher capacity, undercutting on price",
    x = "Capacity Utilization (%)", y = "Density", fill = NULL
  ) +
  theme_antitrust() +
  theme(legend.position = "bottom")

# Combine
(p1) / (p2 | p3) + plot_annotation(
  title = "Maverick Detection Dashboard",
  subtitle = "Identifying disruptive competitors in coordinated effects analysis",
  caption = "Synthetic data for illustration. Replace with actual pricing and capacity data."
)
```

**Interpreting maverick screens:**

1. **Coefficient of Variation (CV)**: Mavericks typically show CV 1.5–3× the industry average. High volatility signals frequent promotions, aggressive price responses, or willingness to sacrifice short-term margins.

2. **Price positioning**: Mavericks often price 5–15% below industry average while maintaining or expanding volume. Calculate: $\text{Price Gap} = \frac{P_{\text{firm}} - P_{\text{industry}}}{P_{\text{industry}}}$

3. **Capacity utilization**: Firms running >85% capacity utilization despite lower prices may be pursuing volume-based strategies that disrupt coordination.

4. **Innovation/entry record**: Document the target's history of new product launches, geographic expansion, or entry into adjacent markets—qualitative factors that amplify quantitative signals.

::: {.callout-note title="Practitioner tip"}
Combine quantitative screens with qualitative evidence: internal documents referencing the target as a "disruptor" or "price leader," customer testimony about switching patterns, and board materials discussing competitive responses to the target. Courts and agencies look for convergence across evidence types.
:::

::: {.callout-tip title="Extended example: Hospital merger analysis"}
Hospital mergers are among the most actively litigated in US antitrust, with the FTC challenging roughly one per year. Commercial insurance prices can rise 20--40% after consummation of anticompetitive hospital mergers, as documented in retrospective studies by Dafny (2009) and Gaynor and Town (2012). The analytical framework combines differentiated-products merger analysis (hospitals compete on quality, location, and reputation) with bilateral bargaining models, because hospitals negotiate rates with insurers rather than individual patients. The UPP/simulation toolkit from this chapter applies, with one wrinkle: the "price" is not set in a standard posted-price market.

Hospital "prices" emerge from bilateral negotiations between hospital systems and insurance networks. The key metric is the hospital's willingness-to-pay (WTP) value: how much an insurer's enrollees would pay to have the hospital included in their network. A merger that combines two hospitals with high WTP increases the merged system's bargaining leverage, because the insurer cannot drop both hospitals without losing a large share of its enrollees. The change in WTP (delta-WTP) serves as the hospital-market analog of UPP/GUPPI. It quantifies how much additional bargaining power the merger confers. The FTC has deployed WTP models in cases including *Advocate Health Care* (7th Cir. 2016), *Penn State Hershey/PinnacleHealth* (3rd Cir. 2016), and *Hackensack Meridian/Englewood* (2021), establishing WTP analysis as a standard tool in the agency's hospital merger playbook.

Hospital market definition is perennially contentious. Defendants typically argue for broad geographic markets on the grounds that patients can and do travel for care, while the FTC argues for narrower markets based on patient flow data showing that most patients travel less than 15--20 miles for inpatient services. The hypothetical monopolist framework adapts to this setting: instead of asking whether a price increase would be profitable, the relevant question is whether an insurer could market a viable plan excluding the merged system. Patient origin data, insurer testimony, and claims-level analysis of where patients actually receive care constitute the key evidence. The 2023 Merger Guidelines' emphasis on practical market realities, rather than theoretical substitution possibilities, has generally supported the FTC's narrower market definitions in recent hospital cases.

The South African Private Healthcare Market Inquiry (2014--2019) examined similar dynamics in a very different institutional setting. The Commission found HHIs above 4,000 in several provinces and documented limited patient switching despite tariff differentials across hospital groups. The inquiry's central recommendation---the creation of a Supply-Side Regulator for Health---represents a regulatory response to the same market power problem that the FTC addresses through case-by-case merger enforcement. The contrast is instructive: the US approach relies on ex ante merger control to prevent the accumulation of hospital market power, while South Africa's proposed approach would impose ongoing oversight of an already-concentrated sector. Cross-reference @sec-regulation-remedies for the remedy design implications.

For US hospital mergers, the principal data sources include CMS claims data (Medicare), FAIR Health (commercial claims), state all-payer claims databases (available in roughly 18 states), and the AHA Annual Survey for structural variables such as bed counts, service lines, and system affiliations. For South Africa, the Council for Medical Schemes, SAMRC databases, and provincial hospital utilization records provide analogous information. In both jurisdictions, the key challenge is obtaining commercially insured price data, which is often confidential and requires litigation discovery or regulatory compulsion to access. Budget time for data acquisition and cleaning. Hospital claims data is messy, with inconsistent coding and complex case-mix adjustment requirements.
:::

### Vertical and mixed effects

Vertical mergers combine firms at different levels of the supply chain. The core tension is between **efficiency gains** (elimination of double marginalization) and **foreclosure risks** (raising rivals' costs, input denial).

#### Elimination of Double Marginalization (EDM)

When both upstream and downstream firms have market power, each adds a markup, resulting in higher final prices than a vertically integrated firm would charge. Post-merger, the combined firm internalizes this externality.

```{r}
#| label: merger-edm
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
source("program/R/helpers.R")

# EDM calculation: compare pre- and post-merger prices
edm_example <- tibble::tribble(
  ~scenario, ~upstream_margin, ~downstream_margin, ~upstream_mc, ~description,
  "Pre-merger", 0.30, 0.25, 50, "Double markup: each firm adds margin",
  "Post-merger", 0.00, 0.25, 50, "EDM: upstream margin eliminated"
) |>
  mutate(
    # Pre-merger: downstream pays upstream price = MC / (1 - upstream_margin)
    upstream_price = upstream_mc / (1 - upstream_margin),
    # Final price includes both markups
    final_price = upstream_price / (1 - downstream_margin),
    # Consumer surplus proxy (inverse of price)
    consumer_benefit = 100 / final_price * 100
  )

cat("EDM Example:\n")
cat(paste0("Pre-merger final price: $", round(edm_example$final_price[1], 2), "\n"))
cat(paste0("Post-merger final price: $", round(edm_example$final_price[2], 2), "\n"))
cat(paste0("EDM savings: $", round(edm_example$final_price[1] - edm_example$final_price[2], 2),
           " (", round((1 - edm_example$final_price[2]/edm_example$final_price[1]) * 100, 1), "% reduction)\n"))
```

**Key parameters for EDM estimation:**
- **Upstream margin**: Often unobserved; infer from cost studies, comparable transactions, or bargaining models.
- **Pass-through rate**: What fraction of upstream cost savings reaches consumers? Typically 50-100% in competitive downstream markets, less if downstream is concentrated.
- **Scope of EDM**: Does EDM apply only to the merging downstream firm, or do rivals also benefit from lower input prices?

#### Vertical Upward Pricing Pressure (vUPP)

**The problem:** When a firm integrates upstream and downstream, it gains an incentive to raise its rivals' costs. Why? Because when a rival loses customers to the merged firm's downstream affiliate, the merged firm captures both the downstream profit *and* any margin it would have earned selling inputs to that rival.

**The intuition:** Imagine a content distributor (Netflix, Hulu) merges with a content studio. Post-merger, if the studio raises licensing prices for rival distributors, some subscribers switch to the merged distributor. The merged firm captures that profit. So it has an incentive to raise rivals' costs—even if it loses some upstream revenue—because it more than makes up for it downstream.

vUPP quantifies the merged firm's incentive to raise rivals' costs by increasing input prices or degrading input quality. The intuition: post-merger, the integrated firm captures a share of the profits when rivals lose sales to its downstream affiliate.

$$
\text{vUPP} = D_{RD} \times m_D \times P_D
$$

Where:
- $D_{RD}$: Diversion ratio from rival downstream firms to the merged downstream affiliate
- $m_D$: Downstream margin of the merged firm
- $P_D$: Downstream price

That expression is the *gross* benefit. The net screen subtracts the upstream margin the merged firm forgoes when it stops selling inputs to the rival---a margin earned per unit of *upstream input*, valued at the upstream price, not as a share of the downstream price.

```{r}
#| label: merger-vupp
#| message: false
#| warning: false

# vUPP screen following the Moresi-Salop vGUPPI logic. Both terms are dollars
# per unit of the rival's lost downstream sales: the merged firm gains the
# downstream margin on diverted sales and loses the upstream margin on the
# inputs it would have sold to produce that unit.
vupp_calc <- function(diversion_to_affiliate, downstream_margin, downstream_price,
                      upstream_margin, upstream_price, inputs_per_unit = 1) {
  # Benefit: captured downstream profit on diverted sales
  foreclosure_benefit <- diversion_to_affiliate * downstream_margin * downstream_price

  # Cost: forgone upstream margin on the inputs behind each lost rival unit
  foreclosure_cost <- inputs_per_unit * upstream_margin * upstream_price

  # Net vUPP
  net_vupp <- foreclosure_benefit - foreclosure_cost

  return(list(
    gross_vupp = foreclosure_benefit,
    opportunity_cost = foreclosure_cost,
    net_vupp = net_vupp
  ))
}

# Example: media/content vertical merger. Fixed proportions: one unit of
# licensed content input per subscriber served.
example <- vupp_calc(
  diversion_to_affiliate = 0.25,  # 25% of rivals' lost sales go to affiliate
  downstream_margin = 0.35,       # 35% downstream margin
  downstream_price = 100,         # $100 downstream price
  upstream_margin = 0.15,         # 15% margin on input sales to rival distributors
  upstream_price = 40             # $40 per unit of input sold to rivals
)

cat("vUPP Analysis (Input Foreclosure):\n")
cat(paste0("Gross vUPP (foreclosure benefit): $", round(example$gross_vupp, 2), "\n"))
cat(paste0("Opportunity cost (lost upstream margin): $", round(example$opportunity_cost, 2), "\n"))
cat(paste0("Net vUPP: $", round(example$net_vupp, 2), "\n"))
cat(paste0("Interpretation: ", ifelse(example$net_vupp > 0,
    "Positive net vUPP suggests incentive to foreclose rivals.",
    "Negative net vUPP suggests foreclosure is unprofitable."), "\n"))
```

The back-of-envelope keeps the bookkeeping honest, but it abstracts from two things a full vGUPPI handles: how much of the input price rise the rival passes through to its own customers, which determines how many downstream sales it actually loses, and any departure from fixed-proportions production. Moresi and Salop (2013) develop the complete set of vGUPPI indices for upstream, downstream, and customer foreclosure.

#### Balancing EDM vs. vUPP

The net effect of a vertical merger depends on whether EDM savings outweigh foreclosure harms. Present both calculations with sensitivity ranges:

```{r}
#| label: merger-vertical-balance
#| message: false
#| warning: false

# Create sensitivity matrix
scenarios <- expand.grid(
  edm_magnitude = seq(5, 25, by = 5),  # % price reduction from EDM
  vupp_effect = seq(2, 15, by = 3)      # % price increase to rivals from foreclosure
) |>
  mutate(
    net_effect = edm_magnitude - vupp_effect * 0.5,  # Weight foreclosure by rival share
    assessment = case_when(
      net_effect > 5 ~ "Likely procompetitive",
      net_effect > 0 ~ "Uncertain - needs detailed analysis",
      TRUE ~ "Likely anticompetitive"
    )
  )

ggplot(scenarios, aes(x = edm_magnitude, y = vupp_effect, fill = net_effect)) +
  geom_tile() +
  geom_text(aes(label = round(net_effect, 1)), size = 3, color = "white") +
  scale_fill_gradient2(low = "#D55E00", mid = "#F0E442", high = "#009E73",
                       midpoint = 0, name = "Net effect\n(% points)") +
  labs(
    title = "Vertical merger assessment: EDM vs. foreclosure",
    subtitle = "Net effect = EDM savings - (foreclosure harm × rival share)",
    x = "EDM magnitude (% price reduction)",
    y = "vUPP/foreclosure effect (% price increase to rivals)"
  ) +
  theme_antitrust()
```

**Data sources for vertical analysis:**
- **Input pricing**: Contracts, invoices, internal transfer pricing documents
- **Diversion estimates**: Customer surveys, switching data, or structural demand estimation
- **Margin data**: Cost accounting, segment financials, comparable transactions
- **Raising rivals' costs evidence**: Internal strategy documents discussing input denial, quality degradation, or discriminatory access

See @salop_2018 for the theoretical framework. On agency guidance, read the 2020 Vertical Merger Guidelines [@doj_ftc_vmg_2020] as history rather than current policy: the FTC withdrew them in 2021, and the 2023 Merger Guidelines [@doj_ftc_hmg_2023] now carry the agencies' approach to vertical theories of harm.

### Efficiencies and remedies
- Synergy claims: require verifiable, merger-specific efficiencies with timelines and implementation costs; stress test with sensitivity tables.
- Remedies: structural first; behavioral only if verifiable/monitorable. Link proposed remedies to modeled harms and operational feasibility.

Tie efficiencies to data. For example, if parties cite procurement savings, request SKU-level cost projections and simulate whether those savings offset UPP. For behavioral remedies, document monitoring costs and fallback options (trustees, data rooms) referenced in CMA/DG COMP practice.

### Retrospectives
- Where historical analogs exist, run diff-in-diff/event studies on prices/output/quality. Use rivals and unaffected markets as controls; test pre-trends [@ashenfelter_hosken_2010; @miller_weinberg_2017].
- For platform/vertical cases, examine participation/multi-homing effects and access terms over time.
- Cite retrospective literature to benchmark magnitudes and methods [@weinberg_hosken_2013].
- Use well-known retrospectives (e.g., supermarket/hospital/airline cases) to set expectations on effect sizes and uncertainty.

#### Stock-event diagnostic
```{r}
#| label: merger-stock-event
#| message: false
#| warning: false
#| cache: true
library(dplyr)
library(ggplot2)
library(readr)
library(lubridate)
source("program/R/helpers.R")

# Airline stock data pulled by program/scripts/02_finance_extended.R.
# Guarded read so the chapter renders from a fresh clone.
airlines <- tryCatch({
  read_csv("data/raw/finance_airlines.csv", show_col_types = FALSE) |>
    mutate(date = as.Date(date))
}, error = function(e) NULL)

if (!is.null(airlines)) {
  # American Airlines/US Airways merger was announced Feb 14, 2013.
  # The merging parties themselves cannot appear in this window: US Airways
  # (LCC) delisted at closing, American's pre-merger equity traded as AAMRQ
  # in bankruptcy, and the post-merger AAL ticker only begins trading in
  # December 2013. This is therefore a RIVALS-ONLY event study.
  event_date <- as.Date("2013-02-14")

  # Filter to rivals and event window, compute returns
  prices <- airlines |>
    filter(!symbol %in% c("AAL", "LCC", "AAMRQ")) |>
    filter(between(date, event_date - days(120), event_date + days(60))) |>
    group_by(symbol) |>
    arrange(date) |>
    mutate(ret = log(adjusted) - log(lag(adjusted))) |>
    ungroup() |>
    filter(!is.na(ret))

  # Market model: use SPY as benchmark
  market <- prices |> filter(symbol == "SPY") |> transmute(date, mkt_ret = ret)
  cars <- prices |> filter(symbol != "SPY") |>
    left_join(market, by = "date") |>
    filter(!is.na(mkt_ret)) |>
    mutate(abnormal = ret - mkt_ret, rel_day = as.integer(date - event_date)) |>
    filter(between(rel_day, -20, 20)) |>
    group_by(symbol) |>
    arrange(rel_day) |>
    mutate(car = cumsum(abnormal))

  # Label mapping for rival airlines
  airline_labels <- c("DAL" = "Delta", "UAL" = "United", "LUV" = "Southwest",
                      "JBLU" = "JetBlue", "ALK" = "Alaska")

  p_cars <- ggplot(cars |> mutate(label = airline_labels[symbol]),
         aes(rel_day, car, color = label)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "darkred", linewidth = 0.8) +
    geom_hline(yintercept = 0, linewidth = 0.3, color = "gray70") +
    geom_line(linewidth = 0.9) +
    scale_color_manual(values = c(
      "Delta" = "#0072B2",
      "United" = "#009E73",
      "Southwest" = "#CC79A7",
      "JetBlue" = "#E69F00",
      "Alaska" = "#56B4E9"
    )) +
    labs(title = "Rival Airlines' Cumulative Abnormal Returns: AA/US Airways Announcement",
         subtitle = "Rivals only: the merging parties' tickers (LCC, AAMRQ) are delisted. Announced Feb 14, 2013; window ±20 trading days.",
         x = "Event time (days)", y = "CAR", color = NULL,
         caption = "Source: finance_airlines.csv (tidyquant). SPY used as market benchmark.\nPositive CAR for rivals suggests coordination expectations; negative suggests competitive threat.") +
    theme_antitrust() +
    theme(legend.position = "bottom")
  print(p_cars)

  # Summary table of cumulative abnormal returns
  car_summary <- cars |>
    filter(rel_day == max(rel_day)) |>
    mutate(label = airline_labels[symbol]) |>
    select(label, CAR = car, final_day = rel_day)

  cat("\nCumulative Abnormal Returns (full event window):\n")
  print(knitr::kable(car_summary, digits = 4,
               caption = "CAR for rival airlines around AA/US Airways merger announcement"))
} else {
  cat("Airline stock data not available.",
      "Run program/scripts/02_finance_extended.R to build data/raw/finance_airlines.csv.\n")
}
```
Use CAR patterns as suggestive evidence of coordination or efficiency expectations, but always pair with operational data (capacity, contracts). The rivals-only design is a necessity, not a choice: standard data feeds do not serve the merging parties' delisted histories (US Airways traded as LCC until closing; American's pre-merger equity traded as AAMRQ in bankruptcy), so the announcement effect is read off Delta, United, Southwest, JetBlue, and Alaska. Positive abnormal returns for rival airlines on the announcement date suggest the market expects the merger to facilitate coordinated effects—an "umbrella effect" that benefits all carriers. Negative abnormal returns suggest the merged entity is expected to compete more aggressively.

### Southern African merger evidence
- **Walmart/Massmart (2011).** The Competition Commission and Tribunal analyzed SKU-level sales and procurement data showing Massmart’s 20–25% share in formal general merchandise with limited reach into township grocery segments. Diversion estimates from loyalty-card switching rates indicated minimal unilateral effect, so the case turned on public-interest harms. Conditions ultimately required a R240 million supplier development fund, a two-year moratorium on merger-specific retrenchments, and detailed annual reporting on local procurement shares—providing a template for tying data-backed public-interest claims to remedies.
- **Mediclinic/Matlosana (2014).** Using patient-level discharge data covering 24 specialties, the Commission computed local HHIs above 6,000 and estimated post-merger tariff increases of 8–12% for insured patients. The Tribunal accepted that rival hospitals were more than 150 km away and prohibited the deal, highlighting how granular utilization data can anchor both geographic market definition and competitive-effects narratives in middle-income regions.
- **Heineken/Distell/Capevin (2022).** In evaluating the creation of Newco, the Commission’s demand estimates—calibrated from Nielsen panel data—showed cider/RTD diversion ratios above 0.5 between Hunters, Savanna, and Strongbow, with Newco projected to command roughly 65% share. Conditional approval required a R10 billion investment commitment, maintenance of existing third-party distribution contracts, and shelf-space safeguards for smaller craft brands, illustrating how quantitative evidence on differentiated products fed into both competition and public-interest remedies.

::: {.callout-note title="Method box"}
- Simple merger sim template (see R helpers).
- Event study around merger announcement/close; rival effects.
- Vertical tools: vUPP, EDM, raising rivals’ costs sketches.
:::

::: {.callout-note title="Method box: UPP/GUPPI quick calc"}
See the UPP/GUPPI example above. Expand the template with case-specific diversion estimates (from surveys, loyalty data, or conjoint work) and margin sources (accounting or expert models). Always show sensitivity ranges.
:::

::: {.callout-note title="Qualitative evidence"}
- Integration plans, synergy decks, customer feedback on alternatives.
- Internal pricing and margin analyses; board materials on strategic rationale.
- Remedy feasibility from operations teams and third parties.
:::

::: {.callout-note title="Code box: merger sim skeleton"}
```{r}
#| label: merger-sim-skeleton-callout
#| message: false
#| warning: false
# Quarto executes from the project root (execute-dir: project), so this
# project-relative path works from any chapter.
source("program/R/helpers.R")
# products_df needs columns: product, firm, price, share, mc
# sim <- run_logit_sim(products_df, merging_firms = c("FirmA", "FirmB"))
# sim is a list: alpha, prices_pre, prices_post, delta, ownership_post
# data.frame(product = products_df$product,
#            pre = sim$prices_pre, post = sim$prices_post)
```
:::

::: {.callout-note title="Citations and comparative note"}
- Anchor claims to current US Merger Guidelines (2023) [@doj_ftc_hmg_2023] and legacy 2010 guidance [@doj_ftc_hmg_2010]; include EC Horizontal Guidelines [@ec_hmg_2004] and CMA merger assessment guidelines for comparisons.
- Cite empirical merger retrospective studies when presenting methods or benchmarks (e.g., airlines, hospitals).
- For vertical mergers, cite vUPP/EDM references and any key enforcement actions (e.g., US v. AT&T/Time Warner, EC cases on input foreclosure).
:::

::: {.callout-note title="Case box: Illustrative mergers"}
- Horizontal: airline mergers (UA/CO, DL/NW, AA/US) — retrospectives and coordinated effects; hospital mergers for local market power.
- Vertical: AT&T/Time Warner (US), Microsoft/Activision (platform/distribution) — input foreclosure/EDM debates.
- Mixed conglomerate/ad tech: Google/DoubleClick; media/telecom bundling examples.
:::

## Visualizations

### Market shares and HHI dashboard
This dashboard shows market structure before and after the merger, combining share distributions, HHI calculations, and competitive thresholds.

```{r}
#| label: merger-hhi-dashboard
#| fig-cap: "Merger screening dashboard: shares, HHI, and delta-HHI. Synthetic data for illustration only."
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 8
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
source("program/R/helpers.R")

# Simulated market data (replace with actual transaction/Nielsen data)
market_pre <- tibble::tribble(
  ~firm,      ~share,
  "Firm A",   0.28,
  "Firm B",   0.22,
  "Firm C",   0.18,
  "Firm D",   0.12,
  "Firm E",   0.08,
  "Firm F",   0.06,
  "Fringe",   0.06
)

# Post-merger: A acquires B
market_post <- market_pre |>
  mutate(
    firm = if_else(firm == "Firm B", "Firm A+B", firm),
    share = if_else(firm == "Firm A", share + 0.22, share)
  ) |>
  filter(firm != "Firm B") |>
  arrange(desc(share))

# Calculate HHI
calc_hhi <- function(shares) {
  sum((shares * 100)^2)
}

hhi_pre <- calc_hhi(market_pre$share)
hhi_post <- calc_hhi(market_post$share)
delta_hhi <- hhi_post - hhi_pre

# Prepare data for visualization
market_pre$period <- "Pre-merger"
market_post$period <- "Post-merger"
market_combined <- bind_rows(market_pre, market_post)
market_combined$period <- factor(market_combined$period,
                                  levels = c("Pre-merger", "Post-merger"))

# Plot 1: Share comparison
p1 <- ggplot(market_combined, aes(x = reorder(firm, share), y = share,
                                   fill = period)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("Pre-merger" = "#0072B2",
                                "Post-merger" = "#D55E00")) +
  labs(
    title = "Market Shares: Pre- vs. Post-Merger",
    x = NULL,
    y = "Market Share",
    fill = NULL
  ) +
  theme_antitrust() +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Plot 2: HHI change
hhi_data <- tibble::tribble(
  ~scenario, ~hhi,
  "Pre-merger", hhi_pre,
  "Post-merger", hhi_post
) |>
  mutate(
    scenario = factor(scenario, levels = c("Pre-merger", "Post-merger")),
    concern_level = case_when(
      hhi < 1000 ~ "Unconcentrated",
      hhi < 1800 ~ "Moderately concentrated",
      TRUE ~ "Highly concentrated (structural presumption)"
    )
  )

p2 <- ggplot(hhi_data, aes(x = scenario, y = hhi, fill = scenario)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 1000, linetype = "dashed", color = "gray40",
             linewidth = 0.8) +
  geom_hline(yintercept = 1800, linetype = "dashed", color = "gray40",
             linewidth = 0.8) +
  annotate("text", x = 2.5, y = 1000, label = "1,000 threshold",
           hjust = 0, vjust = -0.5, size = 3) +
  annotate("text", x = 2.5, y = 1800, label = "1,800 (2023 Guidelines)",
           hjust = 0, vjust = -0.5, size = 3) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1000, ymax = 1800,
           fill = "yellow", alpha = 0.05) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1800, ymax = Inf,
           fill = "red", alpha = 0.05) +
  scale_fill_manual(values = c("Pre-merger" = "#0072B2",
                                "Post-merger" = "#D55E00")) +
  labs(
    title = "HHI Analysis",
    subtitle = paste0("Δ HHI = ", round(delta_hhi, 0)),
    x = NULL,
    y = "HHI",
    fill = NULL
  ) +
  theme_antitrust() +
  theme(
    legend.position = "none",
    plot.title.position = "plot"
  )

# Plot 3: Concentration curve
market_post_sorted <- market_post |>
  arrange(desc(share)) |>
  mutate(cumulative_share = cumsum(share))

p3 <- ggplot(market_post_sorted, aes(x = seq_along(firm), y = cumulative_share)) +
  geom_line(color = "#D55E00", linewidth = 1.2) +
  geom_point(color = "#D55E00", size = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
  annotate("text", x = 1, y = 0.5, label = "50% market share",
           hjust = 0, vjust = -0.5, size = 3) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_x_continuous(breaks = seq_along(market_post_sorted$firm),
                     labels = market_post_sorted$firm) +
  labs(
    title = "Post-Merger Concentration Curve",
    subtitle = "Cumulative market share by firm rank",
    x = NULL,
    y = "Cumulative Share"
  ) +
  theme_antitrust() +
  theme(
    plot.title.position = "plot",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Combine plots
(p1 | p2) / p3 + plot_annotation(
  title = "Market Structure Dashboard",
  subtitle = paste0("Pre-merger HHI: ", round(hhi_pre, 0),
                   " | Post-merger HHI: ", round(hhi_post, 0),
                   " | Δ HHI: ", round(delta_hhi, 0)),
  caption = "US Merger Guidelines (2023): Structural presumption applies when
  HHI > 1,800 and Δ HHI > 100. Higher concentration warrants enhanced scrutiny."
)

# Summary table
cat("\nMarket structure summary:\n")
cat(paste0("Pre-merger HHI: ", round(hhi_pre, 0), "\n"))
cat(paste0("Post-merger HHI: ", round(hhi_post, 0), "\n"))
cat(paste0("Change in HHI: ", round(delta_hhi, 0), "\n"))
cat(paste0("\nCombined entity share: ",
           scales::percent(market_post$share[market_post$firm == "Firm A+B"],
                          accuracy = 0.1), "\n"))
```

**Interpretation:**
- **HHI thresholds**: The 2023 US Merger Guidelines establish a structural presumption of illegality when post-merger HHI > 1,800 and the merger increases HHI by more than 100 points.
- **Delta HHI**: Changes above 100 in concentrated markets (HHI > 1,800) trigger the structural presumption.
- **Combined entity**: The merged firm's share and rank indicate potential unilateral effects concerns.
- **Concentration curve**: Shows how quickly the top firms accumulate market share.

Replace simulated data with actual transaction volumes, Nielsen scanner data, or industry-specific sources (e.g., airline MIDT, hospital discharge data).

### Merger simulation waterfall
A waterfall chart decomposes the predicted post-merger price change into its component parts: diversion, margin, efficiencies, and second-order effects. This helps communicate which parameters drive the result.

```{r}
#| label: merger-sim-waterfall
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 7
library(dplyr)
library(ggplot2)
source("program/R/helpers.R")

# Simulation components (from a differentiated products model)
# Replace with actual simulation outputs. The final row is the total: its
# value is computed from the steps and the bar is drawn from zero, so it
# cannot drift out of line with the steps above it.
sim_components <- tibble::tribble(
  ~component,              ~value,     ~description,
  "Base price",            10.00,      "Pre-merger price",
  "First-order UPP",       +0.80,      "Diversion × Margin effect",
  "Internalization",       +0.35,      "Portfolio reoptimization",
  "Efficiency offset",     -0.25,      "Verified cost synergies",
  "Competitive response",  -0.15,      "Rival price reactions",
  "Post-merger price",     NA,         "Predicted equilibrium (computed)"
) |>
  mutate(
    type = case_when(
      component %in% c("Base price", "Post-merger price") ~ "total",
      value > 0 ~ "increase",
      value < 0 ~ "decrease",
      TRUE ~ "neutral"
    ),
    # Totals contribute no step: the base anchors the running sum, and the
    # post-merger total is the running sum itself, drawn from zero.
    step = if_else(component == "Post-merger price", 0, value),
    end = cumsum(step),
    start = if_else(type == "total", 0, lag(end, default = 0)),
    value = if_else(is.na(value), end, value),
    component = factor(component, levels = component)
  )

# Waterfall plot
ggplot(sim_components) +
  geom_rect(aes(xmin = as.numeric(component) - 0.4,
                xmax = as.numeric(component) + 0.4,
                ymin = start, ymax = end, fill = type),
            color = "black", linewidth = 0.5) +
  geom_text(aes(x = as.numeric(component),
                y = (start + end) / 2,
                label = scales::dollar(value, accuracy = 0.01)),
            size = 3.5, fontface = "bold") +
  geom_segment(data = filter(sim_components, !type %in% c("total")),
               aes(x = as.numeric(component) + 0.4,
                   xend = as.numeric(component) + 1 - 0.4,
                   y = end, yend = end),
               linetype = "dashed", color = "gray50") +
  scale_fill_manual(
    values = c(
      "total" = "#0072B2",
      "increase" = "#D55E00",
      "decrease" = "#009E73",
      "neutral" = "#999999"
    ),
    labels = c(
      "total" = "Price level",
      "increase" = "Price increase",
      "decrease" = "Price decrease",
      "neutral" = "Neutral"
    )
  ) +
  scale_x_continuous(
    breaks = seq_along(sim_components$component),
    labels = sim_components$component
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Merger Simulation Waterfall: Price Effect Decomposition",
    subtitle = "Breaking down predicted price change into component effects",
    x = NULL,
    y = "Price ($)",
    fill = "Effect type",
    caption = "Components from differentiated products logit simulation.
    Replace with actual model outputs."
  ) +
  theme_antitrust() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Summary statistics
cat("\nSimulation summary:\n")
cat(paste0("Pre-merger price: ",
           scales::dollar(sim_components$value[1], accuracy = 0.01), "\n"))
cat(paste0("Post-merger price: ",
           scales::dollar(sim_components$end[nrow(sim_components)],
                         accuracy = 0.01), "\n"))
cat(paste0("Price increase: ",
           scales::dollar(sim_components$end[nrow(sim_components)] -
                         sim_components$value[1], accuracy = 0.01),
           " (",
           scales::percent((sim_components$end[nrow(sim_components)] /
                           sim_components$value[1]) - 1, accuracy = 0.1),
           ")\n"))
```

**How to use this waterfall:**
- **First-order UPP**: Direct pricing pressure from internalizing diversion.
- **Internalization**: Additional optimization from portfolio effects (multi-product firms).
- **Efficiency offset**: Cost savings that reduce pricing pressure (must be verifiable and merger-specific).
- **Competitive response**: Predicted reactions from non-merging firms (can amplify or dampen effects).

**Practical tips:**
- Document the source of each component (diversion from surveys, margins from cost data, efficiencies from integration plans).
- Show sensitivity: run alternative scenarios varying key parameters.
- Compare to retrospective evidence: if similar past mergers raised prices by X%, does your simulation predict comparable magnitudes?

Replace simulated values with outputs from your actual merger simulation model (logit, nested logit, BLP, or custom structural model).

::: {.callout-tip title="Key takeaways"}
Structural screens come first. Under the 2023 Merger Guidelines, the presumption of illegality attaches when post-merger HHI exceeds 1,800 and the merger adds more than 100 points, or when the combined share reaches 30% with the same delta. The presumption opens the argument; diversion settles it. How many customers would switch between the merging parties drives UPP, GUPPI, and every simulation in this chapter, so the quality of the switching evidence---surveys, loyalty data, natural experiments---bounds the quality of the answer. Coordinated effects need different evidence: a maverick removed, symmetry increased, transparency enhanced. Those concerns bite even when unilateral effects are modest.

Vertical theories require showing both ability and incentive to foreclose, weighed against the elimination of double marginalization. Efficiencies count only when they are merger-specific and verifiable; synergies claimed in press releases rarely survive scrutiny. Retrospectives keep simulations honest---if comparable past mergers raised prices 2% and your model predicts 5%, explain the gap. And remedies must fix the diagnosed harm. A divestiture that creates a non-viable competitor is worse than none.
:::

## Exercises

1. **Data/code.** Using the merger simulation example in this chapter, change the outside share from 0.25 to 0.40 (more elastic market). How do the predicted price effects change? Explain intuitively why outside share matters for merger simulation.

2. **Conceptual.** Define the "maverick" concept in coordinated effects analysis. How would you quantitatively identify a maverick using pricing data? What additional qualitative evidence would strengthen or weaken the maverick argument?

3. **Case discussion.** Compare the EDM (elimination of double marginalization) defense in a vertical merger with the efficiencies defense in a horizontal merger. What evidence would you need to present for each, and why are agencies generally more skeptical of the vertical version?

4. **Data/code.** Using `data/derived/airline_route_shares.csv`, compute pre-merger HHI and delta-HHI for a hypothetical merger between the two largest carriers on the top 5 NYC routes. Which routes would exceed the 2023 Merger Guidelines structural presumption threshold (HHI > 1,800 and delta > 100)?

5. **Conceptual.** Explain the difference between a first-order price effect approximation (from UPP) and a full equilibrium merger simulation. When is the approximation likely to overstate or understate the actual price effect?

### Data exercise (checkable)

Two differentiated competitors with market shares of 25% and 15% propose to merge.

a. Using delta-HHI = 2 x s_1 x s_2, compute the change in HHI.
b. The diversion ratio from product 1 to product 2 is 0.30, product 2's margin is 0.40, and prices are equal. Compute the GUPPI for product 1.
c. The pre-merger HHI is 1,900. Does the structural presumption (HHI > 1,800, delta-HHI > 100) apply, and what does the GUPPI add?

::: {.callout-tip collapse="true" title="Worked answer"}
a. delta-HHI = 2 x 25 x 15 = **750**.
b. GUPPI_1 = D_12 x m_2 x (p_2 / p_1) = 0.30 x 0.40 x 1 = **0.12 = 12%**.
c. Post-merger HHI = 1,900 + 750 = 2,650 > 1,800 and delta-HHI 750 > 100, so **the structural presumption applies**. The 12% GUPPI sits well above the 5% screening threshold many agencies use, corroborating a unilateral-effects concern beyond the structural screen.
:::

## Looking ahead

Merger analysis produces artifacts—simulation outputs, UPP tables, event-study results—that reappear in **@sec-monopolization** and **@sec-litigation-practice**. The same frameworks apply when assessing whether a dominant firm's conduct harms competition.

Three steps make that reuse easier. Store merger sims, UPP tables, and event-study code in `data/derived/` with README files documenting assumptions and data sources. Note which datasets (Nielsen panels, loyalty data, procurement archives) need anonymization or refreshed pulls for teaching or publication. Keep a reference list of merger retrospectives (airlines, hospitals, beer) with their effect-size estimates, useful for validating simulations against real-world outcomes.

@sec-monopolization applies similar diversion and foreclosure frameworks to unilateral conduct by dominant firms. The simulation and screening tools transfer directly. The difference is that the conduct comes from a single firm rather than a transaction between two.
