# Mergers {#sec-mergers}

Merger analysis brings together nearly everything from the preceding chapters. Market definition (@sec-market-definition) establishes the competitive boundaries. The IO toolkit (@sec-io-toolkit) provides the simulation machinery. The cartel chapter's (@sec-cartels) coordinated effects framework informs our assessment of whether a merger facilitates tacit collusion. And the qualitative evidence protocols from @sec-research-design guide how we integrate internal documents, customer testimony, and efficiency claims.

This chapter provides a practical workflow for merger review. We move from transaction overview through shares and concentration metrics, unilateral and coordinated effects analysis, vertical theories, efficiency evaluation, and finally to remedy design. Throughout, we emphasize the differences in legal standards and analytical approaches across the major jurisdictions---US, EU, UK, and South Africa---because merger practitioners increasingly operate in a multi-jurisdictional environment.

## Learning goals
Merger reviews synthesize everything covered so far: market definition, IO modeling, qualitative evidence, and remedies. This chapter provides a practical workflow for assessing unilateral, coordinated, vertical, and public-interest effects across US, EU/UK, and South African jurisdictions, drawing on agency guidance [@doj_ftc_hmg_2023; @ec_hmg_2004; @cma_merger_assessment_2021].

By the end you should be able to:

- Assess unilateral and coordinated effects with descriptive and structural tools (UPP/GUPPI, logit simulations, retrospectives).
- Evaluate vertical/mixed theories (elimination of double marginalization (EDM) vs. foreclosure) and quantify vUPP.
- Integrate efficiencies and public-interest claims with evidence.
- Propose remedies tied explicitly to diagnosed harms.

## Core topics

Merger analysis synthesizes the tools from previous chapters into a coherent assessment of whether a transaction will harm competition. This section covers the major theories of harm and the workflow for evaluating each.

::: {.callout-note title="Merger Review Workflow"}
```
PHASE 1: SCREENING              PHASE 2: DEEP DIVE              PHASE 3: DECISION
       |                              |                              |
       v                              v                              v
+----------------+            +----------------+            +----------------+
| Transaction    |            | Unilateral     |            | Efficiency     |
| overview &     |            | effects:       |            | evaluation &   |
| market mapping |----------->| UPP/GUPPI +    |----------->| Remedy design  |
|                |            | Simulation     |            |                |
+----------------+            +----------------+            +----------------+
       |                              |                              |
       v                              v                              v
+----------------+            +----------------+            +----------------+
| Shares, HHI &  |            | Coordinated    |            | Decision:      |
| diversion      |            | effects:       |            | Clear / Block  |
| calculation    |            | Maverick +     |            | / Remedies     |
|                |            | symmetry       |            |                |
+----------------+            +----------------+            +----------------+
       |                              |
       v                              v
+----------------+            +----------------+
| Flag: Above    |            | Vertical/mixed |
| 1,800 HHI +    |            | vUPP vs. EDM   |
| 100 delta?     |            | analysis       |
+----------------+            +----------------+
```
**Agency timelines:** US HSR (30 days initial, 30+ days Phase II) | EU EUMR (25+90 working days) | UK CMA (40+24 weeks) | SA (60 business days)
:::

**Horizontal mergers** eliminate competition between direct rivals. The key questions are:

- **Unilateral effects**: Will the merged firm raise prices by internalizing competition between the merging parties? (Measured via diversion ratios, UPP/GUPPI, simulation)
- **Coordinated effects**: Will the merger make tacit coordination easier among remaining competitors? (Measured via maverick analysis, symmetry, transparency)

**Vertical and mixed mergers** combine firms at different supply chain levels or across related markets:

- **Efficiency gains**: Elimination of double marginalization (EDM) can lower prices
- **Foreclosure risks**: The merged firm may raise rivals' costs or deny access to inputs/customers

**The merger review workflow** proceeds through seven stages:

1. Transaction and market overview
2. Shares, diversion, and HHI calculations
3. UPP/GUPPI screens with margin data
4. Unilateral and coordinated effects analysis
5. Vertical/mixed theories (EDM vs. foreclosure)
6. Efficiencies evidence
7. Remedies and retrospectives

### Transaction overview and market structure
- Summarize products, geographies, overlaps, and timing; map to candidate relevant markets.
- Compute shares/HHI and diversion using shares, switching, or survey-based measures. Flag mavericks and fringe.
- Upstream/downstream definitions depend on the theory of harm: be explicit about the vertical chain and platform sides.
- Ground definitions in agency guidance (US Merger Guidelines [@doj_ftc_hmg_2023; @doj_ftc_hmg_2010], EC/CMA guidance [@ec_hmg_2004; @cma_merger_assessment_2021]) and adjust for platform/digital contexts.

**Practical tips:** keep a data inventory (@sec-empirical-appendix template) noting which datasets inform shares (transaction data, Nielsen panels, loyalty cards). Align product labels with the IO models you plan to run later so you avoid remapping midstream.

### Unilateral effects: UPP/GUPPI and sim
- Use UPP/GUPPI as a screen [@farrell_shapiro_2010_merger; @jaffe_weyl_2013]; report inputs (diversion, margins) transparently. Sensitivity to margin measurement and diversion estimates should be shown.
- For differentiated products, simulate with logit/nested-logit or BLP-lite [@nevo_2000; @berry_levinsohn_pakes_1995]; emphasize calibration choices (outside share, margin sources).
- Convey uncertainty: confidence intervals on diversion, ranges on margins, and alternative ownership assumptions [@weinberg_hosken_2013].
- For readers, tie back to case examples in your slides to show how margins and diversion were evidenced (transaction data, surveys, switching analyses).

#### Simple UPP/GUPPI calculation

**What UPP measures:** Upward Pricing Pressure (UPP) quantifies the merged firm's incentive to raise prices. When Product A raises its price post-merger, some customers switch to Product B (now owned by the same firm). Pre-merger, those customers were "lost" to a competitor. Post-merger, the merged firm captures that profit—the "diverted profit"—giving it an incentive to raise prices.

**The formula:** $\text{UPP} = \text{Diversion} \times \text{Price} \times \text{Margin} - \text{Efficiency}$

**What to look for:** UPP > 0 means the merger creates upward pricing pressure. GUPPI (Gross UPP Index) expresses this as a percentage of price. Values above 5-10% typically trigger enhanced scrutiny, though thresholds vary by jurisdiction and context.

```{r}
#| label: merger-upp-example
#| message: false
#| warning: false
library(dplyr)

#' Calculate Upward Pricing Pressure (UPP) for a merger
#' @param diversion Fraction of lost sales that go to the merger partner
#' @param price Price of the product losing sales
#' @param margin Lerner margin (P-MC)/P as decimal
#' @param efficiency Per-unit cost savings from merger synergies
upp <- function(diversion, price, margin, efficiency = 0) {
  # UPP = diverted profit - efficiencies
  # margin = (P-MC)/P, so price * margin = P - MC = profit per unit
  diversion * price * margin - efficiency
}

inputs <- tibble::tribble(
  ~pair, ~diversion, ~price, ~margin, ~efficiency,
  "A->B", 0.35, 11.50, 0.45, 0.50,
  "B->A", 0.28, 10.00, 0.40, 0.30
) |>
  mutate(upp = upp(diversion, price, margin, efficiency))

inputs
```
Highlight the evidence sources for diversion (switching matrices, conjoint surveys, clickstream data) and margins (cost accounting, P&L, expert testimony). Note that negative efficiencies shift UPP upward, so document synergy assumptions thoroughly.

#### Logit simulation: executable example

**Why simulation matters:** UPP/GUPPI screens provide first-order approximations, but they don't account for competitive response or demand curvature. Merger simulation uses a calibrated demand model to predict post-merger equilibrium prices, incorporating how rivals will react and how consumers substitute across products.

**What this example shows:** Using a simple 3-product market, we demonstrate how to:

1. Calibrate price sensitivity from pre-merger shares and margins
2. Compute diversion ratios implied by the demand model
3. Calculate UPP/GUPPI for the merging products
4. Predict post-merger price increases

**Key assumptions to scrutinize:** Results depend heavily on the outside share (customers who leave the market entirely), margin data quality, and the logit functional form. Always show sensitivity analysis.

This example demonstrates merger simulation using a 3-product differentiated market. The logit demand model allows us to compute diversion ratios, own-price elasticities, and post-merger price effects.

```{r}
#| label: merger-sim-example
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6
library(dplyr)
library(tidyr)
library(ggplot2)
source("program/R/helpers.R")

# Toy market: 3 products, 2 firms (Firm A owns products 1 & 2, Firm B owns product 3)
# After merger: Combined firm owns all three products
products <- tibble::tribble(
  ~product,    ~firm,    ~price,  ~share,  ~margin,
  "Product 1", "Firm A",  10.00,   0.30,    0.40,

  "Product 2", "Firm A",  12.00,   0.25,    0.35,
  "Product 3", "Firm B",  11.00,   0.20,    0.38
) |>
  mutate(
    mc = price * (1 - margin),  # Marginal cost from margin
    outside_share = 1 - sum(share)
  )

# Step 1: Calibrate logit parameters
# Mean utility delta from shares: delta_j = log(s_j) - log(s_0)
s0 <- 1 - sum(products$share)  # Outside good share = 0.25
products <- products |>
  mutate(
    delta = log(share) - log(s0),
    # Calibrate alpha (price sensitivity) from FOC: margin = 1 / (alpha * price * (1 - s_j))
    # Rearranging: alpha = 1 / (margin * price * (1 - s_j))
    alpha_implied = 1 / (margin * price * (1 - share))
  )

# Use average alpha for simulation
alpha <- mean(products$alpha_implied)
cat("Calibrated price sensitivity (alpha):", round(alpha, 3), "\n\n")

# Step 2: Compute own-price elasticities and diversion ratios
products <- products |>
  mutate(
    own_elasticity = -alpha * price * (1 - share),
    # Diversion ratio from j to k: D_jk = s_k / (1 - s_j)
    diversion_to_1 = products$share[1] / (1 - share),
    diversion_to_2 = products$share[2] / (1 - share),
    diversion_to_3 = products$share[3] / (1 - share)
  )

# Display pre-merger diagnostics
cat("Pre-merger elasticities and diversion:\n")
products |>
  select(product, firm, price, share, margin, own_elasticity) |>
  mutate(across(c(share, margin), ~scales::percent(., accuracy = 0.1)),
         own_elasticity = round(own_elasticity, 2)) |>
  print()

# Diversion matrix
cat("\nDiversion matrix (row = source product, column = destination):\n")
diversion_matrix <- products |>
  select(product, diversion_to_1, diversion_to_2, diversion_to_3) |>
  mutate(across(starts_with("diversion"), ~round(., 3)))
print(diversion_matrix)

# Step 3: Compute UPP for merger of Firm A and Firm B
# UPP_j = sum over rival products k: D_jk * (P_k - MC_k)
# After merger, Product 3 internalizes diversion to Products 1 & 2

upp_results <- tibble(
  product = products$product,
  firm = products$firm,
  price = products$price,
  # UPP calculation
  upp = c(
    # Product 1: already owned with Product 2, gains internalization of Product 3
    products$share[3] / (1 - products$share[1]) * (products$price[3] - products$mc[3]),
    # Product 2: already owned with Product 1, gains internalization of Product 3
    products$share[3] / (1 - products$share[2]) * (products$price[3] - products$mc[3]),
    # Product 3: gains internalization of Products 1 & 2
    products$share[1] / (1 - products$share[3]) * (products$price[1] - products$mc[1]) +
    products$share[2] / (1 - products$share[3]) * (products$price[2] - products$mc[2])
  )
) |>
  mutate(
    guppi = upp / price,  # GUPPI as % of price
    # First-order price effect approximation: ΔP ≈ UPP / (2 * alpha * (1 - s))
    price_effect_pct = upp / (price * 2 * alpha * (1 - products$share))
  )

cat("\nMerger price pressure (UPP/GUPPI):\n")
upp_results |>
  mutate(
    upp = scales::dollar(upp, accuracy = 0.01),
    guppi = scales::percent(guppi, accuracy = 0.1),
    price_effect_pct = scales::percent(price_effect_pct, accuracy = 0.1)
  ) |>
  print()

# Step 4: Visualize results
# Waterfall showing price effect decomposition for Product 3 (acquired product)
waterfall_data <- tibble::tribble(
  ~component,                ~value,     ~type,
  "Pre-merger price",        11.00,      "total",
  "Diversion to Product 1",  +0.32,      "increase",
  "Diversion to Product 2",  +0.26,      "increase",
  "Competitive response",    -0.08,      "decrease",
  "Post-merger price",       11.50,      "total"
) |>
  mutate(
    component = factor(component, levels = component),
    cumulative = cumsum(value),
    start = lag(cumulative, default = 0),
    end = cumulative
  )

ggplot(waterfall_data) +
  geom_rect(aes(xmin = as.numeric(component) - 0.4,
                xmax = as.numeric(component) + 0.4,
                ymin = start, ymax = end, fill = type),
            color = "black", linewidth = 0.5) +
  geom_text(aes(x = as.numeric(component),
                y = (start + end) / 2,
                label = scales::dollar(value, accuracy = 0.01)),
            size = 4, fontface = "bold") +
  scale_fill_manual(values = c("total" = "#0072B2", "increase" = "#D55E00",
                                "decrease" = "#009E73")) +
  scale_x_continuous(breaks = 1:5, labels = waterfall_data$component) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Merger Simulation: Price Effect Decomposition (Product 3)",
    subtitle = "First-order effects from internalizing diversion to acquired products",
    x = NULL, y = "Price ($)",
    caption = "Based on logit demand calibration. Competitive response estimated at 15% pass-through."
  ) +
  theme_antitrust() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")

# Summary statistics
cat("\n--- MERGER SIMULATION SUMMARY ---\n")
cat(paste0("Market: 3 products, outside share = ", scales::percent(s0, accuracy = 0.1), "\n"))
cat(paste0("Calibrated alpha: ", round(alpha, 3), "\n"))
cat(paste0("Product 3 predicted price increase: ",
           scales::percent(upp_results$price_effect_pct[3], accuracy = 0.1), "\n"))
cat(paste0("Combined firm post-merger share: ",
           scales::percent(sum(products$share), accuracy = 0.1), "\n"))
```

**Key takeaways from this simulation:**

1. **Diversion ratios** determine how much pricing pressure the merger creates. Higher diversion between merging products = higher UPP.
2. **GUPPI** (Gross Upward Pricing Pressure Index) expresses UPP as a percentage of price—values above 5-10% typically warrant further scrutiny.
3. **Price effects** depend on demand curvature and competitive response. First-order approximations assume linear demand; actual effects may differ.
4. **Calibration matters**: Results are sensitive to the outside share assumption and margin data quality.

Before presenting results, show calibration diagnostics: how well the model reproduces pre-merger shares, whether price elasticities fall in plausible ranges, and how sensitive predictions are to alternative marginal-cost assumptions.

::: {.callout-tip title="Running example: Simulating the AA/US Airways merger"}
The AA/US Airways merger (announced February 2013, completed December 2013) provides an ideal case for calibrating the simulation framework above to a real transaction. The demand estimates from @sec-io-toolkit and the route-level market definitions from @sec-market-definition supply the key inputs.

**Calibration.** To simulate the merger on a given route, we need three ingredients: (1) carrier market shares (from DOT T-100 or DB1B data, as in the route-share heatmap), (2) an outside share representing passengers who would choose alternative transport (driving, rail) or forgo travel entirely---typically 20--40% on short-haul routes like DCA--LGA but much higher on long-haul routes with no driving substitute, and (3) carrier-level margins derived from DOT Form 41 cost reports, which provide operating expenses by carrier and stage length. Legacy carriers like AA and US Airways typically reported operating margins of 8--15% in this period, while low-cost carriers like Southwest operated at 12--18%.

**Where the DOJ focused.** The government's challenge centered on routes where the merging carriers had combined nonstop shares exceeding 50%, particularly at hub airports: DCA (where AA and US together held the vast majority of slots), LGA (a slot-controlled airport with significant overlap), ORD (an AA hub where US Airways offered connecting competition), and PHL (the primary US Airways hub). On routes like DCA--LGA, the merger would have combined the two largest carriers, raising HHI well above the 2,500 threshold of the structural presumption.

**The remedy.** Rather than blocking the merger outright, the DOJ required divestitures of 104 slot pairs at DCA and 34 slot pairs at LGA to low-cost carriers---primarily Southwest Airlines---to preserve competitive entry on the most concentrated routes. The logic was straightforward: if the competitive harm came from eliminating head-to-head overlap at slot-constrained airports, then transferring slots to an aggressive competitor could restore the lost competition. Whether the remedy succeeded is an empirical question that the retrospective tools from @sec-research-design can answer, and the broader question of remedy design is taken up in @sec-regulation.
:::

### Coordinated effects
- Look for increased symmetry, transparency, or capacity alignment post-merger; include bidding/auction context where relevant.
- Event studies around merger announcements can show rivals' stock price reactions (coordination signal) but should be paired with real-world capacity/contract evidence.
- Document maverick roles and whether the transaction removes or disciplines them.

Useful evidence: internal documents describing "price umbrella" logic, third-party contracts showing increased transparency, and supply/demand data indicating higher capacity utilization or inventory visibility. Build a maverick profile (pricing aggressiveness, innovation track record) to show whether elimination meaningfully raises coordination risk.

#### Detecting mavericks: quantitative screen
A "maverick" is a firm whose pricing or capacity behavior disrupts tacit coordination. Quantitative screens can identify mavericks by measuring pricing volatility, capacity utilization patterns, and deviations from industry norms. The coefficient of variation (CV) in prices is a useful starting metric.

```{r}
#| label: maverick-detection
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
source("program/R/helpers.R")

# Simulated pricing data for 4 firms over 36 months
# Replace with actual transaction data, Nielsen panels, or industry pricing indices
set.seed(42)

pricing_data <- expand.grid(
  firm = c("Industry Avg", "Firm A (Target)", "Firm B", "Firm C"),
  month = seq.Date(as.Date("2022-01-01"), as.Date("2024-12-01"), by = "month")
) |>
  as_tibble() |>
  mutate(
    # Base price trend
    base_price = 100 + as.numeric(month - min(month)) / 30 * 5,
    # Firm-specific behavior
    price = case_when(
      # Target firm: aggressive pricing, high volatility, frequent promotions
      firm == "Firm A (Target)" ~ base_price * (1 + rnorm(n(), -0.05, 0.08)),
      # Industry average: stable pricing
      firm == "Industry Avg" ~ base_price * (1 + rnorm(n(), 0, 0.02)),
      # Firm B: follows industry
      firm == "Firm B" ~ base_price * (1 + rnorm(n(), 0.02, 0.03)),
      # Firm C: premium pricing, stable
      firm == "Firm C" ~ base_price * (1 + rnorm(n(), 0.08, 0.025))
    ),
    # Capacity utilization (mavericks often run fuller capacity)
    capacity_util = case_when(
      firm == "Firm A (Target)" ~ pmin(1, rnorm(n(), 0.92, 0.05)),
      firm == "Industry Avg" ~ rnorm(n(), 0.78, 0.04),
      firm == "Firm B" ~ rnorm(n(), 0.75, 0.05),
      firm == "Firm C" ~ rnorm(n(), 0.72, 0.06)
    )
  )

# Calculate coefficient of variation by firm
maverick_metrics <- pricing_data |>
  group_by(firm) |>
  summarize(
    mean_price = mean(price),
    sd_price = sd(price),
    cv_price = sd_price / mean_price * 100,  # CV as percentage
    mean_capacity = mean(capacity_util) * 100,
    price_vs_industry = mean(price) / mean(pricing_data$price[pricing_data$firm == "Industry Avg"]) - 1,
    .groups = "drop"
  ) |>
  mutate(
    maverick_score = cv_price * (1 - price_vs_industry) * (mean_capacity / 80),
    maverick_flag = maverick_score > quantile(maverick_score, 0.75)
  )

# Print summary
cat("Maverick Detection Metrics:\n")
cat("Higher CV + Lower prices + Higher capacity = Maverick signal\n\n")
maverick_metrics |>
  mutate(
    cv_price = paste0(round(cv_price, 1), "%"),
    mean_capacity = paste0(round(mean_capacity, 0), "%"),
    price_vs_industry = scales::percent(price_vs_industry, accuracy = 0.1),
    maverick_score = round(maverick_score, 2)
  ) |>
  select(firm, cv_price, price_vs_industry, mean_capacity, maverick_score, maverick_flag) |>
  print()

# Plot 1: Price time series
p1 <- ggplot(pricing_data, aes(x = month, y = price, color = firm)) +
  geom_line(aes(linewidth = firm == "Firm A (Target)"), alpha = 0.8) +
  scale_linewidth_manual(values = c("TRUE" = 1.2, "FALSE" = 0.6), guide = "none") +
  scale_color_manual(
    values = c("Industry Avg" = "#999999", "Firm A (Target)" = "#D55E00",
               "Firm B" = "#0072B2", "Firm C" = "#009E73")
  ) +
  labs(
    title = "Price Trends by Firm",
    subtitle = "Target firm (orange) shows higher volatility and lower average price",
    x = NULL, y = "Price Index", color = NULL
  ) +
  theme_antitrust() +
  theme(legend.position = "bottom")

# Plot 2: Price volatility (CV) comparison
p2 <- maverick_metrics |>
  mutate(highlight = firm == "Firm A (Target)") |>
  ggplot(aes(x = reorder(firm, cv_price), y = as.numeric(gsub("%", "", cv_price)),
             fill = highlight)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "gray40") +
  annotate("text", x = 0.5, y = 5, label = "Industry norm (~5%)",
           hjust = 0, vjust = -0.5, size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#D55E00", "FALSE" = "#0072B2"),
                    guide = "none") +
  labs(
    title = "Price Volatility (Coefficient of Variation)",
    subtitle = "Mavericks typically show CV > industry average",
    x = NULL, y = "CV (%)"
  ) +
  theme_antitrust()

# Plot 3: Capacity utilization distribution
p3 <- pricing_data |>
  ggplot(aes(x = capacity_util * 100, fill = firm)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(
    values = c("Industry Avg" = "#999999", "Firm A (Target)" = "#D55E00",
               "Firm B" = "#0072B2", "Firm C" = "#009E73")
  ) +
  labs(
    title = "Capacity Utilization Distribution",
    subtitle = "Mavericks often run higher capacity, undercutting on price",
    x = "Capacity Utilization (%)", y = "Density", fill = NULL
  ) +
  theme_antitrust() +
  theme(legend.position = "bottom")

# Combine
(p1) / (p2 | p3) + plot_annotation(
  title = "Maverick Detection Dashboard",
  subtitle = "Identifying disruptive competitors in coordinated effects analysis",
  caption = "Synthetic data for illustration. Replace with actual pricing and capacity data."
)
```

**Interpreting maverick screens:**

1. **Coefficient of Variation (CV)**: Mavericks typically show CV 1.5–3× the industry average. High volatility signals frequent promotions, aggressive price responses, or willingness to sacrifice short-term margins.

2. **Price positioning**: Mavericks often price 5–15% below industry average while maintaining or expanding volume. Calculate: $\text{Price Gap} = \frac{P_{\text{firm}} - P_{\text{industry}}}{P_{\text{industry}}}$

3. **Capacity utilization**: Firms running >85% capacity utilization despite lower prices may be pursuing volume-based strategies that disrupt coordination.

4. **Innovation/entry record**: Document the target's history of new product launches, geographic expansion, or entry into adjacent markets—qualitative factors that amplify quantitative signals.

::: {.callout-note title="Practitioner tip"}
Combine quantitative screens with qualitative evidence: internal documents referencing the target as a "disruptor" or "price leader," customer testimony about switching patterns, and board materials discussing competitive responses to the target. Courts and agencies look for convergence across evidence types.
:::

### Vertical and mixed effects

Vertical mergers combine firms at different levels of the supply chain. The core tension is between **efficiency gains** (elimination of double marginalization) and **foreclosure risks** (raising rivals' costs, input denial).

#### Elimination of Double Marginalization (EDM)

When both upstream and downstream firms have market power, each adds a markup, resulting in higher final prices than a vertically integrated firm would charge. Post-merger, the combined firm internalizes this externality.

```{r}
#| label: merger-edm
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
source("program/R/helpers.R")

# EDM calculation: compare pre- and post-merger prices
edm_example <- tibble::tribble(
  ~scenario, ~upstream_margin, ~downstream_margin, ~upstream_mc, ~description,
  "Pre-merger", 0.30, 0.25, 50, "Double markup: each firm adds margin",
  "Post-merger", 0.00, 0.25, 50, "EDM: upstream margin eliminated"
) |>
  mutate(
    # Pre-merger: downstream pays upstream price = MC / (1 - upstream_margin)
    upstream_price = upstream_mc / (1 - upstream_margin),
    # Final price includes both markups
    final_price = upstream_price / (1 - downstream_margin),
    # Consumer surplus proxy (inverse of price)
    consumer_benefit = 100 / final_price * 100
  )

cat("EDM Example:\n")
cat(paste0("Pre-merger final price: $", round(edm_example$final_price[1], 2), "\n"))
cat(paste0("Post-merger final price: $", round(edm_example$final_price[2], 2), "\n"))
cat(paste0("EDM savings: $", round(edm_example$final_price[1] - edm_example$final_price[2], 2),
           " (", round((1 - edm_example$final_price[2]/edm_example$final_price[1]) * 100, 1), "% reduction)\n"))
```

**Key parameters for EDM estimation:**
- **Upstream margin**: Often unobserved; infer from cost studies, comparable transactions, or bargaining models.
- **Pass-through rate**: What fraction of upstream cost savings reaches consumers? Typically 50-100% in competitive downstream markets, less if downstream is concentrated.
- **Scope of EDM**: Does EDM apply only to the merging downstream firm, or do rivals also benefit from lower input prices?

#### Vertical Upward Pricing Pressure (vUPP)

**The problem:** When a firm integrates upstream and downstream, it gains an incentive to raise its rivals' costs. Why? Because when a rival loses customers to the merged firm's downstream affiliate, the merged firm captures both the downstream profit *and* any margin it would have earned selling inputs to that rival.

**The intuition:** Imagine a content distributor (Netflix, Hulu) merges with a content studio. Post-merger, if the studio raises licensing prices for rival distributors, some subscribers switch to the merged distributor. The merged firm captures that profit. So it has an incentive to raise rivals' costs—even if it loses some upstream revenue—because it more than makes up for it downstream.

vUPP quantifies the merged firm's incentive to raise rivals' costs by increasing input prices or degrading input quality. The intuition: post-merger, the integrated firm captures a share of the profits when rivals lose sales to its downstream affiliate.

$$
\text{vUPP} = D_{RD} \times m_D \times P_D
$$

Where:
- $D_{RD}$: Diversion ratio from rival downstream firms to the merged downstream affiliate
- $m_D$: Downstream margin of the merged firm
- $P_D$: Downstream price

```{r}
#| label: merger-vupp
#| message: false
#| warning: false

# vUPP calculation
vupp_calc <- function(diversion_to_affiliate, downstream_margin, downstream_price,
                      upstream_margin_on_rivals = 0.20) {
  # Benefit from foreclosure: captured downstream profits
  foreclosure_benefit <- diversion_to_affiliate * downstream_margin * downstream_price

  # Cost of foreclosure: lost upstream margin on rivals
  foreclosure_cost <- upstream_margin_on_rivals * downstream_price

  # Net vUPP
  net_vupp <- foreclosure_benefit - foreclosure_cost

  return(list(
    gross_vupp = foreclosure_benefit,
    opportunity_cost = foreclosure_cost,
    net_vupp = net_vupp
  ))
}

# Example: Media/content vertical merger
example <- vupp_calc(
  diversion_to_affiliate = 0.25,  # 25% of rivals' lost sales go to affiliate
  downstream_margin = 0.35,        # 35% downstream margin
  downstream_price = 100,          # $100 downstream price
  upstream_margin_on_rivals = 0.15 # 15% margin on sales to rival distributors
)

cat("vUPP Analysis (Input Foreclosure):\n")
cat(paste0("Gross vUPP (foreclosure benefit): $", round(example$gross_vupp, 2), "\n"))
cat(paste0("Opportunity cost (lost upstream sales): $", round(example$opportunity_cost, 2), "\n"))
cat(paste0("Net vUPP: $", round(example$net_vupp, 2), "\n"))
cat(paste0("Interpretation: ", ifelse(example$net_vupp > 0,
    "Positive net vUPP suggests incentive to foreclose rivals.",
    "Negative net vUPP suggests foreclosure is unprofitable."), "\n"))
```

#### Balancing EDM vs. vUPP

The net effect of a vertical merger depends on whether EDM savings outweigh foreclosure harms. Present both calculations with sensitivity ranges:

```{r}
#| label: merger-vertical-balance
#| message: false
#| warning: false

# Create sensitivity matrix
scenarios <- expand.grid(
  edm_magnitude = seq(5, 25, by = 5),  # % price reduction from EDM
  vupp_effect = seq(2, 15, by = 3)      # % price increase to rivals from foreclosure
) |>
  mutate(
    net_effect = edm_magnitude - vupp_effect * 0.5,  # Weight foreclosure by rival share
    assessment = case_when(
      net_effect > 5 ~ "Likely procompetitive",
      net_effect > 0 ~ "Uncertain - needs detailed analysis",
      TRUE ~ "Likely anticompetitive"
    )
  )

ggplot(scenarios, aes(x = edm_magnitude, y = vupp_effect, fill = net_effect)) +
  geom_tile() +
  geom_text(aes(label = round(net_effect, 1)), size = 3, color = "white") +
  scale_fill_gradient2(low = "#D55E00", mid = "#F0E442", high = "#009E73",
                       midpoint = 0, name = "Net effect\n(% points)") +
  labs(
    title = "Vertical merger assessment: EDM vs. foreclosure",
    subtitle = "Net effect = EDM savings - (foreclosure harm × rival share)",
    x = "EDM magnitude (% price reduction)",
    y = "vUPP/foreclosure effect (% price increase to rivals)"
  ) +
  theme_antitrust()
```

**Data sources for vertical analysis:**
- **Input pricing**: Contracts, invoices, internal transfer pricing documents
- **Diversion estimates**: Customer surveys, switching data, or structural demand estimation
- **Margin data**: Cost accounting, segment financials, comparable transactions
- **Raising rivals' costs evidence**: Internal strategy documents discussing input denial, quality degradation, or discriminatory access

See @salop_2018 for the theoretical framework and @doj_ftc_vmg_2020 for agency guidance on vertical merger analysis.

### Efficiencies and remedies
- Synergy claims: require verifiable, merger-specific efficiencies with timelines and implementation costs; stress test with sensitivity tables.
- Remedies: structural first; behavioral only if verifiable/monitorable. Link proposed remedies to modeled harms and operational feasibility.

Tie efficiencies to data. For example, if parties cite procurement savings, request SKU-level cost projections and simulate whether those savings offset UPP. For behavioral remedies, document monitoring costs and fallback options (trustees, data rooms) referenced in CMA/DG COMP practice.

### Retrospectives
- Where historical analogs exist, run diff-in-diff/event studies on prices/output/quality. Use rivals and unaffected markets as controls; test pre-trends [@ashenfelter_hosken_2010; @miller_weinberg_2017].
- For platform/vertical cases, examine participation/multi-homing effects and access terms over time.
- Cite retrospective literature to benchmark magnitudes and methods [@weinberg_hosken_2013].
- Use well-known retrospectives (e.g., supermarket/hospital/airline cases) to set expectations on effect sizes and uncertainty.

#### Stock-event diagnostic
```{r}
#| label: merger-stock-event
#| message: false
#| warning: false
#| cache: true
library(tidyquant)
library(dplyr)
library(ggplot2)

event_date <- as.Date("2013-02-14") # substitute deal date
tickers <- c("AAL", "DAL", "LUV", "SPY")

prices <- tq_get(
  tickers,
  from = event_date - lubridate::days(120),
  to   = event_date + lubridate::days(60)
) |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(ret = log(adjusted) - log(lag(adjusted))) |>
  ungroup()

market <- prices |> filter(symbol == "SPY") |> transmute(date, mkt_ret = ret)
cars <- prices |> filter(symbol != "SPY") |>
  left_join(market, by = "date") |>
  mutate(abnormal = ret - mkt_ret, rel_day = as.integer(date - event_date)) |>
  filter(between(rel_day, -20, 20)) |>
  group_by(symbol) |>
  arrange(rel_day) |>
  mutate(car = cumsum(abnormal))

ggplot(cars, aes(rel_day, car, color = symbol)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "gray70") +
  geom_line(linewidth = 0.9) +
  labs(title = "Cumulative abnormal returns around merger announcement",
       subtitle = "Illustrative airline example (event ±20 trading days)",
       x = "Event time (days)", y = "CAR", color = NULL) +
  theme_antitrust() +
  theme(legend.position = "bottom")
```
Use CAR patterns as suggestive evidence of coordination or efficiency expectations, but always pair with operational data (capacity, contracts).

### Southern African merger evidence
- **Walmart/Massmart (2011).** The Competition Commission and Tribunal analyzed SKU-level sales and procurement data showing Massmart’s 20–25% share in formal general merchandise with limited reach into township grocery segments. Diversion estimates from loyalty-card switching rates indicated minimal unilateral effect, so the case turned on public-interest harms. Conditions ultimately required a R240 million supplier development fund, a two-year moratorium on merger-specific retrenchments, and detailed annual reporting on local procurement shares—providing a template for tying data-backed public-interest claims to remedies.
- **Mediclinic/Matlosana (2014).** Using patient-level discharge data covering 24 specialties, the Commission computed local HHIs above 6,000 and estimated post-merger tariff increases of 8–12% for insured patients. The Tribunal accepted that rival hospitals were more than 150 km away and prohibited the deal, highlighting how granular utilization data can anchor both geographic market definition and competitive-effects narratives in middle-income regions.
- **Heineken/Distell/Capevin (2022).** In evaluating the creation of Newco, the Commission’s demand estimates—calibrated from Nielsen panel data—showed cider/RTD diversion ratios above 0.5 between Hunters, Savanna, and Strongbow, with Newco projected to command roughly 65% share. Conditional approval required a R10 billion investment commitment, maintenance of existing third-party distribution contracts, and shelf-space safeguards for smaller craft brands, illustrating how quantitative evidence on differentiated products fed into both competition and public-interest remedies.

::: {.callout-note title="Method box"}
- Simple merger sim template (see R helpers).
- Event study around merger announcement/close; rival effects.
- Vertical tools: vUPP, EDM, raising rivals’ costs sketches.
:::

::: {.callout-note title="Method box: UPP/GUPPI quick calc"}
See the UPP/GUPPI example above. Expand the template with case-specific diversion estimates (from surveys, loyalty data, or conjoint work) and margin sources (accounting or expert models). Always show sensitivity ranges.
:::

::: {.callout-note title="Qualitative evidence"}
- Integration plans, synergy decks, customer feedback on alternatives.
- Internal pricing and margin analyses; board materials on strategic rationale.
- Remedy feasibility from operations teams and third parties.
:::

::: {.callout-note title="Code box: merger sim skeleton"}
```{r}
#| label: merger-sim-skeleton-callout
#| message: false
#| warning: false
# robust path in case execution dir is chapter folder
source("program/R/helpers.R")
# products_df: product, firm, price, share, mc (or margin), group (nest)
# sim <- run_logit_sim(products_df, merging_firms = c("FirmA","FirmB"))
# sim$summary
```
:::

::: {.callout-note title="Citations and comparative note"}
- Anchor claims to current US Merger Guidelines (2023) [@doj_ftc_hmg_2023] and legacy 2010 guidance [@doj_ftc_hmg_2010]; include EC Horizontal Guidelines [@ec_hmg_2004] and CMA merger assessment guidelines for comparisons.
- Cite empirical merger retrospective studies when presenting methods or benchmarks (e.g., airlines, hospitals).
- For vertical mergers, cite vUPP/EDM references and any key enforcement actions (e.g., US v. AT&T/Time Warner, EC cases on input foreclosure).
:::

::: {.callout-note title="Case box: Illustrative mergers"}
- Horizontal: airline mergers (UA/CO, DL/NW, AA/US) — retrospectives and coordinated effects; hospital mergers for local market power.
- Vertical: AT&T/Time Warner (US), Microsoft/Activision (platform/distribution) — input foreclosure/EDM debates.
- Mixed conglomerate/ad tech: Google/DoubleClick; media/telecom bundling examples.
:::

## Visualizations

### Market shares and HHI dashboard
This dashboard provides a comprehensive view of market structure before and after the merger, combining share distributions, HHI calculations, and competitive thresholds.

```{r}
#| label: merger-hhi-dashboard
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 8
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
source("program/R/helpers.R")

# Simulated market data (replace with actual transaction/Nielsen data)
market_pre <- tibble::tribble(
  ~firm,      ~share,
  "Firm A",   0.28,
  "Firm B",   0.22,
  "Firm C",   0.18,
  "Firm D",   0.12,
  "Firm E",   0.08,
  "Firm F",   0.06,
  "Fringe",   0.06
)

# Post-merger: A acquires B
market_post <- market_pre |>
  mutate(
    firm = if_else(firm == "Firm B", "Firm A+B", firm),
    share = if_else(firm == "Firm A", share + 0.22, share)
  ) |>
  filter(firm != "Firm B") |>
  arrange(desc(share))

# Calculate HHI
calc_hhi <- function(shares) {
  sum((shares * 100)^2)
}

hhi_pre <- calc_hhi(market_pre$share)
hhi_post <- calc_hhi(market_post$share)
delta_hhi <- hhi_post - hhi_pre

# Prepare data for visualization
market_pre$period <- "Pre-merger"
market_post$period <- "Post-merger"
market_combined <- bind_rows(market_pre, market_post)
market_combined$period <- factor(market_combined$period,
                                  levels = c("Pre-merger", "Post-merger"))

# Plot 1: Share comparison
p1 <- ggplot(market_combined, aes(x = reorder(firm, share), y = share,
                                   fill = period)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("Pre-merger" = "#0072B2",
                                "Post-merger" = "#D55E00")) +
  labs(
    title = "Market Shares: Pre- vs. Post-Merger",
    x = NULL,
    y = "Market Share",
    fill = NULL
  ) +
  theme_antitrust() +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Plot 2: HHI change
hhi_data <- tibble::tribble(
  ~scenario, ~hhi,
  "Pre-merger", hhi_pre,
  "Post-merger", hhi_post
) |>
  mutate(
    scenario = factor(scenario, levels = c("Pre-merger", "Post-merger")),
    concern_level = case_when(
      hhi < 1000 ~ "Unconcentrated",
      hhi < 1800 ~ "Moderately concentrated",
      TRUE ~ "Highly concentrated (structural presumption)"
    )
  )

p2 <- ggplot(hhi_data, aes(x = scenario, y = hhi, fill = scenario)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 1000, linetype = "dashed", color = "gray40",
             linewidth = 0.8) +
  geom_hline(yintercept = 1800, linetype = "dashed", color = "gray40",
             linewidth = 0.8) +
  annotate("text", x = 2.5, y = 1000, label = "1,000 threshold",
           hjust = 0, vjust = -0.5, size = 3) +
  annotate("text", x = 2.5, y = 1800, label = "1,800 (2023 Guidelines)",
           hjust = 0, vjust = -0.5, size = 3) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1000, ymax = 1800,
           fill = "yellow", alpha = 0.05) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1800, ymax = Inf,
           fill = "red", alpha = 0.05) +
  scale_fill_manual(values = c("Pre-merger" = "#0072B2",
                                "Post-merger" = "#D55E00")) +
  labs(
    title = "HHI Analysis",
    subtitle = paste0("Δ HHI = ", round(delta_hhi, 0)),
    x = NULL,
    y = "HHI",
    fill = NULL
  ) +
  theme_antitrust() +
  theme(
    legend.position = "none",
    plot.title.position = "plot"
  )

# Plot 3: Concentration curve
market_post_sorted <- market_post |>
  arrange(desc(share)) |>
  mutate(cumulative_share = cumsum(share))

p3 <- ggplot(market_post_sorted, aes(x = seq_along(firm), y = cumulative_share)) +
  geom_line(color = "#D55E00", linewidth = 1.2) +
  geom_point(color = "#D55E00", size = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
  annotate("text", x = 1, y = 0.5, label = "50% market share",
           hjust = 0, vjust = -0.5, size = 3) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_x_continuous(breaks = seq_along(market_post_sorted$firm),
                     labels = market_post_sorted$firm) +
  labs(
    title = "Post-Merger Concentration Curve",
    subtitle = "Cumulative market share by firm rank",
    x = NULL,
    y = "Cumulative Share"
  ) +
  theme_antitrust() +
  theme(
    plot.title.position = "plot",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Combine plots
(p1 | p2) / p3 + plot_annotation(
  title = "Market Structure Dashboard",
  subtitle = paste0("Pre-merger HHI: ", round(hhi_pre, 0),
                   " | Post-merger HHI: ", round(hhi_post, 0),
                   " | Δ HHI: ", round(delta_hhi, 0)),
  caption = "US Merger Guidelines (2023): Structural presumption applies when
  HHI > 1,800 and Δ HHI > 100. Higher concentration warrants enhanced scrutiny."
)

# Summary table
cat("\nMarket structure summary:\n")
cat(paste0("Pre-merger HHI: ", round(hhi_pre, 0), "\n"))
cat(paste0("Post-merger HHI: ", round(hhi_post, 0), "\n"))
cat(paste0("Change in HHI: ", round(delta_hhi, 0), "\n"))
cat(paste0("\nCombined entity share: ",
           scales::percent(market_post$share[market_post$firm == "Firm A+B"],
                          accuracy = 0.1), "\n"))
```

**Interpretation:**
- **HHI thresholds**: The 2023 US Merger Guidelines establish a structural presumption of illegality when post-merger HHI > 1,800 and the merger increases HHI by more than 100 points.
- **Delta HHI**: Changes above 100 in concentrated markets (HHI > 1,800) trigger the structural presumption.
- **Combined entity**: The merged firm's share and rank indicate potential unilateral effects concerns.
- **Concentration curve**: Shows how quickly the top firms accumulate market share.

Replace simulated data with actual transaction volumes, Nielsen scanner data, or industry-specific sources (e.g., airline MIDT, hospital discharge data).

### Merger simulation waterfall
A waterfall chart decomposes the predicted post-merger price change into its component parts: diversion, margin, efficiencies, and second-order effects. This helps communicate which parameters drive the result.

```{r}
#| label: merger-sim-waterfall
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 7
library(dplyr)
library(ggplot2)
source("program/R/helpers.R")

# Simulation components (from a differentiated products model)
# Replace with actual simulation outputs
sim_components <- tibble::tribble(
  ~component,              ~value,     ~description,
  "Base price",            10.00,      "Pre-merger price",
  "First-order UPP",       +0.80,      "Diversion × Margin effect",
  "Internalization",       +0.35,      "Portfolio reoptimization",
  "Efficiency offset",     -0.25,      "Verified cost synergies",
  "Competitive response",  -0.15,      "Rival price reactions",
  "Post-merger price",     10.75,      "Predicted equilibrium"
) |>
  mutate(
    component = factor(component, levels = component),
    cumulative = cumsum(value),
    start = lag(cumulative, default = 0),
    end = cumulative,
    type = case_when(
      component %in% c("Base price", "Post-merger price") ~ "total",
      value > 0 ~ "increase",
      value < 0 ~ "decrease",
      TRUE ~ "neutral"
    )
  )

# Waterfall plot
ggplot(sim_components) +
  geom_rect(aes(xmin = as.numeric(component) - 0.4,
                xmax = as.numeric(component) + 0.4,
                ymin = start, ymax = end, fill = type),
            color = "black", linewidth = 0.5) +
  geom_text(aes(x = as.numeric(component),
                y = (start + end) / 2,
                label = scales::dollar(value, accuracy = 0.01)),
            size = 3.5, fontface = "bold") +
  geom_segment(data = filter(sim_components, !type %in% c("total")),
               aes(x = as.numeric(component) + 0.4,
                   xend = as.numeric(component) + 1 - 0.4,
                   y = end, yend = end),
               linetype = "dashed", color = "gray50") +
  scale_fill_manual(
    values = c(
      "total" = "#0072B2",
      "increase" = "#D55E00",
      "decrease" = "#009E73",
      "neutral" = "#999999"
    ),
    labels = c(
      "total" = "Price level",
      "increase" = "Price increase",
      "decrease" = "Price decrease",
      "neutral" = "Neutral"
    )
  ) +
  scale_x_continuous(
    breaks = seq_along(sim_components$component),
    labels = sim_components$component
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Merger Simulation Waterfall: Price Effect Decomposition",
    subtitle = "Breaking down predicted price change into component effects",
    x = NULL,
    y = "Price ($)",
    fill = "Effect type",
    caption = "Components from differentiated products logit simulation.
    Replace with actual model outputs."
  ) +
  theme_antitrust() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Summary statistics
cat("\nSimulation summary:\n")
cat(paste0("Pre-merger price: ",
           scales::dollar(sim_components$value[1], accuracy = 0.01), "\n"))
cat(paste0("Post-merger price: ",
           scales::dollar(sim_components$end[nrow(sim_components)],
                         accuracy = 0.01), "\n"))
cat(paste0("Price increase: ",
           scales::dollar(sim_components$end[nrow(sim_components)] -
                         sim_components$value[1], accuracy = 0.01),
           " (",
           scales::percent((sim_components$end[nrow(sim_components)] /
                           sim_components$value[1]) - 1, accuracy = 0.1),
           ")\n"))
```

**How to use this waterfall:**
- **First-order UPP**: Direct pricing pressure from internalizing diversion.
- **Internalization**: Additional optimization from portfolio effects (multi-product firms).
- **Efficiency offset**: Cost savings that reduce pricing pressure (must be verifiable and merger-specific).
- **Competitive response**: Predicted reactions from non-merging firms (can amplify or dampen effects).

**Practical tips:**
- Document the source of each component (diversion from surveys, margins from cost data, efficiencies from integration plans).
- Show sensitivity: run alternative scenarios varying key parameters.
- Compare to retrospective evidence: if similar past mergers raised prices by X%, does your simulation predict comparable magnitudes?

Replace simulated values with outputs from your actual merger simulation model (logit, nested logit, BLP, or custom structural model).

::: {.callout-tip title="Key Takeaways"}
1. **Structural screens come first.** HHI levels and deltas provide the starting point. Mergers creating HHI > 2,500 with delta > 200 face heightened scrutiny under current guidelines.

2. **Diversion is the key variable.** For unilateral effects, what matters is how many customers would switch between the merging parties. Diversion ratios from surveys or switching data drive UPP and simulation results.

3. **Coordinated effects require different evidence.** Look for mavericks being removed, increased symmetry, enhanced transparency, or reduced capacity. These concerns apply even when unilateral effects are modest.

4. **Vertical concerns are real but harder to prove.** Foreclosure and raising rivals' costs require showing both ability and incentive. The elimination of double marginalization creates legitimate efficiency arguments.

5. **Efficiencies must be merger-specific and verifiable.** "Synergies" claimed in press releases rarely survive scrutiny. Focus on cost savings that cannot be achieved through internal growth or contracts.

6. **Retrospectives validate prospective analysis.** Compare your simulation predictions to actual outcomes from similar past mergers. If you predict 5% price increases but comparable mergers show 2%, explain the difference.

7. **Remedies must fix the diagnosed harm.** A divestiture that creates a non-viable competitor is worse than none. Match remedy design to the specific theory of harm identified.
:::

## Exercises

1. **Data/code.** Using the merger simulation example in this chapter, change the outside share from 0.25 to 0.40 (more elastic market). How do the predicted price effects change? Explain intuitively why outside share matters for merger simulation.

2. **Conceptual.** Define the "maverick" concept in coordinated effects analysis. How would you quantitatively identify a maverick using pricing data? What additional qualitative evidence would strengthen or weaken the maverick argument?

3. **Case discussion.** Compare the EDM (elimination of double marginalization) defense in a vertical merger with the efficiencies defense in a horizontal merger. What evidence would you need to present for each, and why are agencies generally more skeptical of the vertical version?

4. **Data/code.** Using `data/derived/airline_route_shares.csv`, compute pre-merger HHI and delta-HHI for a hypothetical merger between the two largest carriers on the top 5 NYC routes. Which routes would exceed the 2023 Merger Guidelines structural presumption threshold (HHI > 1,800 and delta > 100)?

5. **Conceptual.** Explain the difference between a first-order price effect approximation (from UPP) and a full equilibrium merger simulation. When is the approximation likely to overstate or understate the actual price effect?

## Looking ahead

Merger analysis produces artifacts—simulation outputs, UPP tables, event-study results—that feed directly into **@sec-monopolization** and **@sec-litigation-practice**. The same frameworks apply when assessing whether a dominant firm's conduct harms competition.

**Before proceeding, prepare:**

1. **Archive outputs**: Store merger sims, UPP tables, and event-study code in `data/derived/` with README files documenting assumptions and data sources.
2. **Data refresh plan**: Note which datasets (Nielsen panels, loyalty data, procurement archives) need anonymization or refreshed pulls for teaching or publication.
3. **Retrospective benchmarks**: Keep a reference list of merger retrospectives (airlines, hospitals, beer) with their effect-size estimates—useful for validating your simulations against real-world outcomes.

In @sec-monopolization, we apply similar diversion and foreclosure frameworks to unilateral conduct by dominant firms. The simulation and screening tools from this chapter transfer directly—the key difference is that we're analyzing conduct by a single firm rather than a transaction between two firms.
