---
title: "Mergers"
---

## Learning goals
- Assess unilateral and coordinated effects with both descriptive and model-based tools.
- Run and interpret UPP/GUPPI, merger simulation, and merger retrospectives.
- Integrate efficiencies claims with evidence; design remedies.

## Core topics
- Market structure pre/post; diversion from shares and switching data.
- UPP/GUPPI screens; thresholds and caveats.
- Merger simulation: differentiated products logit/nested logit; calibrating margins and ownership.
- Coordinated effects: maverick analysis, transparency, capacity and bidding dynamics.
- Retrospectives: diff-in-diff/event studies with rivals and controls.
- Horizontal, vertical, and mixed mergers; define “upstream” and “downstream” relative to the theory of harm and adapt analysis to industry specifics.
- Workflow: (1) transaction and market overview; (2) shares/diversion/HHI; (3) UPP/GUPPI and margins; (4) unilateral and coordinated effects (structural or reduced form); (5) vertical/mixed theories (EDM vs. foreclosure); (6) efficiencies evidence; (7) remedies and retrospectives.

### Transaction overview and market structure
- Summarize products, geographies, overlaps, and timing; map to candidate relevant markets.
- Compute shares/HHI and diversion using shares, switching, or survey-based measures. Flag mavericks and fringe.
- Upstream/downstream definitions depend on the theory of harm: be explicit about the vertical chain and platform sides.
- Ground definitions in agency guidance (US Merger Guidelines [@doj_ftc_hmg_2023; @doj_ftc_hmg_2010], EC/CMA guidance [@ec_hmg_2004; @cma_merger_assessment_2021]) and adjust for platform/digital contexts.

### Unilateral effects: UPP/GUPPI and sim
- Use UPP/GUPPI as a screen; report inputs (diversion, margins) transparently. Sensitivity to margin measurement and diversion estimates should be shown.
- For differentiated products, simulate with logit/nested-logit or BLP-lite; emphasize calibration choices (outside share, margin sources).
- Convey uncertainty: confidence intervals on diversion, ranges on margins, and alternative ownership assumptions.
- For readers, tie back to case examples in your slides to show how margins and diversion were evidenced (transaction data, surveys, switching analyses).

### Coordinated effects
- Look for increased symmetry, transparency, or capacity alignment post-merger; include bidding/auction context where relevant.
- Event studies around merger announcements can show rivals’ stock price reactions (coordination signal) but should be paired with real-world capacity/contract evidence.
- Document maverick roles and whether the transaction removes or disciplines them.

### Vertical and mixed effects
- EDM vs. foreclosure: quantify both. vUPP and upward pricing pressure on rivals’ access should be shown alongside EDM magnitudes.
- Identify potential raising rivals’ costs channels (input denial, degradation, parity clauses) and model likely price effects.
- Conglomerate/mixed cases: bundling/tying risk, data advantage transfer, and cross-market leverage should be explicitly linked to facts.
- When defining “upstream” vs. “downstream,” anchor to transaction flows and incentives; platform settings may require mapping sides rather than tiers.

### Efficiencies and remedies
- Synergy claims: require verifiable, merger-specific efficiencies with timelines and implementation costs; stress test with sensitivity tables.
- Remedies: structural first; behavioral only if verifiable/monitorable. Link proposed remedies to modeled harms and operational feasibility.

### Retrospectives
- Where historical analogs exist, run diff-in-diff/event studies on prices/output/quality. Use rivals and unaffected markets as controls; test pre-trends.
- For platform/vertical cases, examine participation/multi-homing effects and access terms over time.
- Cite retrospective literature to benchmark magnitudes and methods.
- Use well-known retrospectives (e.g., supermarket/hospital/airline cases [@ashenfelter_hosken_2010]) to set expectations on effect sizes and uncertainty.

::: {.callout-note title="Method box"}
- Simple merger sim template (see R helpers).
- Event study around merger announcement/close; rival effects.
- Vertical tools: vUPP, EDM, raising rivals’ costs sketches.
:::

::: {.callout-note title="Method box: UPP/GUPPI quick calc"}
```{r}
#| label: merger-upp-example
upp_data <- data.frame(
  product = c("A","B"),
  price   = c(10, 12),
  share   = c(0.25, 0.20),
  margin  = c(0.35, 0.30) # as share of price
)
# diversion from A to B (and B to A)
div_AB <- 0.25
div_BA <- 0.20

upp_A <- div_AB * upp_data$price[upp_data$product=="B"] * (1 - upp_data$margin[upp_data$product=="B"])
upp_B <- div_BA * upp_data$price[upp_data$product=="A"] * (1 - upp_data$margin[upp_data$product=="A"])
upp_A; upp_B
```
Interpretation: positive UPP suggests upward pricing pressure absent offsetting efficiencies; replace with case-specific diversion/margins.
:::

::: {.callout-note title="Qualitative evidence"}
- Integration plans, synergy decks, customer feedback on alternatives.
- Internal pricing and margin analyses; board materials on strategic rationale.
- Remedy feasibility from operations teams and third parties.
:::

::: {.callout-note title="Code box: merger sim skeleton"}
```{r}
#| label: merger-sim-skeleton
#| message: false
#| warning: false
# robust path in case execution dir is chapter folder
source("../R/helpers.R")
# products_df: product, firm, price, share, mc (or margin), group (nest)
# sim <- run_logit_sim(products_df, merging_firms = c("FirmA","FirmB"))
# sim$summary
```
:::

::: {.callout-note title="Citations and comparative note"}
- Anchor claims to current US Merger Guidelines (2023) [@doj_ftc_hmg_2023] and legacy 2010 guidance [@doj_ftc_hmg_2010]; include EC Horizontal Guidelines [@ec_hmg_2004] and CMA merger assessment guidelines for comparisons.
- Cite empirical merger retrospective studies when presenting methods or benchmarks (e.g., airlines, hospitals).
- For vertical mergers, cite vUPP/EDM references and any key enforcement actions (e.g., US v. AT&T/Time Warner, EC cases on input foreclosure).
:::

::: {.callout-note title="Case box: Illustrative mergers"}
- Horizontal: airline mergers (UA/CO, DL/NW, AA/US) — retrospectives and coordinated effects; hospital mergers for local market power.
- Vertical: AT&T/Time Warner (US), Microsoft/Activision (platform/distribution) — input foreclosure/EDM debates.
- Mixed conglomerate/ad tech: Google/DoubleClick; media/telecom bundling examples.
- Use class notes/slides (`notes and slides/pdf/7_Antitrust_and_Merger_Review_Economic_Analysis_and_Legal_Practice.pdf`, `notes and slides/Lecture note on merger analysis.pdf`) to populate specifics and citations.
:::
