---
title: "Digital Markets and Platforms"
---

## Learning goals
Digital markets combine traditional IO concepts with product design, governance, and behavioral nudges. Drawing on DMA/Digital Markets Unit briefings and recent enforcement actions, this chapter helps you:

- Analyze multi-sided participation, indirect network effects, and multi-homing incentives.
- Quantify self-preferencing, ranking changes, and default effects on user/merchant behavior.
- Evaluate data advantages, access restrictions, and algorithmic pricing risks.
- Integrate product requirement documents (PRDs), API policies, and telemetry with econometric evidence for global enforcement settings.

Expect to combine telemetry, clickstream data, partner contracts, and qualitative product documentation with econometric tests and simulation scaffolds.

## Core topics
- Platform demand and multi-homing: participation elasticities, steering, matching efficiency.
- Self-preferencing and ranking: click-through impacts, A/B test interpretation, fairness metrics.
- Data advantages and economies of scope; switching costs, lock-in, and interoperability.
- Algorithmic pricing and tacit collusion risk.
- Workflow for platform cases: market definition, theory of harm, mechanism measurement, effects, and remedies (choice screens, access mandates, data portability, separation).

## Platform workflow overview

1. **Define sides and metrics.** Identify user, merchant, advertiser, developer, courier, or publisher sides. Track MAU/DAU, GMV, multi-homing rates, engagement depth, and conversion metrics.  
2. **Map fees and monetization.** Build ladders for commissions, subscriptions, ad loads, and data monetization; compute pass-through and elasticity. For two-sided market theory, see @rochet_tirole_2003 and @armstrong_2006.  
3. **Theory of harm.** Common mechanisms include self-preferencing, default tying, MFN/parity clauses, data leveraging, interoperability restrictions, algorithmic ranking biases, and algorithmic collusion.  
4. **Mechanism measurement.** Quantify ranking impacts, default effects, fee changes, and data advantages; document multi-homing constraints and API access terms.  
5. **Effects & remedies.** Measure partner churn, user switching, price/fee changes, and innovation impacts. Propose remedies (choice screens, fair-ranking reports, API access rules, data portability, structural separation) tied to measured harms.

## Participation, multi-homing, and network effects

- **Participation curves:** Model each side's activity as a function of fees, quality, and participation on the other side. Use logit, linear, or Poisson models with cross-side terms [@evans_schmalensee_2007; @hagiu_wright_2015].  
- **Multi-homing rates:** Estimate the share of users or merchants active on rival platforms. OIPMI surveys and EU DMA filings include methodologies you can reuse [@parker_van_alstyne_2005].  
- **Match quality:** Track conversion rates, completion times, or satisfaction metrics per platform side to show how conduct affects outcomes.  
- **Elasticities:** Two-sided elasticities inform fee caps and remedy design (e.g., how merchant participation responds to lower commissions or improved ranking fairness).

### Two-sided demand scaffold
```{r}
#| label: digital-two-sided
#| message: false
#| warning: false
library(dplyr)
library(fixest)

# Synthetic example; replace with platform telemetry
panel <- expand.grid(platform = c("Alpha","Beta"), period = 2018:2023) |>
  mutate(
    fees_users = runif(n(), 0, 4),
    fees_merchants = runif(n(), 15, 30),
    quality = runif(n(), 0.5, 1),
    users = exp(5 - 0.25 * fees_users + 0.4 * quality + rnorm(n(), 0, 0.2)),
    merchants = exp(4 - 0.12 * fees_merchants + 0.3 * users / 1e5 + rnorm(n(), 0, 0.2))
  )

user_model <- feols(log(users) ~ fees_users + quality + log(merchants) | platform + period, data = panel)
merchant_model <- feols(log(merchants) ~ fees_merchants + log(users) | platform + period, data = panel)
summary(user_model)
summary(merchant_model)
```
Replace the synthetic `panel` with sanitized telemetry (monthly active users/merchants) or aggregated OIPMI data stored in `data/examples/digital-two-sided.csv`. Document the estimated cross-side elasticities in your memo.

## Ranking, self-preferencing, and defaults

### Ranking impact visualization
```{r}
#| label: digital-ranking
#| message: false
#| warning: false
library(dplyr)
library(tidyr)
library(ggplot2)

ranking <- tibble::tribble(
  ~slot, ~first_party_ctr, ~third_party_ctr,
  1, 0.16, 0.08,
  2, 0.13, 0.06,
  3, 0.11, 0.05,
  4, 0.08, 0.04,
  5, 0.06, 0.03
) |>
  pivot_longer(-slot, names_to = "listing", values_to = "ctr")

ggplot(ranking, aes(slot, ctr, color = listing)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("#1b9e77", "#d95f02"), labels = c("First-party", "Third-party")) +
  labs(
    title = "Conversion by ranking slot",
    subtitle = "Illustrative data; use OIPMI or platform A/B metrics when available",
    x = "Slot (1 = top)",
    y = "Conversion rate",
    color = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "bottom")
```
Populate the tibble with actual ranking data (e.g., OIPMI buy-box audits, CMA Amazon Marketplace analysis) or maintain sanitized values in `data/examples/digital-ranking.csv`. The CMA's market studies are publicly available at [gov.uk/cma](https://www.gov.uk/government/organisations/competition-and-markets-authority).

### Default-choice event study scaffold
```{r}
#| label: digital-default-event
#| message: false
#| warning: false
#| eval: false
library(fixest)

# df requires: device_id, week, default_event (event date), share_rival
# event_model <- feols(share_rival ~ i(week, default_event, ref = -1) | device_id + week, data = df)
# etable(event_model)
```
Use telemetry from choice screens (EU Android choice screen, DMA compliance reports) or partner-provided logs. If live data are unavailable, build synthetic event panels mirroring DMA summary stats (store in `data/examples/digital-default.csv`).

## Southern African digital enforcement snapshots
- **Online Intermediation Platforms Market Inquiry (OIPMI, 2020–2023).** Transaction-level data from marketplaces, app stores, delivery apps, and travel platforms revealed Google Search’s >90% share, 25–30% delivery commissions, and buy-box algorithms favoring first-party listings. Remedies require fair-ranking reports, SME fee caps, and anti-steering rights inside aggregator apps.  
- **Media and Digital Platforms Inquiry (2023–ongoing).** Impression-level logs from Meta, Google, and local publishers show Google capturing ~65% of programmatic display spend, while local publishers net <30% after fees, framing bargaining remedies for news publishers.  
- **GovChat/Meta interim relief (2021).** WhatsApp Business API telemetry demonstrated GovChat’s public-service reach (>8 million users) and negligible incremental cost; interim relief preserved access while exclusion claims proceeded, blending platform data with public-interest considerations.

## Methodologies

::: {.callout-note title="Method box"}
- Platform fee/price pass-through across sides.  
- Event studies on ranking/default changes; DMA/DMA-style interventions.  
- Two-sided logit or bargaining sketches to test fee caps and interoperability remedies.
:::

::: {.callout-note title="Qualitative evidence"}
- Product requirement documents, experiment summaries, and governance memos.  
- Partner contracts (parity/MFN, exclusivity), API access terms, data-sharing policies.  
- User research/surveys capturing multi-homing, switching costs, and friction sources.
:::

::: {.callout-note title="Citations and comparative note"}
- US cases (Epic v. Apple, DOJ Google Search and Ad Tech, FTC Amazon Marketplace), EU/DMA (Google Shopping/Android, DMA obligations), UK (CMA digital market studies), KFTC/JFTC app store cases, SAMR digital enforcement.  
- Empirical studies on defaults, ranking effects, and multi-homing (e.g., Luca et al., 2016, "Does Google Content Degrade Google Search?," Harvard Business School Working Paper; Edelman & Lai, 2016, "Design of Search Engine Services," *Journal of Economics & Management Strategy* 25(1): 187-218).  
- Flag differences in parity/MFN treatment and remedy preferences (EU fair-ranking obligations vs. US injunctive relief).
:::

## Visualizations and data sourcing
- **Default impact chart (Roadmap 9.1):** Source DMA compliance reports (available at [ec.europa.eu/dma](https://ec.europa.eu/)), CMA choice-screen experiments, or sanitized telemetry. If unavailable, simulate from DMA summary stats.  
- **Multi-homing Sankey (Roadmap 9.2):** Use OIPMI survey microdata or replicate with local surveys; store in `data/examples/digital-sankey.csv` until live data arrive.  
- **Ranking change event analysis (Roadmap 9.3):** Combine OIPMI buy-box audits or platform-provided logs with the ranking scaffold above.  
- **Fee ladder diagrams:** Pull commission schedules from OIPMI appendices or global public filings (App Store, Play Store, Amazon Marketplace).

Document each dataset in `data/README.md` with provenance, confidentiality notes, and instructions for swapping synthetic vs. production data. When we "fill with real data," replace the synthetic CSVs, re-render figures, and cache sanitized outputs for publication builds.

### Southern African digital enforcement snapshots
- **Online Intermediation Platforms Market Inquiry (OIPMI, 2020–2023).** The Competition Commission collected transaction-level data from e-commerce marketplaces, app stores, food delivery apps, and travel platforms. Findings highlighted Google Search’s >90% share of general search queries, food-delivery commissions clustering at 25–30%, and marketplace buy-box algorithms that favored first-party listings (e.g., Takealot Retail) over third-party sellers with identical fulfillment metrics. Final remedies require fair-ranking reports, caps on marketplace storage/fulfillment fees for SMEs, and anti-steering relief that lets restaurants promote direct-order channels inside aggregator apps.
- **Advertising and digital news media (Media and Digital Platforms Market Inquiry, 2023–ongoing).** Building on OIPMI data, the Commission is ingesting impression-level logs from Meta, Google, and local publishers to quantify revenue imbalances in display advertising. The interim statement reports that Google commands roughly 65% of display ad spend routed through programmatic channels, while local publishers face average take-rates below 30% after demand-side/platform fees—data points that frame prospective bargaining remedies for news publishers across Southern Africa.
- **GovChat/Meta interim relief (Competition Commission v. Meta Platforms, 2021).** The Commission used WhatsApp Business API telemetry to demonstrate GovChat’s role in reaching 8+ million South African users for public-service messaging, and it showed how Meta’s proposed off-boarding would foreclose a civic-use case despite negligible incremental infrastructure cost. Interim relief preserved access while the Tribunal weighed the exclusion case, offering a blueprint for combining platform-side telemetry with public-interest arguments in developing markets.

::: {.callout-note title="Method box"}
- Platform price/fee pass-through; elasticity across sides.
- Event studies on default changes or ranking tweaks.
- Simple two-sided logit/bargaining sketches; diversion across sides.
:::

::: {.callout-note title="Qualitative evidence"}
- Product requirement docs and experiments; governance and content policies.
- Partner contracts (parity/MFN, exclusivity), API access terms.
- User research surveys on multi-homing and switching.
:::

::: {.callout-note title="Citations and comparative note"}
- Cite platform cases and regulations: US (e.g., Epic v. Apple, DOJ Google Search/Ad Tech), EU (Google Shopping/Android, [DMA](https://ec.europa.eu/info/strategy/priorities-2019-2024/europe-fit-digital-age/digital-markets-act-ensuring-fair-and-open-digital-markets_en)), UK ([CMA market investigations](https://www.gov.uk/cma-cases)), and other jurisdictions (e.g., Korea KFTC app stores where relevant).
- Reference empirical studies on defaults and ranking effects, and any A/B or natural experiment evidence from platforms.
- Note differences in treatment of MFNs/parity and self-preferencing across jurisdictions.
:::

## Enhanced Visualizations

### Multi-homing Sankey diagram
Multi-homing patterns reveal competitive constraints and switching costs. A Sankey diagram shows how users flow across platforms, helping visualize whether platforms compete head-to-head or serve distinct niches.

```{r}
#| label: digital-multihoming-sankey
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 8
library(dplyr)
library(ggplot2)
library(ggalluvial)

# Simulated multi-homing survey data
# Replace with OIPMI survey data or DMA compliance reports
set.seed(234)
n_users <- 1000

multihoming <- tibble(
  user_id = 1:n_users,
  platform_a = sample(c("Active", "Inactive"), n_users, replace = TRUE,
                     prob = c(0.7, 0.3)),
  platform_b = sample(c("Active", "Inactive"), n_users, replace = TRUE,
                     prob = c(0.5, 0.5)),
  platform_c = sample(c("Active", "Inactive"), n_users, replace = TRUE,
                     prob = c(0.3, 0.7))
) |>
  mutate(
    pattern = case_when(
      platform_a == "Active" & platform_b == "Active" & platform_c == "Active" ~ "All three",
      platform_a == "Active" & platform_b == "Active" ~ "A + B",
      platform_a == "Active" & platform_c == "Active" ~ "A + C",
      platform_b == "Active" & platform_c == "Active" ~ "B + C",
      platform_a == "Active" ~ "A only",
      platform_b == "Active" ~ "B only",
      platform_c == "Active" ~ "C only",
      TRUE ~ "None"
    )
  )

# Aggregate flows
flows <- multihoming |>
  count(pattern) |>
  mutate(
    pct = n / sum(n),
    pattern = factor(pattern,
                    levels = c("A only", "B only", "C only",
                              "A + B", "A + C", "B + C", "All three", "None"))
  ) |>
  arrange(pattern)

# Bar chart showing multi-homing patterns
p1 <- ggplot(flows, aes(x = pattern, y = pct, fill = pattern)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1)),
            vjust = -0.5, size = 3.5) +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.1))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Multi-homing Patterns Across Platforms",
    subtitle = "User activity across Platform A, B, and C",
    x = NULL,
    y = "Share of Users",
    caption = "Illustrative data. Replace with OIPMI survey or DMA compliance reports."
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title.position = "plot"
  )

p1

# Summary statistics
cat("\nMulti-homing summary:\n")
cat(paste0("Single-homing (one platform only): ",
          scales::percent(sum(flows$pct[flows$pattern %in%
                              c("A only", "B only", "C only")]),
                         accuracy = 0.1), "\n"))
cat(paste0("Multi-homing (2+ platforms): ",
          scales::percent(sum(flows$pct[grepl("\\+|three", flows$pattern)]),
                         accuracy = 0.1), "\n"))
```

**Interpretation:**
- **High single-homing**: Suggests strong lock-in, switching costs, or network effects. Platforms may have market power.
- **High multi-homing**: Indicates competitive pressure; users actively compare options. But check if multi-homing is symmetric or if one platform dominates usage.
- **Asymmetric patterns**: If Platform A appears in most combinations, it may be a "must-have" platform with gatekeeping power.

**Data sources:**
- OIPMI survey microdata (South Africa)
- DMA compliance reports (EU)
- CMA market studies (UK)
- Platform-provided telemetry (subject to confidentiality)

### Choice screen / default effect visualization
Quantify how defaults affect market shares using difference-in-differences or event studies around choice screen implementations.

```{r}
#| label: digital-default-impact
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 7
library(dplyr)
library(ggplot2)
library(patchwork)

# Simulated data: Android choice screen impact (EU DMA context)
# Replace with actual DMA compliance data or CMA experiments
set.seed(345)
weeks <- -24:24
countries <- c("Control (no choice screen)", "Treatment (choice screen)")

default_data <- expand.grid(
  week = weeks,
  country = countries
) |>
  mutate(
    # Pre-intervention: both countries similar
    # Post-intervention: treatment shows rival gain
    rival_share = case_when(
      country == "Control (no choice screen)" ~ 0.05 + 0.0005 * week + rnorm(n(), 0, 0.01),
      week < 0 ~ 0.05 + 0.0005 * week + rnorm(n(), 0, 0.01),
      TRUE ~ 0.05 + 0.0005 * week + 0.08 + rnorm(n(), 0, 0.01)
    ),
    rival_share = pmax(0.01, pmin(0.25, rival_share))
  )

# Time series plot
p1 <- ggplot(default_data, aes(x = week, y = rival_share,
                                color = country, linetype = country)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40",
             linewidth = 1) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  annotate("text", x = 0, y = 0.22, label = "Choice screen\nintroduced",
           hjust = -0.1, size = 3.5, fontface = "italic") +
  scale_color_manual(values = c("Control (no choice screen)" = "#999999",
                                 "Treatment (choice screen)" = "#0072B2")) +
  scale_linetype_manual(values = c("Control (no choice screen)" = "dashed",
                                   "Treatment (choice screen)" = "solid")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Impact of Choice Screen on Rival Search Engine Share",
    subtitle = "Difference-in-Differences: Treatment vs. Control Countries",
    x = "Weeks relative to choice screen introduction",
    y = "Rival search engine share",
    color = NULL,
    linetype = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Coefficient plot (event study coefficients)
# Simulated DiD estimates
did_coefs <- tibble(
  week = weeks[weeks != -1]  # reference period
) |>
  mutate(
    estimate = 0.08 * as.numeric(week >= 0) + rnorm(n(), 0, 0.01),
    ci_lower = estimate - 1.96 * 0.01,
    ci_upper = estimate + 1.96 * 0.01
  )

p2 <- ggplot(did_coefs, aes(x = week, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40",
             linewidth = 1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
              alpha = 0.2, fill = "#0072B2") +
  geom_line(color = "#0072B2", linewidth = 1) +
  geom_point(color = "#0072B2", size = 2) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Event Study Coefficients",
    subtitle = "Treatment effect by week (relative to week -1)",
    x = "Weeks relative to choice screen",
    y = "Effect on rival share (percentage points)",
    caption = "Illustrative data. Replace with DMA compliance reports or CMA experiments."
  ) +
  theme_classic(base_size = 12) +
  theme(plot.title.position = "plot")

# Combined plot
p1 / p2

# Summary statistics
pre_treatment <- mean(default_data$rival_share[
  default_data$country == "Treatment (choice screen)" & default_data$week < 0])
post_treatment <- mean(default_data$rival_share[
  default_data$country == "Treatment (choice screen)" & default_data$week >= 0])
effect <- post_treatment - pre_treatment

cat("\nChoice screen impact summary:\n")
cat(paste0("Pre-intervention rival share: ",
          scales::percent(pre_treatment, accuracy = 0.1), "\n"))
cat(paste0("Post-intervention rival share: ",
          scales::percent(post_treatment, accuracy = 0.1), "\n"))
cat(paste0("Estimated treatment effect: ",
          scales::percent(effect, accuracy = 0.1),
          " (", round(effect / pre_treatment * 100, 0), "% increase)\n"))
```

**How to use this analysis:**
- **Pre-trends**: Check parallel trends before intervention to validate DiD assumptions.
- **Treatment effect**: Quantify how much rival share increased after choice screen.
- **Heterogeneity**: Analyze by device type, user demographics, or country.
- **Persistence**: Track whether effects sustain over time or decay.

**DMA context:**
The EU Digital Markets Act requires designated gatekeepers to offer choice screens for browsers, search engines, and other services. This analysis framework can evaluate compliance and effectiveness.

### Platform fee structure visualization
Commission and fee structures are critical evidence in platform cases. Show how fees vary by seller tier, product category, or fulfillment method.

```{r}
#| label: digital-fee-structure
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 7
library(dplyr)
library(ggplot2)
library(tidyr)

# Fee structures from major platforms (public sources)
# App Store, Play Store, Amazon Marketplace, food delivery
fees <- tibble::tribble(
  ~platform,            ~category,          ~base_commission, ~fulfillment_fee, ~payment_fee,
  "App Store",          "Apps",             0.30,             0.00,              0.00,
  "App Store",          "Apps (< $1M)",     0.15,             0.00,              0.00,
  "Play Store",         "Apps",             0.30,             0.00,              0.00,
  "Play Store",         "Apps (< $1M)",     0.15,             0.00,              0.00,
  "Amazon Marketplace", "FBM (self-fulfill)", 0.15,          0.00,              0.03,
  "Amazon Marketplace", "FBA (Amazon fulfill)", 0.15,        0.20,              0.03,
  "Food Delivery A",    "Restaurant",       0.28,             0.00,              0.02,
  "Food Delivery B",    "Restaurant",       0.32,             0.00,              0.02
) |>
  mutate(
    total_fee = base_commission + fulfillment_fee + payment_fee,
    total_fee_pct = scales::percent(total_fee, accuracy = 0.1)
  )

# Stacked bar chart
fees_long <- fees |>
  pivot_longer(cols = c(base_commission, fulfillment_fee, payment_fee),
               names_to = "fee_type", values_to = "rate") |>
  mutate(
    fee_type = factor(fee_type,
                     levels = c("payment_fee", "fulfillment_fee", "base_commission"),
                     labels = c("Payment Processing", "Fulfillment", "Base Commission"))
  )

ggplot(fees_long, aes(x = paste(platform, category, sep = "\n"),
                      y = rate, fill = fee_type)) +
  geom_col(width = 0.7) +
  geom_text(data = fees,
            aes(x = paste(platform, category, sep = "\n"),
                y = total_fee, label = total_fee_pct, fill = NULL),
            vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(
    values = c("Base Commission" = "#D55E00",
               "Fulfillment" = "#0072B2",
               "Payment Processing" = "#009E73")
  ) +
  labs(
    title = "Platform Fee Structures by Category",
    subtitle = "Total take rate shown at top of each bar",
    x = NULL,
    y = "Fee Rate",
    fill = "Fee Component",
    caption = "Data from public platform documentation (2024). FBM = Fulfilled by Merchant; FBA = Fulfilled by Amazon."
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Summary table
cat("\nFee structure summary:\n")
fees |>
  select(platform, category, total_fee) |>
  mutate(total_fee = scales::percent(total_fee, accuracy = 0.1)) |>
  print(n = Inf)
```

**Key insights:**
- **App stores**: 30% standard (15% for small developers under $1M revenue).
- **E-commerce marketplaces**: 15-18% base + 15-25% fulfillment (if using platform logistics).
- **Food delivery**: 28-32% commission + 2-3% payment processing.
- **Bundling**: Platforms often bundle services (payment, fulfillment, advertising) making switching costly.

**Competitive assessment:**
- Compare fees across platforms to evaluate competitive pressure.
- Assess whether high fees reflect service value or market power.
- Evaluate whether fee structures create barriers to multi-homing.

### Market share evolution for digital platforms
Track platform market share over time to identify tipping points and competitive dynamics.

```{r}
#| label: digital-market-evolution
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 7
library(dplyr)
library(ggplot2)
library(lubridate)

# Market share evolution from public sources and regulatory filings
# Illustrative data for search engines, social media, or e-commerce
market_evolution <- tibble::tribble(
  ~date,         ~platform,      ~share,
  "2015-01-01",  "Platform A",   0.88,
  "2016-01-01",  "Platform A",   0.90,
  "2017-01-01",  "Platform A",   0.91,
  "2018-01-01",  "Platform A",   0.92,
  "2019-01-01",  "Platform A",   0.91,
  "2020-01-01",  "Platform A",   0.90,
  "2021-01-01",  "Platform A",   0.89,
  "2022-01-01",  "Platform A",   0.88,
  "2023-01-01",  "Platform A",   0.87,
  "2015-01-01",  "Platform B",   0.08,
  "2016-01-01",  "Platform B",   0.07,
  "2017-01-01",  "Platform B",   0.06,
  "2018-01-01",  "Platform B",   0.05,
  "2019-01-01",  "Platform B",   0.06,
  "2020-01-01",  "Platform B",   0.07,
  "2021-01-01",  "Platform B",   0.08,
  "2022-01-01",  "Platform B",   0.09,
  "2023-01-01",  "Platform B",   0.10,
  "2015-01-01",  "Others",       0.04,
  "2016-01-01",  "Others",       0.03,
  "2017-01-01",  "Others",       0.03,
  "2018-01-01",  "Others",       0.03,
  "2019-01-01",  "Others",       0.03,
  "2020-01-01",  "Others",       0.03,
  "2021-01-01",  "Others",       0.03,
  "2022-01-01",  "Others",       0.03,
  "2023-01-01",  "Others",       0.03
) |>
  mutate(date = as.Date(date))

# Area chart
ggplot(market_evolution, aes(x = date, y = share, fill = platform)) +
  geom_area(alpha = 0.8) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_fill_manual(
    values = c("Platform A" = "#0072B2",
               "Platform B" = "#009E73",
               "Others" = "#999999")
  ) +
  labs(
    title = "Digital Platform Market Share Evolution",
    subtitle = "Example: Search engine market shares 2015-2023",
    x = NULL,
    y = "Market Share",
    fill = "Platform",
    caption = "Illustrative data. Replace with StatCounter, SimilarWeb, or regulatory filings."
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Summary statistics
cat("\nMarket concentration trends:\n")
cat(paste0("2015 HHI: ", round(sum((market_evolution$share[
  market_evolution$date == "2015-01-01"] * 100)^2), 0), "\n"))
cat(paste0("2023 HHI: ", round(sum((market_evolution$share[
  market_evolution$date == "2023-01-01"] * 100)^2), 0), "\n"))
cat(paste0("\nPlatform A share trend: ",
          scales::percent(market_evolution$share[
            market_evolution$date == "2015-01-01" &
            market_evolution$platform == "Platform A"], accuracy = 0.1),
          " (2015) → ",
          scales::percent(market_evolution$share[
            market_evolution$date == "2023-01-01" &
            market_evolution$platform == "Platform A"], accuracy = 0.1),
          " (2023)\n"))
```

**Data sources:**
- **StatCounter**: Global browser, OS, search engine stats
- **SimilarWeb**: Website and app traffic data
- **Sensor Tower / App Annie**: Mobile app market intelligence
- **Regulatory filings**: EC, CMA, FTC, DMA compliance reports
- **OIPMI**: South African platform market shares

**Use cases:**
- Demonstrate sustained dominance or competitive entry
- Identify tipping points or competitive interventions
- Support market definition and dominance assessments

## Looking ahead
Store all digital platform analyses in `data/derived/digital/` with clear provenance notes. When DMA compliance data become available, replace synthetic Sankey flows and choice-screen impacts with actual telemetry. Document API access restrictions and data-sharing agreements in `data/README.md`. Cross-reference these visualizations with Chapter 12 (litigation practice) for expert report templates and Chapter 13 (empirical appendix) for diagnostic checklists.
