# Digital Markets and Platforms {#sec-digital-markets}

The tools and frameworks from the preceding chapters---market definition (@sec-market-definition), IO modeling (@sec-io-toolkit), merger analysis (@sec-mergers), and monopolization doctrine (@sec-monopolization)---apply to digital markets, but with distinctive challenges. Multi-sided platforms, network effects, zero-price business models, and rapid technological change all complicate traditional antitrust analysis. This chapter shows how to adapt the general toolkit to these specific features.

We also introduce new evidence types that matter especially in digital markets: telemetry data, clickstream analysis, API documentation, and product requirements documents. The intersection of technical product knowledge and economic reasoning defines much of the frontier in digital platform enforcement, from the US cases against Google and Meta to the EU's Digital Markets Act and South Africa's Online Intermediation Platforms Market Inquiry.

## Learning goals
Digital markets combine traditional IO concepts with product design, governance, and behavioral nudges. Drawing on DMA compliance materials, the UK's digital-markets regime under the Digital Markets, Competition and Consumers Act 2024, and recent enforcement actions, this chapter helps you:

- Analyze multi-sided participation, indirect network effects, and multi-homing incentives.
- Quantify self-preferencing, ranking changes, and default effects on user/merchant behavior.
- Evaluate data advantages, access restrictions, and algorithmic pricing risks.
- Integrate product requirement documents (PRDs), API policies, and telemetry with econometric evidence for global enforcement settings.

Expect to combine telemetry, clickstream data, partner contracts, and qualitative product documentation with econometric tests and simulation scaffolds.

::: {.callout-note title="Key acronyms"}
**OIPMI** = Online Intermediation Platforms Market Inquiry (South African Competition Commission, 2020–2023). An investigation into digital platform competition covering marketplaces, app stores, food delivery, travel booking, and software platforms. Provides transaction data, commission structures, and market share analyses referenced throughout this chapter.
:::

## Core topics

In June 2017, the European Commission fined Google 2.42 billion euros for systematically favoring its own comparison-shopping service in search results---a decision that launched a decade of platform antitrust. The case turned on a deceptively simple empirical question: did Google's algorithmic ranking decisions steer traffic away from rival comparison-shopping sites and toward Google Shopping? The answer required analyzing terabytes of clickstream data, modeling user behavior across search result positions, and disentangling algorithmic design choices from organic quality differences.

The South African Competition Commission's OIPMI inquiry (2020--2023) confronted parallel questions in a developing-market context: whether Takealot's marketplace algorithm favored its own retail arm, whether food-delivery commission structures foreclosed small restaurants, and whether Google's search dominance warranted the same remedies in Johannesburg as in Brussels. The US DOJ's monopolization suits against Google followed — and both have now produced liability findings, in search (D.D.C. 2024) and in advertising technology (E.D. Va. 2025). Digital platform enforcement is global, with common empirical challenges that recur across jurisdictions even as legal standards diverge.

Digital platform analysis extends traditional IO tools to multi-sided markets where network effects, data advantages, and zero-price business models create distinct competitive dynamics. This section covers the key mechanisms and evidence types.

**Multi-sided markets** connect different user groups (buyers/sellers, users/advertisers, drivers/riders) who value each other's participation:

- **Participation and multi-homing**: How do users on each side respond to fees, quality, and participation on other sides?
- **Network effects**: Do more users on one side attract more users on other sides? How strong is the feedback loop?
- **Switching costs and lock-in**: What prevents users from moving to rival platforms?

**Platform conduct concerns** [see @khan_2017 for the foundational critique of consumer welfare standards in platform markets; @stigler_committee_2019 for the policy reform agenda] include:

- **Self-preferencing**: Does the platform favor its own products in search rankings or recommendations?
- **Ranking manipulation**: Do algorithm changes benefit the platform at users' expense?
- **Default tying**: Are defaults structured to foreclose rivals?
- **Data leverage**: Does the platform use seller/developer data to compete against them?

**The platform case workflow** proceeds through five stages:

1. Define sides and metrics (MAU, GMV, multi-homing rates)
2. Map fees and monetization (commissions, ad loads, data collection)
3. Identify theory of harm (self-preferencing, foreclosure, data leverage)
4. Measure mechanisms (ranking impacts, default effects, fee pass-through)
5. Assess effects and design remedies (choice screens, access mandates, portability)

## Platform workflow overview

::: {.callout-note title="Platform Competition Analysis Workflow"}
```
SCOPING                   ANALYSIS                  REMEDIES
   |                          |                         |
   v                          v                         v
+--------------+        +--------------+        +--------------+
| Define       |        | Quantify     |        | Design       |
| sides &      |------->| mechanisms:  |------->| remedies:    |
| metrics      |        | - Ranking    |        | - Choice     |
|              |        | - Defaults   |        |   screens    |
+--------------+        | - Fees       |        | - API access |
   |                    +--------------+        | - Fair       |
   v                          |                |   ranking    |
+--------------+              v                +--------------+
| Map          |        +--------------+
| monetization:|        | Test         |
| - Commissions|        | theories:    |
| - Ad loads   |        | - DiD/Event  |
| - Data       |        | - A/B logs   |
+--------------+        | - Telemetry  |
   |                    +--------------+
   v
+--------------+
| Identify     |
| theory of    |
| harm:        |
| - Self-pref  |
| - Foreclosure|
| - Data lever.|
+--------------+
```
**Data sources:** Platform telemetry | DMA compliance | OIPMI surveys | Partner contracts | API documentation
:::

1. **Define sides and metrics.** Identify user, merchant, advertiser, developer, courier, or publisher sides. Track MAU/DAU, GMV, multi-homing rates, engagement depth, and conversion metrics.
2. **Map fees and monetization.** Build ladders for commissions, subscriptions, ad loads, and data monetization; compute pass-through and elasticity. For two-sided market theory, see @rochet_tirole_2003 and @armstrong_2006; the demand estimation framework in @sec-io-toolkit provides the empirical foundation for these models.
3. **Theory of harm.** Common mechanisms include self-preferencing, default tying, MFN/parity clauses, data leveraging, interoperability restrictions, algorithmic ranking biases, and algorithmic collusion.
4. **Mechanism measurement.** Quantify ranking impacts, default effects, fee changes, and data advantages; document multi-homing constraints and API access terms.
5. **Effects & remedies.** Measure partner churn, user switching, price/fee changes, and innovation impacts. Propose remedies (choice screens, fair-ranking reports, API access rules, data portability, structural separation) tied to measured harms.

## Theory-to-metric mapping

The table below maps common platform theories of harm to measurable metrics and appropriate empirical tests. Use it as an investigation scoping checklist when you first encounter a digital markets matter. Not every column will be relevant---prioritize tests where identification is strongest given the data you can actually obtain, and note where platform telemetry (which requires cooperation or compulsion to access) is essential versus where publicly available data suffices.

| Theory of Harm | Key Metric | Empirical Test | Data Sources |
|:---------------|:-----------|:---------------|:-------------|
| **Self-preferencing** | Downstream traffic share; CTR by listing type | Diff-in-Diff around algorithm changes; A/B test analysis | Platform telemetry, OIPMI audits, CMA studies |
| **Ranking manipulation** | Position-adjusted CTR; conversion by slot | Regression of outcomes on ranking position × first-party indicator | Clickstream logs, buy-box audits |
| **Default tying** | Rival market share pre/post choice screen | Event study around remedy implementation | DMA compliance reports, browser/search share data |
| **Foreclosure (input denial)** | API latency; error rates; access denial frequency | Regression of 3rd-party performance on access restrictions | API logs, developer complaints, uptime monitoring |
| **Foreclosure (raising rivals' costs)** | Fee differential; cost-to-serve by channel | Comparison of platform vs. direct channel costs | Fee schedules, merchant P&L data |
| **Data leverage** | Copycat product launch timing; feature matching | Survival analysis of 3rd-party products post-data access | Product launch dates, feature audits, internal documents |
| **MFN/parity clauses** | Price dispersion across platforms | Variance analysis; price comparison pre/post MFN removal | Price scraping, platform listings, DMA studies |
| **Algorithmic collusion** | Price correlation; response times; convergence patterns | Granger causality; structural break tests; simulation (cf. cartel screens in @sec-cartels) | Transaction-level pricing data, algorithm audits |
| **Switching costs/lock-in** | Multi-homing rates; churn by tenure | Duration models; switching cost estimation | Survey data, telemetry, DMA surveys |
| **Network effects exploitation** | Participation elasticity; tipping point dynamics | Two-sided demand estimation; natural experiments | Platform growth data, entry/exit events |

**How to use this table:**

1. **Scoping phase:** Identify which theories of harm are plausible given the market context and available complaints.
2. **Data requests:** Use the "Data Sources" column to structure information requests to platforms and third parties.
3. **Analysis design:** Match empirical tests to available data; prioritize tests with strongest identification.
4. **Triangulation:** Combine quantitative evidence with qualitative sources (internal documents, partner interviews, PRDs).

::: {.callout-note title="Adapting to zero-price markets"}
When users pay with attention rather than money, adapt metrics accordingly:

- Replace `fees_users` with **ad load** (ads per session, ad-to-content ratio)
- Use **time spent** or **engagement depth** as the "price" users pay
- Apply SSNDQ framework: would a hypothetical monopolist profitably degrade **quality** by 5-10%?
- Track quality metrics: load times, relevance scores, content diversity, privacy erosion
:::

## Participation, multi-homing, and network effects

- **Participation curves:** Model each side's activity as a function of fees, quality, and participation on the other side. Use logit, linear, or Poisson models with cross-side terms [@evans_schmalensee_2007; @hagiu_wright_2015].  
- **Multi-homing rates:** Estimate the share of users or merchants active on rival platforms. OIPMI surveys and EU DMA filings include methodologies you can reuse [@parker_van_alstyne_2005].  
- **Match quality:** Track conversion rates, completion times, or satisfaction metrics per platform side to show how conduct affects outcomes.  
- **Elasticities:** Two-sided elasticities inform fee caps and remedy design (e.g., how merchant participation responds to lower commissions or improved ranking fairness).

### Two-sided demand scaffold
```{r}
#| label: digital-two-sided
#| fig-cap: "Two-sided platform participation. Synthetic data for illustration only."
#| message: false
#| warning: false
library(dplyr)
library(fixest)
source("program/R/helpers.R")

# Synthetic example; replace with platform telemetry
panel <- expand.grid(platform = c("Alpha","Beta"), period = 2018:2023) |>
  mutate(
    fees_users = runif(n(), 0, 4),
    fees_merchants = runif(n(), 15, 30),
    quality = runif(n(), 0.5, 1),
    users = exp(5 - 0.25 * fees_users + 0.4 * quality + rnorm(n(), 0, 0.2)),
    merchants = exp(4 - 0.12 * fees_merchants + 0.3 * users / 1e5 + rnorm(n(), 0, 0.2))
  )

user_model <- feols(log(users) ~ fees_users + quality + log(merchants) | platform + period, data = panel)
merchant_model <- feols(log(merchants) ~ fees_merchants + log(users) | platform + period, data = panel)
summary(user_model)
summary(merchant_model)
```
Replace the synthetic `panel` with sanitized telemetry (monthly active users/merchants) or aggregated OIPMI data stored in `data/examples/digital-two-sided.csv`. Document the estimated cross-side elasticities in your memo.

### Zero-price demand model (attention markets)

When users pay nothing in monetary terms, traditional demand estimation breaks down---there is no price variation to exploit for identification. But users do pay, in attention, time, data, and exposure to advertising. The framework below adapts standard demand tools to these non-monetary costs, treating ad load, data collection intensity, and quality degradation as the "prices" that platforms charge on the user side. This aligns with the SSNDQ (Small but Significant Non-transitory Decrease in Quality) framework that competition authorities increasingly use as a substitute for SSNIP in zero-price markets.

**The challenge:** Many platforms don't charge users money---instead, users "pay" with attention (watching ads) and data (surrendering personal information). Standard price-based demand models don't directly apply.

**The solution:** Replace monetary fees with **attention costs**: ad load, time spent, or quality degradation. This lets us estimate how sensitive users are to non-price dimensions of competition and apply the SSNDQ (Small but Significant Non-transitory Decrease in Quality) framework used by competition authorities.

**What this model shows:** We estimate how user engagement (daily active users) responds to "attention costs" like ad load and data collection, as well as quality metrics like load time and content relevance. This lets us simulate whether a hypothetical monopolist could profitably degrade quality.

```{r}
#| label: digital-zero-price
#| message: false
#| warning: false
library(dplyr)
library(fixest)
source("program/R/helpers.R")

# Zero-price platform: users "pay" with attention (ad exposure)
# Synthetic example; replace with platform telemetry
set.seed(789)
attention_panel <- expand.grid(
  platform = c("Alpha", "Beta"),
  period = 2018:2023
) |>
  mutate(
    # Attention costs (ads per session, data collection intensity)
    ad_load = runif(n(), 3, 12),        # Ads per session
    data_collection = runif(n(), 0, 1), # Privacy cost index (0-1)

    # Quality metrics
    load_time_ms = runif(n(), 500, 3000),   # Page load time
    relevance_score = runif(n(), 0.5, 0.95), # Content relevance

    # User engagement (dependent variable)
    # Users respond negatively to ads/data collection, positively to quality
    daily_active_users = exp(
      12 - 0.08 * ad_load - 0.5 * data_collection -
      0.0003 * load_time_ms + 2 * relevance_score +
      rnorm(n(), 0, 0.15)
    ),

    # Advertiser side responds to user base
    advertiser_spend = exp(
      8 + 0.6 * log(daily_active_users) + 0.1 * ad_load +
      rnorm(n(), 0, 0.1)
    )
  )

# User-side model: attention cost elasticity
user_model <- feols(
  log(daily_active_users) ~ ad_load + data_collection +
    load_time_ms + relevance_score | platform + period,
  data = attention_panel
)

# Advertiser-side model: cross-side effects
advertiser_model <- feols(
  log(advertiser_spend) ~ log(daily_active_users) + ad_load | platform + period,
  data = attention_panel
)

cat("=== ZERO-PRICE DEMAND MODEL ===\n\n")
cat("User-side (attention cost) elasticities:\n")
cat(paste0("  Ad load effect: ", round(coef(user_model)["ad_load"], 4),
           " (1 more ad/session → ",
           round(coef(user_model)["ad_load"] * 100, 1), "% DAU change)\n"))
cat(paste0("  Data collection effect: ", round(coef(user_model)["data_collection"], 4),
           " (privacy cost)\n"))
cat(paste0("  Load time effect: ", round(coef(user_model)["load_time_ms"], 6),
           " (per ms)\n"))
cat(paste0("  Relevance effect: ", round(coef(user_model)["relevance_score"], 4),
           " (quality boost)\n\n"))

cat("Advertiser-side elasticities:\n")
cat(paste0("  User base elasticity: ", round(coef(advertiser_model)["log(daily_active_users)"], 3),
           " (cross-side network effect)\n"))
cat(paste0("  Ad inventory value: ", round(coef(advertiser_model)["ad_load"], 4),
           " (more slots → more spend)\n"))

# SSNDQ analysis: what if quality degrades 5-10%?
cat("\n=== SSNDQ SIMULATION ===\n")
baseline_dau <- mean(attention_panel$daily_active_users)
# 10% quality degradation = 10% more ads + 10% worse relevance
degraded_dau <- baseline_dau * exp(
  coef(user_model)["ad_load"] * 0.8 +      # 10% more ads (e.g., 8 → 8.8)
  coef(user_model)["relevance_score"] * -0.08  # 10% worse relevance
)
dau_loss <- (baseline_dau - degraded_dau) / baseline_dau

cat(paste0("Baseline DAU: ", format(round(baseline_dau), big.mark = ","), "\n"))
cat(paste0("After 10% quality degradation: ", format(round(degraded_dau), big.mark = ","), "\n"))
cat(paste0("Predicted user loss: ", scales::percent(dau_loss, accuracy = 0.1), "\n"))
cat(paste0("Would monopolist profit? Depends on ad revenue vs. user loss tradeoff.\n"))
```

**Key adaptations for zero-price markets:**

| Traditional Market | Zero-Price Market | Measurement |
|:-------------------|:------------------|:------------|
| Price | Ad load (ads/session) | Platform telemetry |
| Price | Time cost (session duration) | Engagement metrics |
| Price | Privacy cost (data collection) | Privacy policy scoring |
| Quality | Relevance, load time, UX | A/B tests, user surveys |
| Quantity | DAU, MAU, sessions | Platform analytics |

**SSNDQ framework application:**

1. Define quality metrics relevant to users (speed, relevance, privacy, content diversity)
2. Estimate user sensitivity to each quality dimension
3. Simulate: would a hypothetical monopolist profitably degrade quality by 5-10%?
4. If users wouldn't switch despite degradation → market power concern

:::{.callout-note title="Case box: European Commission v. Google Shopping (2017)"}
The Commission found that Google systematically positioned its own comparison shopping service at the top of search results while demoting rival comparison shopping sites through algorithmic adjustments [@eu_google_shopping_2017]. Position drove the harm: the Commission's evidence showed the top generic search result receives roughly 35% of clicks, with click-through rates falling steeply down the page. The €2.42 billion fine---the largest antitrust penalty at the time---rested on empirical analysis of traffic differentials; in the United Kingdom, the decision found, rival comparison shopping sites lost around 85% of their traffic after the algorithmic changes while traffic to Google's own service increased roughly 45-fold. The remedy required Google to give rival comparison services equal treatment through an auction-based mechanism for product listing ad slots. The General Court upheld the decision in 2021, and the Court of Justice dismissed Google's further appeal in September 2024 (C-48/22 P), definitively confirming the fine and that self-preferencing by a dominant platform can constitute abuse even absent a traditional leveraging theory.
:::

:::{.callout-note title="Case box: Epic Games v. Apple (2021)"}
Epic challenged Apple's 30% App Store commission and anti-steering rules that prevented developers from directing users to cheaper external payment options [@us_epic_apple_2021]. Judge Gonzalez Rogers found Apple was *not* a monopolist under the Sherman Act, partly because the relevant market was defined as "digital mobile gaming transactions"---a two-sided market where Apple competed with Google Play and console platforms. However, Apple violated California's Unfair Competition Law by preventing developers from communicating alternative payment methods to users. The injunction requiring Apple to allow external payment links reshaped app store economics globally. The case demonstrated the stakes of market definition in platform cases: a narrower single-sided market (iOS app distribution) might have yielded a monopoly finding, while the broader two-sided framing shielded Apple from Section 2 liability. The injunction also grew teeth: in April 2025 the same judge found Apple in willful violation of the anti-steering order, barred the 27% commission Apple had imposed on off-app purchases and its "scare screens" around external links, and referred the matter for possible criminal contempt.
:::

:::{.callout-note title="Where the platform cases stand (mid-2026)"}
The headline investigations this chapter discusses now have outcomes.

- ***United States v. Google* (search), D.D.C.** August 2024: Judge Mehta found Google unlawfully maintained monopolies in general search services and general search text ads through exclusive default agreements. The September 2025 remedies opinion declined the government's request to divest Chrome; it barred exclusive default and preinstallation agreements, required Google to share certain search index and user-interaction data with qualified competitors, and installed technical-committee oversight. Appeals were pending as of this writing.
- ***United States v. Google* (ad tech), E.D. Va.** April 2025: Judge Brinkema found Section 2 liability in the publisher ad server and ad exchange markets and unlawful tying of DFP to AdX; the advertiser ad network claim failed. Remedies proceedings followed.
- ***FTC v. Meta*, D.D.C.** November 2025: judgment for Meta. Judge Boasberg held the FTC failed to prove that Meta currently holds monopoly power in the alleged personal-social-networking market.
- **DMA enforcement.** The Commission issued its first non-compliance fines in April 2025 --- €500 million against Apple (App Store anti-steering) and €200 million against Meta ("consent or pay") --- and adopted interoperability specification decisions against Apple in March 2025.
- **Google Shopping.** The CJEU dismissed Google's appeal in September 2024 (C-48/22 P), upholding the €2.42 billion fine.
- **United Kingdom.** The Digital Markets, Competition and Consumers Act 2024 brought the digital-markets regime into force in January 2025. The CMA opened strategic-market-status investigations into Google's general search business and the Apple and Google mobile ecosystems in 2025 and, as of this writing, has begun making designations.
:::

## Ranking, self-preferencing, and defaults

### Ranking impact visualization

Ranking position is the central competitive lever on digital platforms. Empirical studies consistently find that the top search result captures 25--35% of clicks, while the tenth result captures less than 3%. Control over ranking is, in effect, control over demand---a point that explains why self-preferencing allegations focus so heavily on algorithmic ranking decisions. The visualization below quantifies this leverage using click-through rate data by position, showing how first-party listings systematically outperform third-party listings even at the same slot.

```{r}
#| label: digital-ranking
#| message: false
#| warning: false
library(dplyr)
library(tidyr)
library(ggplot2)

ranking <- tibble::tribble(
  ~slot, ~first_party_ctr, ~third_party_ctr,
  1, 0.16, 0.08,
  2, 0.13, 0.06,
  3, 0.11, 0.05,
  4, 0.08, 0.04,
  5, 0.06, 0.03
) |>
  pivot_longer(-slot, names_to = "listing", values_to = "ctr")

ggplot(ranking, aes(slot, ctr, color = listing)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("#1b9e77", "#d95f02"), labels = c("First-party", "Third-party")) +
  labs(
    title = "Conversion by ranking slot",
    subtitle = "Illustrative data; use OIPMI or platform A/B metrics when available",
    x = "Slot (1 = top)",
    y = "Conversion rate",
    color = NULL
  ) +
  theme_antitrust() +
  theme(legend.position = "bottom")
```
Populate the tibble with actual ranking data (e.g., OIPMI buy-box audits, CMA Amazon Marketplace analysis) or maintain sanitized values in `data/examples/digital-ranking.csv`. The CMA's market studies are publicly available at [gov.uk/cma](https://www.gov.uk/government/organisations/competition-and-markets-authority).

### Default-choice event study
```{r}
#| label: digital-default-event
#| message: false
#| warning: false
library(fixest)
library(dplyr)
library(ggplot2)
source("program/R/helpers.R")

# Synthetic panel: 50 markets over 20 months, choice screen at month 11
set.seed(42)
n_markets <- 50
n_periods <- 20
treat_period <- 11

panel <- expand.grid(market_id = 1:n_markets, month = 1:n_periods) |>
  mutate(
    treated = market_id <= 25,  # Half get choice screen
    post = month >= treat_period,
    rel_month = month - treat_period,
    # Default search engine share: drops for treated markets post-intervention
    default_share = 0.92 - 0.001 * month +
      ifelse(treated & post, -0.07, 0) +
      rnorm(n(), 0, 0.015)
  )

# Event study with market and period FEs
es <- feols(default_share ~ i(rel_month, treated, ref = -1) |
              market_id + month, data = panel)

# Extract and plot coefficients
coefs <- broom::tidy(es, conf.int = TRUE) |>
  filter(grepl("rel_month", term)) |>
  mutate(rel_month = as.numeric(gsub("rel_month::(.*):treated::TRUE", "\\1", term))) |>
  bind_rows(tibble(rel_month = -1, estimate = 0, conf.low = 0, conf.high = 0))

ggplot(coefs, aes(x = rel_month, y = estimate)) +
  geom_hline(yintercept = 0, color = "gray40") +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "gray50") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, fill = "#0072B2") +
  geom_line(color = "#0072B2", linewidth = 1) +
  geom_point(color = "#0072B2", size = 2.5) +
  annotate("text", x = 0, y = 0.01, label = "Choice screen\nintroduced",
           hjust = -0.1, size = 3.5, fontface = "italic") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Default search engine share after choice screen",
    subtitle = "Event study: treated vs. control markets (synthetic data)",
    x = "Months relative to choice screen",
    y = "Effect on default engine share",
    caption = "Replace with DMA compliance data or CMA choice screen experiments."
  ) +
  theme_antitrust()
```
Replace the synthetic panel with DMA compliance reports, CMA choice screen experiments, or telemetry from browser/search default changes.

## Methodologies

::: {.callout-note title="Method box"}
- Platform fee/price pass-through across sides.  
- Event studies on ranking/default changes, following the design-based approach in @sec-research-design; DMA-style interventions.  
- Two-sided logit or bargaining sketches to test fee caps and interoperability remedies.
:::

::: {.callout-note title="Qualitative evidence"}
- Product requirement documents, experiment summaries, and governance memos.  
- Partner contracts (parity/MFN, exclusivity), API access terms, data-sharing policies.  
- User research/surveys capturing multi-homing, switching costs, and friction sources.
:::

::: {.callout-note title="Citations and comparative note"}
- US cases [@us_epic_apple_2021; @us_google_search_2023; @us_google_adtech_2023], EU/DMA [@eu_google_shopping_2017; @eu_google_android_2018; @eu_dma_2022], UK (CMA digital market studies), KFTC/JFTC app store cases, SAMR digital enforcement.
- Empirical studies on defaults, ranking effects, and multi-homing [e.g., @luca_2016; @edelman_lai_2016].
- Flag differences in parity/MFN treatment and remedy preferences (EU fair-ranking obligations vs. US injunctive relief).
:::

### Algorithmic collusion: a Q-learning simulation

One of the most distinctive challenges in digital market competition is the possibility that pricing algorithms---even when independently developed and deployed---may learn to coordinate on supra-competitive prices without any explicit communication between firms. This is not hypothetical: in simulation, Q-learning pricing agents playing a repeated Bertrand game systematically converge to prices well above the one-shot Nash equilibrium and sustain them with punishment-and-recovery responses to deviations [@calvano_etal_2020; @oecd_algorithms_2017].

The mechanism is straightforward. Each algorithm observes the current state (the prices set in the previous period), chooses an action (a price from a discrete grid), and receives a reward (profit). Over repeated iterations, the algorithm learns a Q-table mapping states to actions that maximizes cumulative reward. When two such algorithms interact, they discover that mutual price increases yield higher rewards than price competition---and they learn to punish deviations by reverting toward competitive pricing, creating a self-enforcing collusive pattern that mirrors the trigger strategies of repeated-game theory.

```{r}
#| label: digital-algorithmic-collusion
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6
library(dplyr)
library(ggplot2)
library(zoo)
source("program/R/helpers.R")

# Q-learning pricing simulation: two firms, repeated Bertrand game
# Demand: q_i = 100 - 4*p_i + 2*p_j, marginal cost c = 5.
# On the grid {20, 24, 28}, p = 20 is each firm's best response to EVERY
# rival price (one-shot profits: 900 vs. 836/644 against a rival at 20), so
# (20, 20) is the unique stage-game Nash. Symmetric pricing at 28 earns
# 1012 each -- more than Nash -- but a unilateral cut to 20 earns 1140, so
# sustaining 28 requires learned punishment. That makes 28 genuinely
# collusive, not just another static equilibrium.

set.seed(42)
n_iterations <- 20000
price_levels <- c(20, 24, 28)  # Nash, intermediate, collusive
n_prices <- length(price_levels)
p_nash <- 20
p_collusive <- 28

demand <- function(p_i, p_j) max(0, 100 - 4 * p_i + 2 * p_j)
profit <- function(p_i, p_j) demand(p_i, p_j) * (p_i - 5)

# Two independent learners. Each firm keeps its OWN Q-matrix:
# rows index the state (last period's price pair), columns the action
# (this period's own price).
q_learning_sim <- function(n_iter, lr = 0.1, discount = 0.95,
                            eps_start = 0.5, eps_end = 0.001) {
  n_states <- n_prices^2
  Q1 <- matrix(0, nrow = n_states, ncol = n_prices)
  Q2 <- matrix(0, nrow = n_states, ncol = n_prices)

  state_index <- function(a_own, a_rival) (a_own - 1) * n_prices + a_rival

  choose_action <- function(Q, s, eps) {
    if (runif(1) < eps) return(sample.int(n_prices, 1))
    best <- which(Q[s, ] == max(Q[s, ]))
    if (length(best) > 1) best <- sample(best, 1)
    best
  }

  prices_out <- matrix(0, nrow = n_iter, ncol = 2)
  eps <- eps_start
  eps_decay <- (eps_end / eps_start)^(1 / n_iter)

  # Random initial price pair
  a1 <- sample.int(n_prices, 1)
  a2 <- sample.int(n_prices, 1)
  s1 <- state_index(a1, a2)  # firm 1's view: (own last, rival last)
  s2 <- state_index(a2, a1)  # firm 2's view

  for (t in 1:n_iter) {
    a1_new <- choose_action(Q1, s1, eps)
    a2_new <- choose_action(Q2, s2, eps)

    r1 <- profit(price_levels[a1_new], price_levels[a2_new])
    r2 <- profit(price_levels[a2_new], price_levels[a1_new])

    s1_new <- state_index(a1_new, a2_new)
    s2_new <- state_index(a2_new, a1_new)

    Q1[s1, a1_new] <- Q1[s1, a1_new] +
      lr * (r1 + discount * max(Q1[s1_new, ]) - Q1[s1, a1_new])
    Q2[s2, a2_new] <- Q2[s2, a2_new] +
      lr * (r2 + discount * max(Q2[s2_new, ]) - Q2[s2, a2_new])

    s1 <- s1_new
    s2 <- s2_new
    prices_out[t, ] <- price_levels[c(a1_new, a2_new)]
    eps <- eps * eps_decay
  }
  prices_out
}

results <- q_learning_sim(n_iterations)

collusion_data <- tibble(
  iteration = 1:n_iterations,
  firm1 = results[, 1],
  firm2 = results[, 2],
  avg = (firm1 + firm2) / 2,
  rolling = zoo::rollmean(avg, k = 500, fill = NA)
)

ggplot(collusion_data, aes(x = iteration)) +
  geom_line(aes(y = firm1), color = "#0072B2", alpha = 0.15, linewidth = 0.3) +
  geom_line(aes(y = firm2), color = "#D55E00", alpha = 0.15, linewidth = 0.3) +
  geom_line(aes(y = rolling), color = "#333333", linewidth = 1.2) +
  geom_hline(yintercept = p_nash, linetype = "dashed", color = "#009E73", linewidth = 0.8) +
  geom_hline(yintercept = p_collusive, linetype = "dashed", color = "#D55E00", linewidth = 0.8) +
  annotate("text", x = n_iterations * 0.96, y = p_nash - 0.5,
           label = "One-shot Nash (20)",
           color = "#009E73", hjust = 1, size = 3) +
  annotate("text", x = n_iterations * 0.96, y = p_collusive + 0.5,
           label = "Collusive price (28)",
           color = "#D55E00", hjust = 1, size = 3) +
  labs(
    title = "Algorithmic Collusion: Q-Learning Price Convergence",
    subtitle = "Two independent pricing algorithms, each with its own Q-matrix and no communication",
    x = "Iteration", y = "Price",
    caption = "Simulation: independent Q-learning agents in a repeated Bertrand game (q = 100 - 4p + 2p_rival, c = 5).\nBlack line: 500-period rolling average of the two firms' prices."
  ) +
  theme_antitrust() +
  theme(legend.position = "none")

early_mean <- mean(collusion_data$avg[1:1000])
late_mean <- mean(collusion_data$avg[(n_iterations - 999):n_iterations])
conv_ratio <- (late_mean - p_nash) / (p_collusive - p_nash)

cat("\nPrice convergence summary:\n")
cat(sprintf("First 1,000 iterations (exploration): mean price = %.2f\n", early_mean))
cat(sprintf("Last 1,000 iterations (after learning): mean price = %.2f\n", late_mean))
cat(sprintf("One-shot Nash on the grid: %.2f | Joint-profit price on the grid: %.2f\n",
            p_nash, p_collusive))
cat(sprintf("Convergence ratio: %.0f%% of the gap from Nash toward the collusive price\n",
            100 * conv_ratio))
```

**Interpretation.** Read the convergence ratio as the share of the Nash-to-collusion price gap the algorithms close: 0% means they settle at the one-shot Nash price of 20, 100% at the joint-profit price of 28. In repeated runs of this deliberately small simulation (three prices, 20,000 episodes), final prices typically land a third to a half of the way up the gap---supra-competitive, sustained by each agent's learned response of reverting toward 20 when its rival cuts price, though short of full monopoly pricing. The full-scale result is @calvano_etal_2020: with finer price grids and millions of training episodes, their Q-learning agents systematically converged on supra-competitive prices and enforced them with punishment-and-recovery schemes---a deviation triggers a price war, followed by a gradual return to the collusive level---without any communication between the agents.

**Antitrust implications.** This result challenges the traditional legal framework for cartel enforcement, which requires proof of an "agreement" or "concerted practice." Q-learning algorithms achieve collusive outcomes through independent optimization---there is no meeting of minds, no communication, no explicit coordination. Competition authorities are grappling with how to address this gap. The OECD's 2017 report on algorithms and collusion [@oecd_algorithms_2017] identified four scenarios: (1) messenger algorithms (humans collude, algorithms implement), (2) hub-and-spoke (common algorithm provider), (3) predictable agent (algorithms respond to each other's predictable behavior), and (4) autonomous machine (algorithms independently learn to collude). The first three fit within existing legal frameworks; the fourth does not. The EU's Digital Markets Act and the UK's digital-markets regime under the Digital Markets, Competition and Consumers Act 2024 (in force since January 2025) give regulators ex ante tools that do not depend on proving agreement, while US enforcement remains anchored in conspiracy doctrine.

### Southern African digital enforcement snapshots
- **Online Intermediation Platforms Market Inquiry (OIPMI, 2020–2023).** The Competition Commission collected transaction-level data from e-commerce marketplaces, app stores, food delivery apps, and travel platforms. Findings highlighted Google Search’s >90% share of general search queries, food-delivery commissions clustering at 25–30%, and marketplace buy-box algorithms that favored first-party listings (e.g., Takealot Retail) over third-party sellers with identical fulfillment metrics. Final remedies require fair-ranking reports, caps on marketplace storage/fulfillment fees for SMEs, and anti-steering relief that lets restaurants promote direct-order channels inside aggregator apps.
- **Advertising and digital news media (Media and Digital Platforms Market Inquiry, 2023–ongoing).** Building on OIPMI data, the Commission is ingesting impression-level logs from Meta, Google, and local publishers to quantify revenue imbalances in display advertising. The interim statement reports that Google commands roughly 65% of display ad spend routed through programmatic channels, while local publishers face average take-rates below 30% after demand-side/platform fees—data points that frame prospective bargaining remedies for news publishers across Southern Africa.
- **GovChat/Meta interim relief (Competition Commission v. Meta Platforms, 2021).** The Commission used WhatsApp Business API telemetry to demonstrate GovChat’s role in reaching 8+ million South African users for public-service messaging, and it showed how Meta’s proposed off-boarding would foreclose a civic-use case despite negligible incremental infrastructure cost. Interim relief preserved access while the Tribunal weighed the exclusion case, offering a blueprint for combining platform-side telemetry with public-interest arguments in developing markets.

## Enhanced Visualizations

### Multi-homing patterns
Multi-homing patterns reveal competitive constraints and switching costs. The following visualization shows the distribution of users across platform combinations, helping identify whether platforms compete head-to-head or serve distinct niches.

```{r}
#| label: digital-multihoming-plot
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 6
library(dplyr)
library(ggplot2)

# Simulated multi-homing survey data
# Replace with OIPMI survey data or DMA compliance reports
set.seed(234)
n_users <- 1000

multihoming <- tibble(
  user_id = 1:n_users,
  platform_a = sample(c("Active", "Inactive"), n_users, replace = TRUE,
                     prob = c(0.7, 0.3)),
  platform_b = sample(c("Active", "Inactive"), n_users, replace = TRUE,
                     prob = c(0.5, 0.5)),
  platform_c = sample(c("Active", "Inactive"), n_users, replace = TRUE,
                     prob = c(0.3, 0.7))
) |>
  mutate(
    pattern = case_when(
      platform_a == "Active" & platform_b == "Active" & platform_c == "Active" ~ "All three",
      platform_a == "Active" & platform_b == "Active" ~ "A + B",
      platform_a == "Active" & platform_c == "Active" ~ "A + C",
      platform_b == "Active" & platform_c == "Active" ~ "B + C",
      platform_a == "Active" ~ "A only",
      platform_b == "Active" ~ "B only",
      platform_c == "Active" ~ "C only",
      TRUE ~ "None"
    )
  )

# Aggregate flows
flows <- multihoming |>
  count(pattern) |>
  mutate(
    pct = n / sum(n),
    pattern = factor(pattern,
                    levels = c("A only", "B only", "C only",
                              "A + B", "A + C", "B + C", "All three", "None"))
  ) |>
  arrange(pattern)

# Bar chart showing multi-homing patterns
p1 <- ggplot(flows, aes(x = pattern, y = pct, fill = pattern)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1)),
            vjust = -0.5, size = 3.5) +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.1))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Multi-homing Patterns Across Platforms",
    subtitle = "User activity across Platform A, B, and C",
    x = NULL,
    y = "Share of Users",
    caption = "Illustrative data. Replace with OIPMI survey or DMA compliance reports."
  ) +
  theme_antitrust() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title.position = "plot"
  )

p1

# Summary statistics
cat("\nMulti-homing summary:\n")
cat(paste0("Single-homing (one platform only): ",
          scales::percent(sum(flows$pct[flows$pattern %in%
                              c("A only", "B only", "C only")]),
                         accuracy = 0.1), "\n"))
cat(paste0("Multi-homing (2+ platforms): ",
          scales::percent(sum(flows$pct[grepl("\\+|three", flows$pattern)]),
                         accuracy = 0.1), "\n"))
```

**Interpretation:**
-   **High single-homing**: Suggests strong lock-in, switching costs, or network effects. Platforms may have market power.
-   **High multi-homing**: Indicates competitive pressure; users actively compare options. But check if multi-homing is symmetric or if one platform dominates usage.
-   **Asymmetric patterns**: If Platform A appears in most combinations, it may be a "must-have" platform with gatekeeping power.

**Data sources:**
-   OIPMI survey microdata (South Africa)
-   DMA compliance reports (EU)
-   CMA market studies (UK)
-   Platform-provided telemetry (subject to confidentiality)

### Choice screen / default effect visualization
Quantify how defaults affect market shares using difference-in-differences or event studies around choice screen implementations.

### Self-preferencing detection exercise

Self-preferencing occurs when a platform uses its control over access to favor its own products or services over those of rivals. The EU's Google Shopping decision (@eu_google_shopping_2017) established that systematic demotion of rival comparison shopping services in search results constituted an abuse of dominance under Article 102 TFEU. Detecting self-preferencing requires comparing the platform's treatment of its own offerings against its treatment of rivals, controlling for quality differences.

The empirical approach involves three steps: (1) measure the platform's ranking or visibility of own vs. rival products; (2) control for quality differences (relevance, price, user ratings); and (3) test whether any residual favoritism is statistically and economically significant.

```{r}
#| label: digital-self-preferencing
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6
library(dplyr)
library(ggplot2)
library(fixest)
source("program/R/helpers.R")

# Simulated search ranking data
# Replace with actual platform ranking data from discovery or DMA reports
set.seed(456)
n_queries <- 500

ranking_data <- tibble(
  query_id = 1:n_queries,
  # Platform's own product vs. rival
  is_own = rbinom(n_queries, 1, 0.3),
  # Quality metrics (observable to the econometrician)
  relevance_score = runif(n_queries, 0, 1),
  user_rating = runif(n_queries, 1, 5),
  price_competitiveness = runif(n_queries, 0, 1),
  # Position assigned by platform (1 = top, higher = worse)
  # Own products get better positions, controlling for quality
  position = case_when(
    is_own == 1 ~ round(2 + 3 * (1 - relevance_score) + rnorm(n_queries, 0, 0.5)),
    TRUE ~ round(4 + 3 * (1 - relevance_score) + rnorm(n_queries, 0, 0.8))
  ),
  position = pmax(1, pmin(10, position))
)

# Regression: does being "own" improve ranking, controlling for quality?
rank_model <- feols(
  position ~ is_own + relevance_score + user_rating + price_competitiveness,
  data = ranking_data
)

cat("Self-preferencing test:\n")
cat("Coefficient on is_own (negative = better ranking for own products):\n")
cat(paste0("  β = ", round(coef(rank_model)["is_own"], 3), "\n"))
cat(paste0("  SE = ", round(sqrt(diag(vcov(rank_model)))["is_own"], 3), "\n"))
cat(paste0("  p-value = ", sprintf("%.4f", coef(summary(rank_model))["is_own", 4]), "\n\n"))

# Visualization: distribution of positions for own vs. rival products
ggplot(ranking_data, aes(x = position, fill = factor(is_own, labels = c("Rival", "Own")))) +
  geom_bar(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("Rival" = "#0072B2", "Own" = "#D55E00")) +
  scale_x_reverse(breaks = 1:10) +
  labs(
    title = "Self-Preferencing Detection: Position Distribution",
    subtitle = "Own products systematically ranked higher, controlling for quality",
    x = "Search Position (1 = top)", y = "Count of Products", fill = NULL,
    caption = "Simulated data. Replace with actual platform ranking data from discovery or DMA reports."
  ) +
  theme_antitrust() +
  theme(legend.position = "bottom")
```

**Interpreting the results:** A statistically significant negative coefficient on `is_own` indicates that the platform's own products receive better positions than rival products, even after controlling for relevance, user ratings, and price competitiveness. This is the pattern that competition authorities look for in self-preferencing cases. The magnitude of the coefficient tells you the average position advantage---a coefficient of -1.5 means own products are ranked 1.5 positions higher on average, which can translate into substantial traffic differences given the steep click-through rate gradient across search positions.

**Data sources:** Platform ranking logs (from discovery or regulatory requests), click-through rate data, user engagement metrics, and product quality scores. The EU's DMA requires designated gatekeepers to provide ranking data to the Commission, creating a new data source for self-preferencing analysis.

```{r}
#| label: digital-default-plot
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 5
library(fixest)
library(dplyr)
library(ggplot2)
library(patchwork)

# Simulated data: Android choice screen impact (EU DMA context)
# Replace with actual DMA compliance data or CMA experiments
set.seed(345)
weeks <- -24:24
countries <- c("Control (no choice screen)", "Treatment (choice screen)")

default_data <- expand.grid(
  week = weeks,
  country = countries
) |>
  mutate(
    # Pre-intervention: both countries similar
    # Post-intervention: treatment shows rival gain
    rival_share = case_when(
      country == "Control (no choice screen)" ~ 0.05 + 0.0005 * week + rnorm(n(), 0, 0.01),
      week < 0 ~ 0.05 + 0.0005 * week + rnorm(n(), 0, 0.01),
      TRUE ~ 0.05 + 0.0005 * week + 0.08 + rnorm(n(), 0, 0.01)
    ),
    rival_share = pmax(0.01, pmin(0.25, rival_share))
  )

# Time series plot
p1 <- ggplot(default_data, aes(x = week, y = rival_share,
                                color = country, linetype = country)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40",
             linewidth = 1) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  annotate("text", x = 0, y = 0.22, label = "Choice screen\nintroduced",
           hjust = -0.1, size = 3.5, fontface = "italic") +
  scale_color_manual(values = c("Control (no choice screen)" = "#999999",
                                 "Treatment (choice screen)" = "#0072B2")) +
  scale_linetype_manual(values = c("Control (no choice screen)" = "dashed",
                                   "Treatment (choice screen)" = "solid")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Impact of Choice Screen on Rival Search Engine Share",
    subtitle = "Difference-in-Differences: Treatment vs. Control Countries",
    x = "Weeks relative to choice screen introduction",
    y = "Rival search engine share",
    color = NULL,
    linetype = NULL
  ) +
  theme_antitrust() +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Coefficient plot (event study coefficients)
# Simulated DiD estimates
did_coefs <- tibble(
  week = weeks[weeks != -1]  # reference period
) |>
  mutate(
    estimate = 0.08 * as.numeric(week >= 0) + rnorm(n(), 0, 0.01),
    ci_lower = estimate - 1.96 * 0.01,
    ci_upper = estimate + 1.96 * 0.01
  )

p2 <- ggplot(did_coefs, aes(x = week, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40",
             linewidth = 1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
              alpha = 0.2, fill = "#0072B2") +
  geom_line(color = "#0072B2", linewidth = 1) +
  geom_point(color = "#0072B2", size = 2) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Event Study Coefficients",
    subtitle = "Treatment effect by week (relative to week -1)",
    x = "Weeks relative to choice screen",
    y = "Effect on rival share (percentage points)",
    caption = "Illustrative data. Replace with DMA compliance reports or CMA experiments."
  ) +
  theme_antitrust() +
  theme(plot.title.position = "plot")

# Combined plot
p1 / p2

# Summary statistics
pre_treatment <- mean(default_data$rival_share[
  default_data$country == "Treatment (choice screen)" & default_data$week < 0])
post_treatment <- mean(default_data$rival_share[
  default_data$country == "Treatment (choice screen)" & default_data$week >= 0])
effect <- post_treatment - pre_treatment

cat("\nChoice screen impact summary:\n")
tibble::tibble(
  metric = c("Pre-treatment share", "Post-treatment share", "Treatment Effect"),
  value = c(pre_treatment, post_treatment, effect)
) |>
  knitr::kable(digits = 3, caption = "Impact of Choice Screen")
```

**How to use this analysis:**
- **Pre-trends**: Check parallel trends before intervention to validate DiD assumptions.
- **Treatment effect**: Quantify how much rival share increased after choice screen.
- **Heterogeneity**: Analyze by device type, user demographics, or country.
- **Persistence**: Track whether effects sustain over time or decay.

**DMA context:**
The EU Digital Markets Act requires designated gatekeepers to offer choice screens for browsers, search engines, and other services. This analysis framework can evaluate compliance and effectiveness.

### Platform fee structure visualization
Commission and fee structures are central evidence in platform cases. Show how fees vary by seller tier, product category, or fulfillment method.

```{r}
#| label: digital-fee-structure
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 7
library(dplyr)
library(ggplot2)
library(tidyr)

# Fee structures from major platforms (public sources)
# App Store, Play Store, Amazon Marketplace, food delivery
fees <- tibble::tribble(
  ~platform,            ~category,          ~base_commission, ~fulfillment_fee, ~payment_fee,
  "App Store",          "Apps",             0.30,             0.00,              0.00,
  "App Store",          "Apps (< $1M)",     0.15,             0.00,              0.00,
  "Play Store",         "Apps",             0.30,             0.00,              0.00,
  "Play Store",         "Apps (< $1M)",     0.15,             0.00,              0.00,
  "Amazon Marketplace", "FBM (self-fulfill)", 0.15,          0.00,              0.03,
  "Amazon Marketplace", "FBA (Amazon fulfill)", 0.15,        0.20,              0.03,
  "Food Delivery A",    "Restaurant",       0.28,             0.00,              0.02,
  "Food Delivery B",    "Restaurant",       0.32,             0.00,              0.02
) |>
  mutate(
    total_fee = base_commission + fulfillment_fee + payment_fee,
    total_fee_pct = scales::percent(total_fee, accuracy = 0.1)
  )

# Stacked bar chart
fees_long <- fees |>
  pivot_longer(cols = c(base_commission, fulfillment_fee, payment_fee),
               names_to = "fee_type", values_to = "rate") |>
  mutate(
    fee_type = factor(fee_type,
                     levels = c("payment_fee", "fulfillment_fee", "base_commission"),
                     labels = c("Payment Processing", "Fulfillment", "Base Commission"))
  )

ggplot(fees_long, aes(x = paste(platform, category, sep = "\n"),
                      y = rate, fill = fee_type)) +
  geom_col(width = 0.7) +
  geom_text(data = fees,
            aes(x = paste(platform, category, sep = "\n"),
                y = total_fee, label = total_fee_pct, fill = NULL),
            vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(
    values = c("Base Commission" = "#D55E00",
               "Fulfillment" = "#0072B2",
               "Payment Processing" = "#009E73")
  ) +
  labs(
    title = "Platform Fee Structures by Category",
    subtitle = "Total take rate shown at top of each bar",
    x = NULL,
    y = "Fee Rate",
    fill = "Fee Component",
    caption = "Data from public platform documentation (2024). FBM = Fulfilled by Merchant; FBA = Fulfilled by Amazon."
  ) +
  theme_antitrust() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Summary table
cat("\nFee structure summary:\n")
fees |>
  select(platform, category, total_fee) |>
  mutate(total_fee = scales::percent(total_fee, accuracy = 0.1)) |>
  print(n = Inf)
```

**Key insights:**
- **App stores**: 30% standard (15% for small developers under $1M revenue).
- **E-commerce marketplaces**: 15-18% base + 15-25% fulfillment (if using platform logistics).
- **Food delivery**: 28-32% commission + 2-3% payment processing.
- **Bundling**: Platforms often bundle services (payment, fulfillment, advertising) making switching costly.

**Competitive assessment:**
- Compare fees across platforms to evaluate competitive pressure.
- Assess whether high fees reflect service value or market power.
- Evaluate whether fee structures create barriers to multi-homing.

### Market share evolution for digital platforms
Track platform market share over time to identify tipping points and competitive dynamics.

```{r}
#| label: digital-market-evolution
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 7
library(dplyr)
library(ggplot2)
library(readr)
library(lubridate)
source("program/R/helpers.R")

# Real platform market share data from EU/US/SA competition authorities
platform_shares <- read_csv("data/derived/platform_market_shares.csv", show_col_types = FALSE)

# Commission rate changes from public filings
commissions <- read_csv("data/derived/platform_commission_rates.csv", show_col_types = FALSE)

# Panel A: Market shares by platform (from regulatory filings)
p1 <- ggplot(platform_shares, aes(x = reorder(platform, share), y = share, fill = jurisdiction)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "darkred", linewidth = 0.8) +
  geom_hline(yintercept = 0.7, linetype = "dotted", color = "darkred", linewidth = 0.8) +
  annotate("text", x = 5.5, y = 0.52, label = "50%", color = "darkred", hjust = 0, size = 3) +
  annotate("text", x = 5.5, y = 0.72, label = "70%", color = "darkred", hjust = 0, size = 3) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c(
    "EU" = "#0072B2", "US" = "#D55E00", "SA" = "#009E73"
  )) +
  coord_flip() +
  labs(
    title = "Platform Market Shares from Regulatory Filings",
    subtitle = "Real data from EU, US, and South African competition authorities",
    x = NULL, y = "Market Share", fill = "Jurisdiction",
    caption = "Source: platform_market_shares.csv. Compiled from EU Commission decisions, US litigation filings, SA OIPMI."
  ) +
  theme_antitrust() +
  theme(legend.position = "bottom")

# Panel B: Commission rate changes over time
p2 <- ggplot(commissions, aes(x = as.Date(date), y = commission_rate, color = platform)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_color_manual(values = c(
    "Apple App Store" = "#0072B2", "Google Play" = "#D55E00",
    "Amazon" = "#E69F00", "Uber Eats" = "#009E73"
  )) +
  labs(
    title = "Platform Commission Rate Changes",
    subtitle = "Real commission rates from public filings and regulatory disclosures",
    x = NULL, y = "Commission Rate", color = NULL,
    caption = "Source: platform_commission_rates.csv. From public filings and regulatory proceedings."
  ) +
  theme_antitrust() +
  theme(legend.position = "bottom")

# Display both panels
if (requireNamespace("patchwork", quietly = TRUE)) {
  library(patchwork)
  p1 / p2 + plot_layout(heights = c(1, 1))
} else {
  print(p1)
  print(p2)
}

# HHI summary
cat("\nPlatform market concentration (HHI by case):\n")
platform_shares |>
  group_by(case_name) |>
  summarise(
    n_firms = n(),
    hhi = sum((share * 100)^2),
    leader = platform[which.max(share)],
    leader_share = max(share),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 0,
               caption = "HHI computed from real platform market share data")
```

These real market share figures from EU, US, and South African regulatory proceedings illustrate the concentration patterns that define digital platform markets. The commission rate panel shows how platform fees have evolved in response to regulatory pressure---Apple's reduction from 30% to 15% for small developers (2020), Google's matching response (2021), and the EU's Digital Markets Act requirements for greater fee transparency.

**Data sources:**
- **StatCounter**: Global browser, OS, search engine stats
- **SimilarWeb**: Website and app traffic data
- **Sensor Tower / App Annie**: Mobile app market intelligence
- **Regulatory filings**: EC, CMA, FTC, DMA compliance reports
- **OIPMI**: South African platform market shares

**Use cases:**
- Demonstrate sustained dominance or competitive entry
- Identify tipping points or competitive interventions
- Support market definition and dominance assessments

## Generative AI and foundation models

The platform competition framework developed above—multi-sided markets, network effects, data as a competitive moat—carries over to AI markets. Foundation models add concerns that merit separate treatment: compute concentration, partnership structures that may confer control without acquisition, and rapidly evolving capabilities that strain static market share analysis.

This section surveys early enforcement thinking on AI markets, adapting concepts from the platform discussion above. As of 2025-2026, enforcement agencies are actively examining AI markets for potential antitrust concerns.

### Market structure concerns

**Vertical integration and compute access:** A small number of cloud providers (AWS, Azure, Google Cloud) control the compute infrastructure essential for training frontier models. Vertical integration between cloud providers and AI model developers (Microsoft/OpenAI, Google/DeepMind, Amazon/Anthropic) creates potential foreclosure concerns similar to traditional media/distribution cases.

**Data moats:** Training data advantages may create barriers to entry that are difficult to replicate. Key questions include:
- Whether data access is essential for competitive AI development
- Whether exclusive data arrangements foreclose rivals
- The role of synthetic data in reducing data barriers

**Partnerships vs. acquisitions:** The Microsoft/OpenAI and Amazon/Anthropic partnership structures may achieve control similar to acquisitions without triggering merger notification. Agencies are scrutinizing whether these arrangements confer de facto control.

### Theory of harm framework

| Concern | Mechanism | Measurement approach |
|:--------|:----------|:--------------------|
| **Compute foreclosure** | Integrated cloud provider degrades rivals' access | API latency/uptime for affiliated vs. unaffiliated models |
| **Training data exclusivity** | Exclusive licensing of high-quality datasets | Document analysis; entry barrier assessment |
| **API pricing squeeze** | Below-cost API pricing to foreclose rivals | Margin analysis; comparison to standalone costs |
| **Distribution bundling** | Tying AI features to dominant platforms | Default/pre-installation analysis (cf. browser cases) |
| **Talent concentration** | Acqui-hires and non-competes limiting rival innovation | Labor market analysis (see @sec-labor-markets) |

### Emerging regulatory frameworks

- **EU AI Act (2024):** Establishes tiered obligations based on risk level; interacts with DMA gatekeeper obligations for AI-integrated platforms.
- **US Executive Order on AI (2023):** Directs agencies to monitor competitive dynamics; potential FTC Section 5 authority for unfair methods.
- **CMA AI Foundation Models review (2023-ongoing):** Examines compute concentration, partnerships, and potential competition concerns.

### Practical implications

When analyzing AI market competition:

1. **Define the relevant market carefully:** Is it foundation models, fine-tuned applications, API access, or end-user AI features? Market boundaries are fluid and contested.
2. **Assess partnership structures:** Evaluate whether minority investments, commercial agreements, and board seats confer control absent formal acquisition.
3. **Consider dynamic competition:** AI capabilities evolve rapidly; static market shares may understate competitive dynamics or overstate durability of positions.
4. **Examine data and compute access:** These inputs may function as essential facilities; apply raising rivals' costs and foreclosure frameworks.

For the latest agency guidance, monitor CMA's AI Foundation Models work and FTC/DOJ statements on AI competition.

## Exercises

1. **Data/code.** Load `data/derived/platform_search_ranking.csv` and estimate a regression of click-through rate on ranking position, with separate slopes for first-party vs. third-party listings. Is the ranking advantage for first-party listings statistically significant? What would this imply for a self-preferencing theory of harm?

2. **Conceptual.** Explain the SSNDQ framework. A social media platform increases ad load from 5 to 7 ads per session. Using the zero-price demand model from this chapter, estimate the predicted effect on daily active users. Under what conditions would this quality degradation constitute an exercise of market power?

3. **Case discussion.** Compare the EU's DMA approach (ex ante regulation of designated gatekeepers) with the US approach (case-by-case antitrust enforcement). What are the advantages and disadvantages of each for addressing self-preferencing by dominant platforms?

4. **Data/code.** Using `data/derived/platform_multihoming_users.csv`, compute multi-homing rates for each platform. If Platform A appears in 80% of multi-homing combinations while Platform B appears in only 30%, what does this asymmetry suggest about competitive constraints and market power?

5. **Conceptual.** A ride-hailing platform argues that high multi-homing rates among drivers prove the market is competitive. Evaluate this argument. Under what conditions could high multi-homing coexist with significant market power on the rider side?

### Data exercise (checkable)

On a search platform, the click-through rate is 35% for the top organic position and 8% for position 4. The platform demotes a rival comparison-shopping service from position 1 to position 4 on a query that runs 1,000,000 times.

a. Compute the rival's lost clicks from the demotion.
b. Why does this metric matter for a self-preferencing theory of harm?

::: {.callout-tip collapse="true" title="Worked answer"}
a. Lost clicks = (0.35 - 0.08) x 1,000,000 = 0.27 x 1,000,000 = **270,000 clicks**.
b. It quantifies the causal traffic diversion from a ranking change -- the empirical core of the *Google Shopping* theory of harm. Tying the click loss to the algorithm change, rather than to organic quality differences, is what separates unlawful self-preferencing from legitimate ranking.
:::

## Looking ahead

Digital markets move fast, so treat the frameworks here as a starting point. @sec-labor-markets applies similar monopsony frameworks to employer-side platform power---gig economy classification, wage-fixing algorithms, and no-poach agreements. @sec-litigation-practice covers how to package digital market evidence into expert reports and trial exhibits. Many of the same event-study and difference-in-differences techniques used here to evaluate platform conduct reappear when assessing remedy effectiveness and merger retrospectives.
