# Empirical Appendix and Templates {.unnumbered}

## Purpose
This appendix is the shared workspace for reusable diagnostics, qualitative protocols, and data inventories referenced throughout the book. Treat it as the "operations manual" for the course: every code chunk or template lives here so you can drop it into chapter-specific notebooks, expert reports, or litigation memos without rewriting boilerplate.

## How to use this appendix
1. **Start here during scoping.** Clone the chronology, data inventory, and interview templates before diving into a matter so legal, economic, and qualitative teams stay synchronized.
2. **Customize but keep provenance.** When you adapt a template, note the source so future readers can trace assumptions.
3. **Feed updates back in.** If a later chapter develops a better diagnostic, link it here to avoid divergence.

## Key resources for quick reference

| Chapter focus | Key resources |
| --- | --- |
| Orientation & institutions | Agency websites: [DOJ Antitrust](https://www.justice.gov/atr), [FTC](https://www.ftc.gov/), [EC Competition](https://ec.europa.eu/competition/), [CMA](https://www.gov.uk/cma) |
| Research design & methods | [Causal Inference: The Mixtape](https://mixtape.scunning.com/), Angrist & Pischke papers |
| Cartels & collusion | [OECD Cartel Guidance](https://www.oecd.org/daf/competition/cartels/) |
| Mergers | US/EU/UK Merger Guidelines (see references) |
| Digital markets | [EU DMA](https://ec.europa.eu/dma), CMA Digital Markets Unit |
| Labor markets | [BLS data](https://www.bls.gov/), [Census LEHD](https://lehd.ces.census.gov/) |

Use this table alongside external references such as [@ashenfelter_hosken_2010] and agency guidelines.

## Case chronology template
When building mixed-method narratives, start with a matrix that ties events to evidence and hypotheses. Duplicate the scaffold below into your project notebook:

| Date | Event / document | Evidence type | Hypothesis link | Follow-ups |
| --- | --- | --- | --- | --- |
| 2018-04-12 | Board memo re: “price discipline” | Document (custodian: CFO) | Supports coordinated effects theory | Interview CFO; pull bid data |
| 2019-02-01 | Customer interview (retailer) | Qualitative | Tests diversion/USP claims | Quantify share of wallet |
| 2021-09-30 | Price spike (SKU 124) | Quantitative (transaction DB) | Flags potential capacity withholding | Overlay maintenance logs |

**Tips:** capture custodian names, bates ranges, or dataset pointers, then push the table into a `gt` or `quarto` table for reports.

## Data inventory worksheet
Pair the chronology with a tabular schema inventory so you know what can feed the quantitative chapters. Customize the tibble below inside your R session.

```{r}
#| label: data-inventory-template
#| message: false
#| warning: false
library(tibble)
library(dplyr)
source("program/R/helpers.R")

data_inventory <- tribble(
  ~system, ~owner, ~grain, ~fields, ~coverage, ~notes,
  "CRM / Salesforce", "US Sales Ops", "Customer-month", "price, quantity, discounts, churn", "2016-01 to 2024-06", "Great for diversion estimates; use OECD cartel screening guidance",
  "Claims data", "SA Health Plan", "Claim-line", "procedure, tariff, provider, patient segment", "2014-01 to 2023-12", "Use for South African retrospectives",
  "Bid tracker", "EU Procurement", "Bid-event", "participants, bids, award, lot size", "2018-01 to present", "Supports cartel screens"
) |>
  arrange(system)

data_inventory
```

Document locating instructions (sharepoint links, API keys) and legal restrictions. Store sanitized metadata in `data/README.md` so others can reproduce.

## Survey and interview kit
Follow established qualitative research standards to keep qualitative work rigorous:
- **Screeners:** Define decision-maker roles, procurement thresholds, and geographic balance; record quotas in a simple CSV.
- **Question bank:** Map each question to a legal element (market definition, competitive effects, efficiencies). Rotate order to reduce priming.
- **Documentation:** Save audio or transcripts, then code responses using the lexicon below. Always log interviewer, location, and consent status.

## Document coding lexicon
Standardize qualitative tagging so it can be merged with quantitative variables:
- **Conduct codes:** `capacity_discipline`, `most_favored`, `exclusive`, `discount_pass_through`.
- **Market context:** `entry_barrier_data`, `supply_chain_shock`, `public_interest_sa`.
- **Procedural:** `remedy_negotiation`, `leniency_proffer`, `public_hearing`.

Maintain the lexicon as a YAML or CSV in `data/derived`.

## Flagship data assets (by jurisdiction)

### United States
- **HSR and merger retrospectives:** Derive pre/post comparisons referencing [FTC](https://www.ftc.gov/) and [DOJ](https://www.justice.gov/atr) reports.
- **Healthcare claims (CMS, FAIR Health):** Price and quantity data for hospital mergers and labor monopsony cases.
- **Public procurement / ABM data:** Use [USAspending.gov](https://www.usaspending.gov/), OCC data, or [BLS](https://www.bls.gov/) Occupational Employment and Wage Statistics for wage-fixing analyses.

### European Union & United Kingdom
- **Case team data rooms:** DG COMP uploads (pricing, cost, user logs).
- **Eurostat + Ofcom datasets:** Demand-side metrics for telecom, digital platforms, or broadcasting.
- **CMA survey repositories:** Useful for designing conjoint or diversion studies.

### South Africa & SADC focus
- **Competition Commission data requests:** Transactional extracts from retailers, healthcare groups, or digital platforms; cross-check with public data from Stats SA.
- **National Treasury tender portal:** Procurement and supplier concentration data for cartel screens.
- **Sector inquiries:** Grocery retail, digital platforms, and transport datasets—log file formats and confidentiality riders here so later chapters can reference them.

### Cross-jurisdiction validation
- Build a blended index (e.g., inflation-adjusted revenue) to harmonize multi-region cases.
- Track currency conversions and PPP adjustments in the inventory template.

## Reusable diagnostics

### Diff-in-diff and event studies
- Run parallel-trend visuals, dynamic leads/lags, and placebo periods (`did::att_gt`, `fixest::sunab`). Reference @cunningham_2021 for intuition and cite @ashenfelter_hosken_2010 for merger retrospectives. For staggered treatment timing, see @callaway_santanna_2021.
- For South African matters with staggered regulatory actions, consider matrix-completion methods from recent econometrics literature [@athey_imbens_2017].

### Matching and weighting
- Propensity score, entropy balancing, and synthetic controls should include balance plots (standardized differences, variance ratios) [@abadie_diamond_hainmueller_2010].
- Store reusable plotting code in `R/helpers.R` (`plot_balance()`). For general guidance on matching methods, see @angrist_pischke_2009.

### Panel FE and inference choices
- Two-way FE with staggered treatment: document estimator choice and weights; cite the relevant methodological literature.
- For few clusters (ports, mines), use wild bootstrap or randomization inference.

### Qualitative + quantitative integration
- Align interview findings with econometric outputs using issue trees. For example, code “capacity discipline” documents and test against utilization regressions.
- Use the chronology template to surface inconsistencies quickly.

## Code scaffolds

### Data dictionary generator
```{r}
#| label: data-dictionary
#| message: false
library(readr)
library(purrr)

generate_dictionary <- function(path) {
  sample <- readr::read_csv(path, n_max = 1000, show_col_types = FALSE)
  purrr::map_dfr(names(sample), \(nm) {
    tibble::tibble(
      variable = nm,
      type = class(sample[[nm]])[1],
      sample_values = paste(head(unique(sample[[nm]]), 3), collapse = "; "),
      notes = ""
    )
  })
}

# Example usage (uncomment when data available):
# dict <- generate_dictionary("data/raw/claims_sample.csv")
# dict
```

### Survey weighting stub
```{r}
#| label: survey-weighting
#| eval: false
library(survey)

design <- svydesign(
  ids = ~1,
  strata = ~region,
  weights = ~weight,
  data = survey_responses
)

svymean(~switch_intent, design)
```

### Chronology visualization
```{r}
#| label: chronology-viz
#| message: false
#| warning: false
library(ggplot2)
library(lubridate)

chronology <- tibble::tribble(
  ~date, ~category, ~label,
  ymd("2018-04-12"), "Document", "Board memo",
  ymd("2019-02-01"), "Interview", "Retailer A",
  ymd("2021-09-30"), "Data", "SKU 124 spike"
)

ggplot(chronology, aes(x = date, y = category, color = category)) +
  geom_point(size = 3) +
  geom_text(aes(label = label), hjust = -0.1, size = 3) +
  labs(title = "Chronology overview", x = NULL, y = NULL) +
  theme_antitrust() +
  theme(legend.position = "none")
```

## Diagnostic Gallery

This section provides reusable diagnostic visualizations that apply across chapters. Use these templates to validate identification assumptions, assess robustness, and communicate uncertainty in expert reports.

### Pre-trends and parallel trends checks
Essential for difference-in-differences designs (mergers, labor, remedies).

```{r}
#| label: diagnostic-pretrends
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 7
library(dplyr)
library(ggplot2)
library(fixest)
library(patchwork)

# Simulated panel data with treatment in period 10
# Replace with your actual panel data
set.seed(123)
periods <- 20
n_units <- 50

panel <- expand.grid(
  unit = 1:n_units,
  period = 1:periods
) |>
  mutate(
    treated = ifelse(unit <= n_units/2, 1, 0),
    post = ifelse(period >= 10, 1, 0),
    # Pre-trends: parallel before period 10
    outcome = 100 + 2 * period + 5 * treated +
              10 * treated * post +
              rnorm(n(), 0, 3)
  )

# Plot 1: Raw trends by group
p1 <- panel |>
  group_by(period, treated) |>
  summarise(mean_outcome = mean(outcome), .groups = "drop") |>
  mutate(group = ifelse(treated == 1, "Treatment", "Control")) |>
  ggplot(aes(x = period, y = mean_outcome, color = group, linetype = group)) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "gray40") +
  geom_line(linewidth = 1.2) +
  annotate("text", x = 10, y = max(panel$outcome) * 0.95,
          label = "Treatment\nperiod", hjust = -0.1, size = 3.5) +
  scale_color_manual(values = c("Treatment" = "#D55E00",
                                 "Control" = "#0072B2")) +
  labs(
    title = "Pre-Trends Check: Raw Outcome Trends",
    subtitle = "Parallel trends before period 10 support DiD identification",
    x = "Period",
    y = "Outcome",
    color = NULL,
    linetype = NULL
  ) +
  theme_antitrust() +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Event study with leads and lags
event_study <- feols(
  outcome ~ i(period, treated, ref = 9) | unit + period,
  data = panel
)

# Extract coefficients
coef_data <- tibble(
  period = as.numeric(gsub("period::", "", names(coef(event_study)))),
  estimate = coef(event_study),
  se = se(event_study)
) |>
  mutate(
    ci_lower = estimate - 1.96 * se,
    ci_upper = estimate + 1.96 * se,
    rel_period = period - 10
  )

# Plot 2: Event study coefficients
p2 <- ggplot(coef_data, aes(x = rel_period, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
             alpha = 0.2, fill = "#0072B2") +
  geom_line(color = "#0072B2", linewidth = 1) +
  geom_point(color = "#0072B2", size = 2.5) +
  annotate("rect", xmin = -Inf, xmax = -1, ymin = -Inf, ymax = Inf,
          fill = "green", alpha = 0.05) +
  annotate("text", x = -5, y = max(coef_data$ci_upper) * 0.9,
          label = "Pre-period\n(should be ≈0)", size = 3, hjust = 0.5) +
  labs(
    title = "Event Study Coefficients (Leads and Lags)",
    subtitle = "Pre-period estimates near zero validate parallel trends assumption",
    x = "Periods relative to treatment (0 = treatment period)",
    y = "Treatment Effect Estimate",
    caption = "Reference period: -1. Shaded region shows 95% confidence interval."
  ) +
  theme_antitrust() +
  theme(plot.title.position = "plot")

# Combine plots
p1 / p2

# Statistical test for pre-trends
pre_period_coefs <- coef_data |>
  filter(rel_period < 0, rel_period != -1)

cat("\nPre-trends test:\n")
cat(paste0("Number of pre-period coefficients: ", nrow(pre_period_coefs), "\n"))
cat(paste0("Mean pre-period coefficient: ",
          round(mean(pre_period_coefs$estimate), 3), "\n"))
cat(paste0("Joint F-test p-value: [run wald_test on event_study model]\n"))
```

**Interpretation:**
- **Pre-trends near zero**: Validates parallel trends assumption
- **Post-treatment divergence**: Confirms treatment effect
- **Confidence intervals**: Should include zero in pre-period
- **Use this**: Mergers (Ch. 06), remedies (Ch. 08), labor (Ch. 10)

### Balance plots for matching/weighting
Show covariate balance before and after matching or weighting.

```{r}
#| label: diagnostic-balance
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8
library(dplyr)
library(ggplot2)
library(tidyr)

# Simulated covariate data (treatment and control groups)
set.seed(234)
n <- 200

covariates <- tibble(
  id = 1:n,
  treated = sample(c(0, 1), n, replace = TRUE),
  firm_size = rnorm(n, mean = 100 + 20 * treated, sd = 30),
  market_share = rnorm(n, mean = 0.15 + 0.05 * treated, sd = 0.08),
  years_operating = rnorm(n, mean = 15 + 5 * treated, sd = 7),
  revenue_growth = rnorm(n, mean = 0.10 + 0.03 * treated, sd = 0.12)
)

# Calculate standardized mean differences
calc_smd <- function(var, group) {
  mean_1 <- mean(var[group == 1])
  mean_0 <- mean(var[group == 0])
  sd_pool <- sqrt((var(var[group == 1]) + var(var[group == 0])) / 2)
  (mean_1 - mean_0) / sd_pool
}

balance_before <- covariates |>
  summarise(across(c(firm_size, market_share, years_operating, revenue_growth),
                  ~calc_smd(.x, treated),
                  .names = "{.col}")) |>
  pivot_longer(everything(), names_to = "variable", values_to = "smd") |>
  mutate(timing = "Before matching")

# Simulate "after matching" with reduced imbalance
# In practice, replace this with actual matched/weighted sample
balance_after <- balance_before |>
  mutate(
    smd = smd * 0.2,  # Matching reduces imbalance by 80%
    timing = "After matching"
  )

variable_levels <- c(
  "Firm size (employees)",
  "Market share",
  "Years operating",
  "Revenue growth"
)

balance_combined <- bind_rows(balance_before, balance_after) |>
  mutate(
    variable = recode(
      variable,
      "firm_size" = "Firm size (employees)",
      "market_share" = "Market share",
      "years_operating" = "Years operating",
      "revenue_growth" = "Revenue growth"
    ),
    variable = factor(variable, levels = variable_levels),
    timing = factor(timing, levels = c("Before matching", "After matching"))
  )

annotation_label <- tibble(
  abs_smd = 0.1,
  variable = variable_levels[1],
  label = "0.1 threshold\n(acceptable)"
)

# Love plot
ggplot(balance_combined, aes(x = abs(smd), y = variable,
                              color = timing, shape = timing)) +
  geom_vline(xintercept = 0.1, linetype = "dashed",
            color = "darkgreen", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray40") +
  geom_point(size = 3) +
  geom_text(
    data = annotation_label,
    aes(x = abs_smd, y = variable, label = label),
    inherit.aes = FALSE,
    hjust = -0.1,
    size = 3
  ) +
  scale_color_manual(values = c("Before matching" = "#D55E00",
                                 "After matching" = "#0072B2")) +
  scale_shape_manual(values = c("Before matching" = 16,
                                "After matching" = 17)) +
  scale_x_continuous(limits = c(0, max(abs(balance_combined$smd)) * 1.1)) +
  labs(
    title = "Covariate Balance: Love Plot",
    subtitle = "Standardized mean differences before and after matching",
    x = "Absolute Standardized Mean Difference",
    y = NULL,
    color = "Timing",
    shape = "Timing",
    caption = "SMD < 0.1 indicates acceptable balance (Austin, 2009)."
  ) +
  theme_antitrust() +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Summary table
cat("\nBalance summary:\n")
balance_combined |>
  select(variable, timing, smd) |>
  pivot_wider(names_from = timing, values_from = smd) |>
  mutate(improvement = abs(`Before matching`) - abs(`After matching`)) |>
  print()
```

**Interpretation:**
- **SMD < 0.1**: Acceptable balance (Austin, 2009)
- **Matching improves balance**: Lines should move toward zero
- **Use this**: Merger retrospectives, labor market studies, remedy evaluations

### Regression specification curve
Show robustness across multiple reasonable specifications.

```{r}
#| label: diagnostic-spec-curve
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 8
library(dplyr)
library(ggplot2)
library(fixest)

# Simulated data for multiple specifications
set.seed(345)
panel_spec <- expand.grid(
  firm = 1:100,
  period = 1:20
) |>
  mutate(
    treatment = ifelse(firm <= 50 & period >= 10, 1, 0),
    control1 = rnorm(n()),
    control2 = rnorm(n()),
    control3 = rnorm(n()),
    outcome = 50 + 5 * treatment + 2 * control1 + 1 * control2 +
              0.5 * control3 + rnorm(n(), 0, 3)
  )

# Run multiple specifications
specs <- list(
  "No controls" = feols(outcome ~ treatment | firm + period, panel_spec),
  "+ Control 1" = feols(outcome ~ treatment + control1 | firm + period, panel_spec),
  "+ Controls 1-2" = feols(outcome ~ treatment + control1 + control2 | firm + period, panel_spec),
  "+ All controls" = feols(outcome ~ treatment + control1 + control2 + control3 | firm + period, panel_spec),
  "No FE" = lm(outcome ~ treatment + control1 + control2 + control3, panel_spec),
  "Firm FE only" = feols(outcome ~ treatment + control1 + control2 + control3 | firm, panel_spec),
  "Period FE only" = feols(outcome ~ treatment + control1 + control2 + control3 | period, panel_spec)
)

# Extract coefficients
spec_results <- tibble(
  specification = names(specs),
  estimate = sapply(specs, function(m) {
    coefs <- coef(m)
    coefs[grepl("treatment", names(coefs))][1]
  }),
  se = sapply(specs, function(m) {
    ses <- se(m)
    ses[grepl("treatment", names(ses))][1]
  })
) |>
  mutate(
    ci_lower = estimate - 1.96 * se,
    ci_upper = estimate + 1.96 * se,
    spec_num = row_number(),
    specification = factor(specification, levels = specification)
  )

# Specification curve plot
ggplot(spec_results, aes(x = spec_num, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
             alpha = 0.2, fill = "#0072B2") +
  geom_line(color = "#0072B2", linewidth = 1.2) +
  geom_point(color = "#0072B2", size = 3) +
  scale_x_continuous(
    breaks = spec_results$spec_num,
    labels = spec_results$specification
  ) +
  labs(
    title = "Specification Curve: Treatment Effect Robustness",
    subtitle = "Estimate remains positive and significant across specifications",
    x = NULL,
    y = "Treatment Effect Estimate",
    caption = "Each point represents a different model specification. Shaded area = 95% CI."
  ) +
  theme_antitrust() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title.position = "plot"
  )

# Summary statistics
cat("\nSpecification curve summary:\n")
cat(paste0("Range of estimates: ",
          round(min(spec_results$estimate), 2), " to ",
          round(max(spec_results$estimate), 2), "\n"))
cat(paste0("All specifications significant? ",
          ifelse(all(spec_results$ci_lower > 0), "Yes", "No"), "\n"))
cat(paste0("Median estimate: ", round(median(spec_results$estimate), 2), "\n"))
```

**Interpretation:**
- **Stable across specs**: Effect robust to model choices
- **All CIs exclude zero**: Consistent significance
- **Use this**: Expert reports, Daubert challenges, sensitivity analysis

### Residual diagnostics
Check model assumptions and identify influential observations.

```{r}
#| label: diagnostic-residuals
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 9
library(dplyr)
library(ggplot2)
library(patchwork)

# Fit model (use your actual model)
model_data <- panel_spec |>
  filter(period >= 10)

model <- feols(outcome ~ treatment + control1 + control2 | firm + period,
              data = model_data)

# Get residuals and fitted values
diagnostics <- model_data |>
  mutate(
    fitted = fitted(model),
    residual = resid(model),
    standardized_residual = residual / sd(residual)
  )

# Plot 1: Residuals vs. fitted
p1 <- ggplot(diagnostics, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(alpha = 0.5, color = "#0072B2") +
  geom_smooth(method = "loess", se = FALSE, color = "#D55E00") +
  labs(
    title = "Residuals vs. Fitted Values",
    subtitle = "Should show no pattern (validates linearity assumption)",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_antitrust()

# Plot 2: Q-Q plot for normality
p2 <- ggplot(diagnostics, aes(sample = standardized_residual)) +
  stat_qq(color = "#0072B2", alpha = 0.6) +
  stat_qq_line(color = "#D55E00", linewidth = 1) +
  labs(
    title = "Q-Q Plot: Residual Normality",
    subtitle = "Points should follow line for normal distribution",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_antitrust()

# Plot 3: Scale-location plot
p3 <- ggplot(diagnostics, aes(x = fitted, y = sqrt(abs(standardized_residual)))) +
  geom_point(alpha = 0.5, color = "#0072B2") +
  geom_smooth(method = "loess", se = FALSE, color = "#D55E00") +
  labs(
    title = "Scale-Location Plot",
    subtitle = "Should show horizontal line (validates homoscedasticity)",
    x = "Fitted Values",
    y = "√|Standardized Residuals|"
  ) +
  theme_antitrust()

# Plot 4: Histogram of residuals
p4 <- ggplot(diagnostics, aes(x = residual)) +
  geom_histogram(bins = 30, fill = "#0072B2", alpha = 0.8, color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#D55E00") +
  labs(
    title = "Distribution of Residuals",
    subtitle = "Should be approximately normal, centered at zero",
    x = "Residuals",
    y = "Count"
  ) +
  theme_antitrust()

# Combine plots
(p1 | p2) / (p3 | p4) + plot_annotation(
  title = "Residual Diagnostic Plots",
  caption = "Check for violations of regression assumptions"
)
```

**What to look for:**
- **Residuals vs. Fitted**: No pattern (confirms linearity)
- **Q-Q Plot**: Points follow line (confirms normality)
- **Scale-Location**: Horizontal line (confirms homoscedasticity)
- **Histogram**: Bell-shaped, centered at zero

### Power analysis visualization
Show statistical power across different sample sizes and effect sizes.

```{r}
#| label: diagnostic-power
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 7
library(dplyr)
library(ggplot2)
library(pwr)

# Calculate power for different scenarios
sample_sizes <- seq(50, 500, by = 25)
effect_sizes <- c(0.2, 0.5, 0.8)  # Small, medium, large (Cohen's d)

power_grid <- expand.grid(
  n = sample_sizes,
  effect_size = effect_sizes
) |>
  mutate(
    power = pwr.t.test(n = n, d = effect_size, sig.level = 0.05,
                      type = "two.sample")$power,
    effect_label = case_when(
      effect_size == 0.2 ~ "Small effect (d = 0.2)",
      effect_size == 0.5 ~ "Medium effect (d = 0.5)",
      effect_size == 0.8 ~ "Large effect (d = 0.8)"
    )
  )

# Power curve
ggplot(power_grid, aes(x = n, y = power, color = effect_label)) +
  geom_hline(yintercept = 0.80, linetype = "dashed",
            color = "darkgreen", linewidth = 0.8) +
  annotate("text", x = max(sample_sizes) * 0.7, y = 0.80,
          label = "80% power threshold", hjust = 0, vjust = -0.5,
          size = 3, color = "darkgreen") +
  geom_line(linewidth = 1.2) +
  geom_point(data = filter(power_grid, power >= 0.79, power <= 0.81),
            size = 3) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_color_manual(values = c(
    "Small effect (d = 0.2)" = "#D55E00",
    "Medium effect (d = 0.5)" = "#0072B2",
    "Large effect (d = 0.8)" = "#009E73"
  )) +
  labs(
    title = "Statistical Power Analysis",
    subtitle = "Sample size requirements for 80% power at α = 0.05",
    x = "Sample Size (per group)",
    y = "Statistical Power",
    color = "Effect Size",
    caption = "Points indicate minimum sample size for 80% power."
  ) +
  theme_antitrust() +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Summary table
power_summary <- power_grid |>
  filter(power >= 0.79, power <= 0.81) |>
  group_by(effect_label) |>
  slice_min(n, n = 1) |>
  ungroup() |>
  select(effect_label, n_required = n, power)

cat("\nMinimum sample size for 80% power:\n")
print(power_summary, n = Inf)
```

**Use this for:**
- Study design and sample size planning
- Explaining null results (underpowered?)
- Justifying sample selection in expert reports

## Additional references
- **FTC resources:** See [ftc.gov](https://www.ftc.gov/) for expert guidance and econometric standards.
- **Causal Inference: The Mixtape** ([mixtape.scunning.com](https://mixtape.scunning.com/)) for causal estimators and diagnostics.
- **OECD cartel guidance:** See [oecd.org/competition](https://www.oecd.org/daf/competition/) for screening methodologies.

Return to this appendix whenever a chapter references "see appendix template" so the workflow stays consistent across teams and jurisdictions.

