# Empirical Appendix and Templates {#sec-empirical-appendix .unnumbered}

## Purpose
This appendix is the shared workspace for reusable diagnostics, qualitative protocols, and data inventories referenced throughout the book. Treat it as the "operations manual" for the course: every code chunk or template lives here so you can drop it into chapter-specific notebooks, expert reports, or litigation memos without rewriting boilerplate.

## How to use this appendix
1. **Start here during scoping.** Clone the chronology, data inventory, and interview templates at the outset so legal, economic, and qualitative teams stay synchronized.
2. **Customize but keep provenance.** When you adapt a template, note the source so future readers can trace assumptions.
3. **Feed updates back in.** If a later chapter develops a better diagnostic, link it here to avoid divergence.

## Key resources for quick reference

| Chapter focus | Key resources |
| --- | --- |
| Orientation & institutions | Agency websites: [DOJ Antitrust](https://www.justice.gov/atr), [FTC](https://www.ftc.gov/), [EC Competition](https://ec.europa.eu/competition/), [CMA](https://www.gov.uk/cma) |
| Research design & methods | [Causal Inference: The Mixtape](https://mixtape.scunning.com/), Angrist & Pischke [@angrist_pischke_2009; @angrist_pischke_2015] |
| Cartels & collusion | [OECD Cartel Guidance](https://www.oecd.org/daf/competition/cartels/) |
| Mergers | US/EU/UK Merger Guidelines [@doj_ftc_hmg_2023; @ec_hmg_2004; @cma_merger_assessment_2021] |
| Digital markets | [EU DMA](https://ec.europa.eu/dma), CMA Digital Markets Unit |
| Labor markets | [BLS data](https://www.bls.gov/), [Census LEHD](https://lehd.ces.census.gov/) |

Use this table alongside external references such as [@ashenfelter_hosken_2010] and agency guidelines.

## Case chronology template
When building mixed-method narratives, start with a matrix that ties events to evidence and hypotheses. Duplicate the scaffold below into your project notebook:

| Date | Event / document | Evidence type | Hypothesis link | Follow-ups |
| --- | --- | --- | --- | --- |
| 2018-04-12 | Board memo re: “price discipline” | Document (custodian: CFO) | Supports coordinated effects theory | Interview CFO; pull bid data |
| 2019-02-01 | Customer interview (retailer) | Qualitative | Tests diversion/USP claims | Quantify share of wallet |
| 2021-09-30 | Price spike (SKU 124) | Quantitative (transaction DB) | Flags potential capacity withholding | Overlay maintenance logs |

**Tips:** capture custodian names, bates ranges, or dataset pointers, then push the table into a `gt` or `quarto` table for reports.

## Data inventory worksheet
Pair the chronology with a tabular schema inventory so you know what can feed the quantitative chapters. Customize the tibble below inside your R session.

```{r}
#| label: data-inventory-template
#| message: false
#| warning: false
library(tibble)
library(dplyr)
source("program/R/helpers.R")

data_inventory <- tribble(
  ~system, ~owner, ~grain, ~fields, ~coverage, ~notes,
  "CRM / Salesforce", "US Sales Ops", "Customer-month", "price, quantity, discounts, churn", "2016-01 to 2024-06", "Great for diversion estimates; use OECD cartel screening guidance",
  "Claims data", "SA Health Plan", "Claim-line", "procedure, tariff, provider, patient segment", "2014-01 to 2023-12", "Use for South African retrospectives",
  "Bid tracker", "EU Procurement", "Bid-event", "participants, bids, award, lot size", "2018-01 to present", "Supports cartel screens"
) |>
  arrange(system)

data_inventory
```

Document locating instructions (sharepoint links, API keys) and legal restrictions. Store sanitized metadata in `data/README.md` so others can reproduce.

## Survey and interview kit
Follow established qualitative research standards to keep qualitative work rigorous:
- **Screeners:** Define decision-maker roles, procurement thresholds, and geographic balance; record quotas in a simple CSV.
- **Question bank:** Map each question to a legal element (market definition, competitive effects, efficiencies). Rotate order to reduce priming.
- **Documentation:** Save audio or transcripts, then code responses using the lexicon below. Always log interviewer, location, and consent status.

## Document coding lexicon
Standardize qualitative tagging so it can be merged with quantitative variables:
- **Conduct codes:** `capacity_discipline`, `most_favored`, `exclusive`, `discount_pass_through`.
- **Market context:** `entry_barrier_data`, `supply_chain_shock`, `public_interest_sa`.
- **Procedural:** `remedy_negotiation`, `leniency_proffer`, `public_hearing`.

Maintain the lexicon as a YAML or CSV in `data/derived`.

## Flagship data assets (by jurisdiction)

### United States
- **HSR and merger retrospectives:** Derive pre/post comparisons referencing [FTC](https://www.ftc.gov/) and [DOJ](https://www.justice.gov/atr) reports.
- **Healthcare claims (CMS, FAIR Health):** Price and quantity data for hospital mergers and labor monopsony cases.
- **Public procurement / ABM data:** Use [USAspending.gov](https://www.usaspending.gov/), OCC data, or [BLS](https://www.bls.gov/) Occupational Employment and Wage Statistics for wage-fixing analyses.

### European Union & United Kingdom
- **Case team data rooms:** DG COMP uploads (pricing, cost, user logs).
- **Eurostat + Ofcom datasets:** Demand-side metrics for telecom, digital platforms, or broadcasting.
- **CMA survey repositories:** Useful for designing conjoint or diversion studies.

### South Africa & SADC focus
- **Competition Commission data requests:** Transactional extracts from retailers, healthcare groups, or digital platforms; cross-check with public data from Stats SA.
- **National Treasury tender portal:** Procurement and supplier concentration data for cartel screens.
- **Sector inquiries:** Grocery retail, digital platforms, and transport datasets—log file formats and confidentiality riders here so later chapters can reference them.

### Cross-jurisdiction validation
- Build a blended index (e.g., inflation-adjusted revenue) to harmonize multi-region cases.
- Track currency conversions and PPP adjustments in the inventory template.

## Reusable diagnostics

### Diff-in-diff and event studies
- Run parallel-trend visuals, dynamic leads/lags, and placebo periods (`did::att_gt`, `fixest::sunab`). Reference @cunningham_2021 for intuition and cite @ashenfelter_hosken_2010 for merger retrospectives. For staggered treatment timing, see @callaway_santanna_2021.
- For South African matters with staggered regulatory actions, consider matrix-completion methods from recent econometrics literature [@athey_bayati_imbens_2021].

### Matching and weighting

Matching and weighting methods come into play whenever you need to construct a credible comparison group from observational data. Merger retrospectives, labor market studies, remedy evaluations, and damages analyses all rely on them. The core idea is to reweight or subsample the control group so that its observable characteristics match those of the treated group, reducing selection bias and improving the credibility of causal estimates.

**Propensity score matching (PSM)** estimates the probability of treatment given observed covariates, then matches treated and control units with similar propensity scores. The key diagnostics are balance tests: after matching, the standardized mean difference (SMD) for each covariate should be below 0.1 [@austin_2009]. Love plots (showing SMD before and after matching) are the standard visualization. The limitation of PSM is that it only balances observed covariates; unobserved confounders remain a threat to identification.

**Entropy balancing** [@hainmueller_2012] is an alternative that directly reweights the control group to match the treated group's moments (means, variances, and optionally higher moments) on selected covariates. Unlike PSM, entropy balancing achieves exact balance on the specified moments by construction, and the weights have a clear information-theoretic interpretation (they minimize the Kullback-Leibler divergence from uniform weights). In antitrust applications, entropy balancing is particularly useful for merger retrospectives where the pre-merger characteristics of treated and control markets can be precisely matched.

**Coarsened exact matching (CEM)** partitions the covariate space into strata and retains only treated-control pairs that fall into the same stratum. This guarantees balance on the coarsened covariates but can result in substantial data loss if the covariate space is high-dimensional. CEM is most useful when you have a small number of key matching variables (e.g., market size, pre-treatment price level, geographic region) and want to ensure exact comparability.

**Synthetic control** [@abadie_diamond_hainmueller_2010] constructs a weighted combination of untreated units that matches the treated unit's pre-treatment trajectory. It is the method of choice for "N=1" cases: a single national merger, a one-off regulatory intervention, or a market inquiry affecting a specific sector. The key diagnostics are pre-treatment fit (the synthetic control should closely track the treated unit before the intervention) and placebo tests (applying the same method to each donor unit should produce smaller effects than the treated unit).

For all matching and weighting methods, balance plots showing standardized differences before and after matching should be included in expert reports. The Love-plot code in the diagnostic gallery below (chunk `diagnostic-balance`) is the reusable template; adapt it rather than rewriting. For general guidance on matching methods, see @angrist_pischke_2009 and @austin_2009.

### Panel FE and inference choices
- Two-way FE with staggered treatment: document estimator choice and weights; cite the relevant methodological literature.
- For few clusters (ports, mines), use wild bootstrap or randomization inference.

### Qualitative + quantitative integration
- Align interview findings with econometric outputs using issue trees. For example, code “capacity discipline” documents and test against utilization regressions.

## Connecting templates to chapter applications

This section maps each diagnostic template to the specific chapters where it is applied, so you can quickly find the right tool for your matter.

**Cartel matters (Chapter 5).** Start with the data inventory worksheet to catalog bid data, invoice records, and leniency statements. Use the variance and correlation screens to flag suspicious pricing patterns. The transition matrix and network graph templates identify bid rotation. For damages estimation, the before/after method with structural break tests provides the overcharge estimate. The document coding lexicon helps tag internal communications for price discipline, market allocation, or capacity restraint language.

**Merger analysis (Chapter 6).** The case chronology template tracks the deal timeline from initial approach through agency review to consummation. Use the matching and weighting methods (PSM, entropy balancing, synthetic control) to construct control groups for merger retrospectives. The diff-in-diff template with staggered treatment estimators evaluates post-merger price effects. The data inventory should capture customer-level transaction data, capacity records, and internal strategy documents.

**Monopolization (Chapter 7).** The margin squeeze test and pass-through diagnostics are the core quantitative tools. The document coding lexicon should include tags for exclusive dealing, tying, self-preferencing, and retaliation. The survey and interview kit gathers customer testimony about foreclosure effects and switching costs.

**Digital markets (Chapter 9).** The data inventory should capture platform telemetry: user activity logs, ranking data, API access records, and multi-homing patterns. The self-preferencing detection regression and algorithmic collusion Q-learning simulation provide distinctive analytical tools not available in traditional antitrust toolkits.

**Labor markets (Chapter 10).** Use the data inventory to catalog HR databases, job posting data, and commuting zone definitions. The matching methods construct comparison groups for policy evaluations (noncompete bans, no-poach settlements). The diff-in-diff template evaluates wage effects of enforcement actions.

**Litigation practice (Chapter 12).** The case chronology template is the backbone of expert report preparation. The data inventory ensures all data sources are documented with custodian names, bates ranges, and confidentiality designations. The survey and interview kit supports class certification and damages estimation.

**Regulation and remedies (Chapter 8).** The diff-in-diff template evaluates remedy effectiveness using pre/post KPI data. The benchmarking scatter plot compares regulated entities on cost and quality dimensions. The remedy monitoring KPI table provides a framework for compliance dashboards.
- Use the chronology template to surface inconsistencies quickly.

## Code scaffolds

### Data dictionary generator
```{r}
#| label: data-dictionary
#| message: false
library(readr)
library(purrr)

generate_dictionary <- function(path) {
  sample <- readr::read_csv(path, n_max = 1000, show_col_types = FALSE)
  purrr::map_dfr(names(sample), \(nm) {
    tibble::tibble(
      variable = nm,
      type = class(sample[[nm]])[1],
      sample_values = paste(head(unique(sample[[nm]]), 3), collapse = "; "),
      notes = ""
    )
  })
}

# Example usage (uncomment when data available):
# dict <- generate_dictionary("data/raw/claims_sample.csv")
# dict
```

### Survey weighting stub
```{r}
#| label: survey-weighting
#| eval: false
library(survey)

design <- svydesign(
  ids = ~1,
  strata = ~region,
  weights = ~weight,
  data = survey_responses
)

svymean(~switch_intent, design)
```

### Chronology visualization
```{r}
#| label: chronology-viz
#| message: false
#| warning: false
library(ggplot2)
library(lubridate)

chronology <- tibble::tribble(
  ~date, ~category, ~label,
  ymd("2018-04-12"), "Document", "Board memo",
  ymd("2019-02-01"), "Interview", "Retailer A",
  ymd("2021-09-30"), "Data", "SKU 124 spike"
)

ggplot(chronology, aes(x = date, y = category, color = category)) +
  geom_point(size = 3) +
  geom_text(aes(label = label), hjust = -0.1, size = 3) +
  labs(title = "Chronology overview", x = NULL, y = NULL) +
  theme_antitrust() +
  theme(legend.position = "none")
```

## Diagnostic Gallery

This section provides reusable diagnostic visualizations that apply across chapters. Use these templates to validate identification assumptions, assess robustness, and communicate uncertainty in expert reports.

### Pre-trends and parallel trends checks
Run these before any difference-in-differences design (mergers, labor, remedies).

```{r}
#| label: diagnostic-pretrends
#| fig-cap: "Pre-trends diagnostic. Synthetic data for illustration only."
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 7
library(dplyr)
library(ggplot2)
library(fixest)
library(patchwork)

# Simulated panel data with treatment in period 10
# Replace with your actual panel data
set.seed(123)
periods <- 20
n_units <- 50

panel <- expand.grid(
  unit = 1:n_units,
  period = 1:periods
) |>
  mutate(
    treated = ifelse(unit <= n_units/2, 1, 0),
    post = ifelse(period >= 10, 1, 0),
    # Pre-trends: parallel before period 10
    outcome = 100 + 2 * period + 5 * treated +
              10 * treated * post +
              rnorm(n(), 0, 3)
  )

# Plot 1: Raw trends by group
p1 <- panel |>
  group_by(period, treated) |>
  summarise(mean_outcome = mean(outcome), .groups = "drop") |>
  mutate(group = ifelse(treated == 1, "Treatment", "Control")) |>
  ggplot(aes(x = period, y = mean_outcome, color = group, linetype = group)) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "gray40") +
  geom_line(linewidth = 1.2) +
  annotate("text", x = 10, y = max(panel$outcome) * 0.95,
          label = "Treatment\nperiod", hjust = -0.1, size = 3.5) +
  scale_color_manual(values = c("Treatment" = "#D55E00",
                                 "Control" = "#0072B2")) +
  labs(
    title = "Pre-Trends Check: Raw Outcome Trends",
    subtitle = "Parallel trends before period 10 support DiD identification",
    x = "Period",
    y = "Outcome",
    color = NULL,
    linetype = NULL
  ) +
  theme_antitrust() +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Event study with leads and lags
event_study <- feols(
  outcome ~ i(period, treated, ref = 9) | unit + period,
  data = panel
)

# Extract coefficients. Names look like "period::1:treated", so pull the
# integer between "period::" and the first ":" rather than gsub-ing the prefix.
keep <- grepl("^period::", names(coef(event_study)))
coef_data <- tibble(
  period = as.numeric(sub("^period::(-?[0-9]+).*$", "\\1", names(coef(event_study))[keep])),
  estimate = coef(event_study)[keep],
  se = se(event_study)[keep]
) |>
  mutate(
    ci_lower = estimate - 1.96 * se,
    ci_upper = estimate + 1.96 * se,
    rel_period = period - 10
  )

# Plot 2: Event study coefficients
p2 <- ggplot(coef_data, aes(x = rel_period, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
             alpha = 0.2, fill = "#0072B2") +
  geom_line(color = "#0072B2", linewidth = 1) +
  geom_point(color = "#0072B2", size = 2.5) +
  annotate("rect", xmin = -Inf, xmax = -1, ymin = -Inf, ymax = Inf,
          fill = "green", alpha = 0.05) +
  annotate("text", x = -5, y = max(coef_data$ci_upper) * 0.9,
          label = "Pre-period\n(should be ≈0)", size = 3, hjust = 0.5) +
  labs(
    title = "Event Study Coefficients (Leads and Lags)",
    subtitle = "Pre-period estimates near zero validate parallel trends assumption",
    x = "Periods relative to treatment (0 = treatment period)",
    y = "Treatment Effect Estimate",
    caption = "Reference period: -1. Shaded region shows 95% confidence interval."
  ) +
  theme_antitrust() +
  theme(plot.title.position = "plot")

# Combine plots
p1 / p2

# Statistical test for pre-trends
pre_period_coefs <- coef_data |>
  filter(rel_period < 0, rel_period != -1)

cat("\nPre-trends test:\n")
cat(paste0("Number of pre-period coefficients: ", nrow(pre_period_coefs), "\n"))
cat(paste0("Mean pre-period coefficient: ",
          round(mean(pre_period_coefs$estimate), 3), "\n"))
# Joint Wald test that all pre-period leads (periods 1-8; reference is 9)
# are zero. fixest::wald() prints the F-statistic and p-value.
fixest::wald(event_study, keep = "period::[1-8]:")
```

**Interpretation:**
- **Pre-trends near zero**: Validates parallel trends assumption
- **Post-treatment divergence**: Confirms treatment effect
- **Confidence intervals**: Should include zero in pre-period
- **Use this**: Mergers (@sec-mergers), remedies (@sec-regulation-remedies), labor (@sec-labor-markets)

### Balance plots for matching/weighting
Show covariate balance before and after matching or weighting.

```{r}
#| label: diagnostic-balance
#| fig-cap: "Covariate balance diagnostic. Synthetic data for illustration only."
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8
library(dplyr)
library(ggplot2)
library(tidyr)

# Simulated covariate data (treatment and control groups)
set.seed(234)
n <- 200

covariates <- tibble(
  id = 1:n,
  treated = sample(c(0, 1), n, replace = TRUE),
  firm_size = rnorm(n, mean = 100 + 20 * treated, sd = 30),
  market_share = rnorm(n, mean = 0.15 + 0.05 * treated, sd = 0.08),
  years_operating = rnorm(n, mean = 15 + 5 * treated, sd = 7),
  revenue_growth = rnorm(n, mean = 0.10 + 0.03 * treated, sd = 0.12)
)

# Calculate standardized mean differences
calc_smd <- function(var, group) {
  mean_1 <- mean(var[group == 1])
  mean_0 <- mean(var[group == 0])
  sd_pool <- sqrt((var(var[group == 1]) + var(var[group == 0])) / 2)
  (mean_1 - mean_0) / sd_pool
}

balance_before <- covariates |>
  summarise(across(c(firm_size, market_share, years_operating, revenue_growth),
                  ~calc_smd(.x, treated),
                  .names = "{.col}")) |>
  pivot_longer(everything(), names_to = "variable", values_to = "smd") |>
  mutate(timing = "Before matching")

# Simulate "after matching" with reduced imbalance
# In practice, replace this with actual matched/weighted sample
balance_after <- balance_before |>
  mutate(
    smd = smd * 0.2,  # Matching reduces imbalance by 80%
    timing = "After matching"
  )

variable_levels <- c(
  "Firm size (employees)",
  "Market share",
  "Years operating",
  "Revenue growth"
)

balance_combined <- bind_rows(balance_before, balance_after) |>
  mutate(
    variable = recode(
      variable,
      "firm_size" = "Firm size (employees)",
      "market_share" = "Market share",
      "years_operating" = "Years operating",
      "revenue_growth" = "Revenue growth"
    ),
    variable = factor(variable, levels = variable_levels),
    timing = factor(timing, levels = c("Before matching", "After matching"))
  )

annotation_label <- tibble(
  abs_smd = 0.1,
  variable = variable_levels[1],
  label = "0.1 threshold\n(acceptable)"
)

# Love plot
ggplot(balance_combined, aes(x = abs(smd), y = variable,
                              color = timing, shape = timing)) +
  geom_vline(xintercept = 0.1, linetype = "dashed",
            color = "darkgreen", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray40") +
  geom_point(size = 3) +
  geom_text(
    data = annotation_label,
    aes(x = abs_smd, y = variable, label = label),
    inherit.aes = FALSE,
    hjust = -0.1,
    size = 3
  ) +
  scale_color_manual(values = c("Before matching" = "#D55E00",
                                 "After matching" = "#0072B2")) +
  scale_shape_manual(values = c("Before matching" = 16,
                                "After matching" = 17)) +
  scale_x_continuous(limits = c(0, max(abs(balance_combined$smd)) * 1.1)) +
  labs(
    title = "Covariate Balance: Love Plot",
    subtitle = "Standardized mean differences before and after matching",
    x = "Absolute Standardized Mean Difference",
    y = NULL,
    color = "Timing",
    shape = "Timing",
    caption = "SMD < 0.1 indicates acceptable balance (Austin, 2009)."
  ) +
  theme_antitrust() +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Summary table
cat("\nBalance summary:\n")
balance_combined |>
  select(variable, timing, smd) |>
  pivot_wider(names_from = timing, values_from = smd) |>
  mutate(improvement = abs(`Before matching`) - abs(`After matching`)) |>
  print()
```

**Interpretation:**
- **SMD < 0.1**: Acceptable balance (Austin, 2009)
- **Matching improves balance**: Lines should move toward zero
- **Use this**: Merger retrospectives, labor market studies, remedy evaluations

### Regression specification curve
Show robustness across multiple reasonable specifications.

```{r}
#| label: diagnostic-spec-curve
#| fig-cap: "Specification-curve diagnostic. Synthetic data for illustration only."
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 8
library(dplyr)
library(ggplot2)
library(fixest)

# Simulated data for multiple specifications
set.seed(345)
panel_spec <- expand.grid(
  firm = 1:100,
  period = 1:20
) |>
  mutate(
    treatment = ifelse(firm <= 50 & period >= 10, 1, 0),
    control1 = rnorm(n()),
    control2 = rnorm(n()),
    control3 = rnorm(n()),
    outcome = 50 + 5 * treatment + 2 * control1 + 1 * control2 +
              0.5 * control3 + rnorm(n(), 0, 3)
  )

# Run multiple specifications
specs <- list(
  "No controls" = feols(outcome ~ treatment | firm + period, panel_spec),
  "+ Control 1" = feols(outcome ~ treatment + control1 | firm + period, panel_spec),
  "+ Controls 1-2" = feols(outcome ~ treatment + control1 + control2 | firm + period, panel_spec),
  "+ All controls" = feols(outcome ~ treatment + control1 + control2 + control3 | firm + period, panel_spec),
  "No FE" = lm(outcome ~ treatment + control1 + control2 + control3, panel_spec),
  "Firm FE only" = feols(outcome ~ treatment + control1 + control2 + control3 | firm, panel_spec),
  "Period FE only" = feols(outcome ~ treatment + control1 + control2 + control3 | period, panel_spec)
)

# Extract coefficients
spec_results <- tibble(
  specification = names(specs),
  estimate = sapply(specs, function(m) {
    coefs <- coef(m)
    coefs[grepl("treatment", names(coefs))][1]
  }),
  se = sapply(specs, function(m) {
    if (inherits(m, "fixest")) {
      ses <- fixest::se(m)
    } else {
      ses <- summary(m)$coefficients[, "Std. Error"]
    }
    ses[grepl("treatment", names(ses))][1]
  })
) |>
  mutate(
    ci_lower = estimate - 1.96 * se,
    ci_upper = estimate + 1.96 * se,
    spec_num = row_number(),
    specification = factor(specification, levels = specification)
  )

# Specification curve plot
ggplot(spec_results, aes(x = spec_num, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
             alpha = 0.2, fill = "#0072B2") +
  geom_line(color = "#0072B2", linewidth = 1.2) +
  geom_point(color = "#0072B2", size = 3) +
  scale_x_continuous(
    breaks = spec_results$spec_num,
    labels = spec_results$specification
  ) +
  labs(
    title = "Specification Curve: Treatment Effect Robustness",
    subtitle = "Estimate remains positive and significant across specifications",
    x = NULL,
    y = "Treatment Effect Estimate",
    caption = "Each point represents a different model specification. Shaded area = 95% CI."
  ) +
  theme_antitrust() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title.position = "plot"
  )

# Summary statistics
cat("\nSpecification curve summary:\n")
cat(paste0("Range of estimates: ",
          round(min(spec_results$estimate), 2), " to ",
          round(max(spec_results$estimate), 2), "\n"))
cat(paste0("All specifications significant? ",
          ifelse(all(spec_results$ci_lower > 0), "Yes", "No"), "\n"))
cat(paste0("Median estimate: ", round(median(spec_results$estimate), 2), "\n"))
```

**Interpretation:**
- **Stable across specs**: Effect robust to model choices
- **All CIs exclude zero**: Consistent significance
- **Use this**: Expert reports, Daubert challenges, sensitivity analysis

### Residual diagnostics
Check model assumptions and identify influential observations.

```{r}
#| label: diagnostic-residuals
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 9
library(dplyr)
library(ggplot2)
library(patchwork)

# Fit model (use your actual model). Estimate on the full panel: treatment
# is treated x post, which varies within firm, so it survives the firm fixed
# effects. Subsetting to post-treatment periods would make treatment constant
# within firm and fixest would silently drop it.
model_data <- panel_spec

model <- feols(outcome ~ treatment + control1 + control2 | firm + period,
              data = model_data)

# Get residuals and fitted values
diagnostics <- model_data |>
  mutate(
    fitted = fitted(model),
    residual = resid(model),
    standardized_residual = residual / sd(residual)
  )

# Plot 1: Residuals vs. fitted
p1 <- ggplot(diagnostics, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(alpha = 0.5, color = "#0072B2") +
  geom_smooth(method = "loess", se = FALSE, color = "#D55E00") +
  labs(
    title = "Residuals vs. Fitted Values",
    subtitle = "Should show no pattern (validates linearity assumption)",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_antitrust()

# Plot 2: Q-Q plot for normality
p2 <- ggplot(diagnostics, aes(sample = standardized_residual)) +
  stat_qq(color = "#0072B2", alpha = 0.6) +
  stat_qq_line(color = "#D55E00", linewidth = 1) +
  labs(
    title = "Q-Q Plot: Residual Normality",
    subtitle = "Points should follow line for normal distribution",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_antitrust()

# Plot 3: Scale-location plot
p3 <- ggplot(diagnostics, aes(x = fitted, y = sqrt(abs(standardized_residual)))) +
  geom_point(alpha = 0.5, color = "#0072B2") +
  geom_smooth(method = "loess", se = FALSE, color = "#D55E00") +
  labs(
    title = "Scale-Location Plot",
    subtitle = "Should show horizontal line (validates homoscedasticity)",
    x = "Fitted Values",
    y = "√|Standardized Residuals|"
  ) +
  theme_antitrust()

# Plot 4: Histogram of residuals
p4 <- ggplot(diagnostics, aes(x = residual)) +
  geom_histogram(bins = 30, fill = "#0072B2", alpha = 0.8, color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#D55E00") +
  labs(
    title = "Distribution of Residuals",
    subtitle = "Should be approximately normal, centered at zero",
    x = "Residuals",
    y = "Count"
  ) +
  theme_antitrust()

# Combine plots
(p1 | p2) / (p3 | p4) + plot_annotation(
  title = "Residual Diagnostic Plots",
  caption = "Check for violations of regression assumptions"
)
```

**What to look for:**
- **Residuals vs. Fitted**: No pattern (confirms linearity)
- **Q-Q Plot**: Points follow line (confirms normality)
- **Scale-Location**: Horizontal line (confirms homoscedasticity)
- **Histogram**: Bell-shaped, centered at zero

### Power analysis visualization
Show statistical power across different sample sizes and effect sizes.

```{r}
#| label: diagnostic-power
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 7
library(dplyr)
library(ggplot2)
library(pwr)

# Calculate power for different scenarios
sample_sizes <- seq(50, 500, by = 25)
effect_sizes <- c(0.2, 0.5, 0.8)  # Small, medium, large (Cohen's d) [@cohen_1988]

power_grid <- expand.grid(
  n = sample_sizes,
  effect_size = effect_sizes
) |>
  rowwise() |>
  mutate(
    power = pwr.t.test(n = n, d = effect_size, sig.level = 0.05,
                      type = "two.sample")$power,
    effect_label = case_when(
      effect_size == 0.2 ~ "Small effect (d = 0.2)",
      effect_size == 0.5 ~ "Medium effect (d = 0.5)",
      effect_size == 0.8 ~ "Large effect (d = 0.8)"
    )
  ) |>
  ungroup()

# Smallest sample size on the grid reaching 80% power, per effect size.
# (A band filter like power >= 0.79 & power <= 0.81 would drop d = 0.8,
# whose power already exceeds 0.95 at the grid minimum.)
power_minimum <- power_grid |>
  filter(power >= 0.80) |>
  group_by(effect_label) |>
  slice_min(n, n = 1) |>
  ungroup()

# Power curve
ggplot(power_grid, aes(x = n, y = power, color = effect_label)) +
  geom_hline(yintercept = 0.80, linetype = "dashed",
            color = "darkgreen", linewidth = 0.8) +
  annotate("text", x = max(sample_sizes) * 0.7, y = 0.80,
          label = "80% power threshold", hjust = 0, vjust = -0.5,
          size = 3, color = "darkgreen") +
  geom_line(linewidth = 1.2) +
  geom_point(data = power_minimum, size = 3) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_color_manual(values = c(
    "Small effect (d = 0.2)" = "#D55E00",
    "Medium effect (d = 0.5)" = "#0072B2",
    "Large effect (d = 0.8)" = "#009E73"
  )) +
  labs(
    title = "Statistical Power Analysis",
    subtitle = "Sample size requirements for 80% power at α = 0.05",
    x = "Sample Size (per group)",
    y = "Statistical Power",
    color = "Effect Size",
    caption = "Points indicate minimum sample size for 80% power."
  ) +
  theme_antitrust() +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

# Summary table
power_summary <- power_minimum |>
  select(effect_label, n_required = n, power)

cat("\nMinimum sample size for 80% power:\n")
print(power_summary, n = Inf)
```

**Use this for:**
- Study design and sample size planning
- Explaining null results (underpowered?)
- Justifying sample selection in expert reports

## AI and Machine Learning in Antitrust

Machine learning is a prediction technology. It finds patterns in high-dimensional data and generates out-of-sample forecasts, often outperforming traditional parametric models. In antitrust work, prediction tasks arise more often than practitioners sometimes recognize: classifying markets as potentially collusive, forecasting demand systems for merger simulation, identifying anomalous bidding patterns in procurement data, and clustering millions of documents for efficient review. In all of these settings ML methods can improve on conventional approaches.

But prediction is not causation, and antitrust enforcement ultimately depends on causal claims. Estimating the overcharge from a cartel conspiracy, measuring the price effect of a consummated merger, or attributing competitive harm to exclusionary conduct — these tasks require the design-based econometric methods developed in earlier chapters of this book. No amount of predictive accuracy can substitute for a credible identification strategy when the question is "what would have happened absent the challenged conduct?" ML can improve the *inputs* to causal analysis, such as demand estimates and matched control groups, without replacing the causal framework itself. Researchers who treat ML as a complement to the methods in @sec-research-design, rather than a substitute, will find the most productive applications.

### ML for cartel screening

One of the most promising applications of machine learning in competition enforcement is the automated screening of markets for collusive behavior. Traditional cartel screens, discussed in @sec-cartels, rely on structural and behavioral markers — bid rotation patterns, price parallelism, variance reductions — applied to individual markets. ML methods can scale these screens across hundreds or thousands of markets simultaneously, flagging those that warrant deeper investigation by enforcement agencies.

**Supervised approaches** train classifiers on markets where cartel activity has been established (through leniency applications, successful prosecutions, or agency decisions) and use the learned patterns to score unscreened markets. Random forests and gradient-boosted trees are particularly well suited to this task because they handle nonlinear interactions among features naturally and provide variable importance measures that aid interpretability. Useful features include the coefficient of variation of bids, the number of bidders per tender, the bid spread (difference between highest and lowest bids), price-cost margins, market concentration indices, and indicators of bidder rotation. The challenge, as with any supervised method, is that the training data is censored: we only observe cartels that were detected, and detected cartels may differ systematically from undetected ones.

**Unsupervised approaches** — clustering algorithms, isolation forests, autoencoders — sidestep the labeling problem by identifying markets whose pricing patterns are anomalous relative to the broader population, without requiring prior knowledge of which markets were cartelized. These methods are valuable precisely when labeled training data is scarce, though they require careful calibration to avoid overwhelming investigators with false positives.

Several competition authorities have begun piloting ML screening tools in practice. Brazil's CADE has experimented with algorithmic screens for public procurement cartels. Korea's KFTC has developed machine-learning-based bid-rigging detection systems. The OECD has published guidance on computational screening methodologies, emphasizing that these tools are investigative aids — they identify candidates for investigation, not substitutes for the evidentiary standard required to establish an infringement.

The following code scaffold illustrates a basic random forest classifier for cartel screening:

```{r}
#| label: ml-cartel-screen
#| eval: false
library(ranger)
library(dplyr)

# One row per bid: the screens need every bidder's bid in each project,
# not just the winning award (the winning bid is constant within a project,
# so its within-project sd would be zero by construction)
bids <- read.csv("data/derived/cartel_cement_bids.csv")

# Feature engineering: within-project dispersion across all submitted bids
features <- bids |>
  group_by(project_id) |>
  summarise(
    cv_bids = sd(bid) / mean(bid),
    n_bidders = n(),
    spread = (max(bid) - min(bid)) / min(bid),
    cartel = first(cartel_period)  # label
  )

# Train random forest (importance = "impurity" is required for importance())
rf_model <- ranger(
  cartel ~ cv_bids + n_bidders + spread,
  data = features,
  probability = TRUE,
  num.trees = 500,
  importance = "impurity"
)

# Variable importance
importance(rf_model)
```

### Algorithmic collusion

A distinct and increasingly urgent concern is that pricing algorithms may themselves facilitate collusion — not through explicit coordination among firms, but through the autonomous convergence of algorithmic pricing strategies on supracompetitive equilibria. When competing firms deploy reinforcement-learning or Q-learning algorithms to set prices in repeated market interactions, experimental research has shown that these algorithms can learn to sustain prices above competitive levels without any direct communication between them [@oecd_algorithms_2017]. The algorithms effectively discover and implement tacit collusion strategies that would be difficult for human managers to coordinate.

This possibility poses a hard problem for antitrust enforcement. Traditional cartel law requires evidence of an agreement or concerted practice: the "smoking gun" communications, meeting records, or parallel conduct plus facilitating factors that investigators rely on to establish liability. When an algorithm arrives at collusive pricing through autonomous learning, there may be no human agreement to discover. The algorithm's pricing behavior *is* the evidence, and interpreting that behavior requires new empirical tools.

Researchers have proposed several approaches to detecting algorithmic collusion empirically. Granger causality tests can assess whether one firm's algorithmic pricing adjustments predict a competitor's subsequent price changes beyond what market fundamentals would explain. Structural break detection methods can identify whether the deployment of pricing algorithms coincided with a regime change in pricing patterns — a sudden increase in price levels, reduction in price variance, or acceleration of price adjustment speed. And A/B testing frameworks, where algorithmic pricing is deployed in some markets but not others, can provide quasi-experimental variation to estimate the causal effect of algorithmic pricing on market outcomes.

The enforcement gap remains significant. Most jurisdictions lack clear legal frameworks for addressing algorithmic tacit collusion that falls short of an explicit agreement. This is an area where the empirical methods will likely develop faster than the legal doctrine.

### NLP and document review

Natural language processing has become an indispensable tool for managing the enormous document productions that characterize modern antitrust investigations. A single merger review or cartel prosecution can generate millions of documents, and traditional manual review is prohibitively expensive and slow.

Topic modeling algorithms (such as Latent Dirichlet Allocation) can automatically organize large document collections into thematic clusters — grouping documents that discuss capacity planning, pricing strategy, competitor intelligence, or supply agreements. Named entity recognition can extract and link references to specific firms, executives, products, and markets across thousands of documents. Semantic search, powered by transformer-based language models, allows investigators to find documents that are conceptually related to a query even when they do not share exact keywords — for instance, finding documents that discuss "price discipline" even when the phrase used is "market stability" or "rational pricing."

These tools work best when linked to econometric findings. If a structural break analysis identifies a specific date range when pricing behavior changed (see @sec-cartels), NLP-assisted search can prioritize documents from that period, looking for contemporaneous discussions of competitor contacts, pricing agreements, or capacity restrictions. Quantitative and qualitative evidence reinforce each other here.

One caveat: NLP-assisted review is a triage and prioritization tool, not a replacement for human expert judgment. Privilege review, assessment of document authenticity, and the interpretive work of connecting documentary evidence to legal theories of harm all require experienced lawyers and economists. The technology accelerates the process of finding the needle in the haystack; it does not eliminate the need for expertise in understanding what the needle means.

### Evidentiary challenges

The use of ML methods in antitrust raises important questions about admissibility and transparency, particularly in U.S. litigation governed by the *Daubert* standard [@daubert_1993] and in other jurisdictions with analogous reliability requirements (see @sec-litigation-practice for the broader framework of expert testimony standards).

ML models face heightened scrutiny because of "black box" concerns: the worry that a model's predictions cannot be meaningfully explained or challenged. The concern is practical. Opposing counsel will argue, often persuasively, that a model whose internal logic cannot be articulated in plain language should not form the basis of a damages estimate or a liability finding. For this reason, practitioners should favor interpretable models when the output will face judicial scrutiny. Random forests with variable importance rankings, gradient-boosted trees with SHAP (SHapley Additive exPlanations) values, and regularized regression methods all provide mechanisms for explaining which features drive predictions and by how much. Deep neural networks, by contrast, are harder to defend in adversarial proceedings, however strong their predictive performance.

Documentation practices matter as much as model choice. Practitioners should maintain detailed records of model selection rationale, hyperparameter tuning procedures, cross-validation protocols, and out-of-sample performance metrics. The goal is to demonstrate that the modeling choices were principled rather than result-driven — that the analyst did not simply try dozens of specifications and report the one most favorable to their client's position.

The distinction between screening and estimation carries particular legal significance. Using ML to prioritize markets for investigation or to identify relevant documents is relatively uncontroversial — these are internal workflow decisions that do not directly determine liability or damages. Using ML to *estimate* an overcharge or to *quantify* competitive harm places the model's output squarely at issue in the proceedings and invites much more rigorous challenge. Practitioners should be clear about which role their ML tools are playing and adjust their transparency and validation practices accordingly.

Finally, the regulatory landscape itself is evolving. The EU's AI Act [@eu_ai_act_2024], which entered into force in 2024, may impose additional transparency and documentation requirements on algorithmic tools used in public enforcement proceedings, including competition investigations. Practitioners operating across jurisdictions should monitor these developments and ensure that their ML workflows satisfy both evidentiary standards and emerging regulatory obligations for algorithmic accountability.

## Additional references
- **FTC resources:** See [ftc.gov](https://www.ftc.gov/) for expert guidance and econometric standards.
- **Causal Inference: The Mixtape** ([mixtape.scunning.com](https://mixtape.scunning.com/)) for causal estimators and diagnostics.
- **OECD cartel guidance:** See [oecd.org/competition](https://www.oecd.org/daf/competition/) for screening methodologies.

Return to this appendix whenever a chapter references "see appendix template" so the workflow stays consistent across teams and jurisdictions.

