---
title: "Day 1 — Foundations"
subtitle: "Institutions, evidence, market definition, and market power"
author: "Antitrust Methods: Research and Practice · Practitioner short course"
format:
  pptx:
    slide-level: 2
execute:
  echo: true
  eval: false
  warning: false
---

# Module 1.1 — The competition-law landscape

## Why institutions come before methods

- Economic analysis answers to **legal standards, agency priorities, and procedural rules** — all three vary by jurisdiction.
- Get the frame right and your analysis speaks to the questions that *decide* a matter, not the ones that are merely interesting.
- Three questions to ask of any matter:
  - Who **enforces** the law here?
  - What must they **prove** (the legal elements)?
  - Where does the **economics** enter (which element does my number go to)?

## South Africa: the institutions

- **Competition Act 89 of 1998** (as amended, incl. the 2019 Amendment Act).
- **Competition Commission** — investigates, screens mergers, refers cases.
- **Competition Tribunal** — adjudicates (Pretoria); decides large mergers and prohibited-practice referrals.
- **Competition Appeal Court (CAC)** — hears appeals; several headline economic findings have been *set aside* on appeal.
- Distinctive: a statutory **public-interest test** in merger review, and **market inquiries** (Grocery Retail, Healthcare, Online Intermediation Platforms).
- Regional reach: conduct affecting **SADC** markets is in scope.

## The Competition Act, by conduct type

| Conduct | South Africa | United States | EU |
|:--|:--|:--|:--|
| Cartels / horizontal | §4 (per se: §4(1)(b)) | Sherman §1 | Art. 101 |
| Abuse of dominance | §8 | Sherman §2 | Art. 102 |
| Vertical restraints | §5; §8(1)(d) if dominant | Sherman §1 (RoR) | Art. 101 + VBER |
| Mergers | §12, §12A (+ public interest §12A(3)) | Clayton §7 | EUMR 139/2004 |
| Price discrimination | §9 (by dominant firm) | Robinson-Patman | Art. 102(c) |

::: notes
Vertical restraints fall under §5 (and §8(1)(d) for a dominant firm), **not** the per se horizontal provision §4(1)(b). This is a common slip.
:::

## Dominance thresholds are statutory in SA (§7)

Unlike the US "monopoly power" inference or EU practice, **SA fixes the thresholds in the statute**:

> - **≥ 45%** → presumed dominant
> - **35–45%** → dominant *unless* it shows it lacks market power
> - **< 35%** → dominant only if market power is *proven*

- US: "monopoly power", in practice inferred above ~70% share.
- EU: "dominance", from ~40% with entry barriers; rebuttable presumption at 50% (*AKZO*).
- **Why it matters:** in SA the share number does legal work directly — the economist's market-share estimate can decide the burden.

## The same conduct, three regimes

- **Object vs. effects:** EU Art. 101 treats "object" restrictions as presumptively restrictive (101(3) still formally available); US separates per se from rule of reason; SA §4(1)(b) lists per se horizontal practices.
- **Public interest:** SA uniquely weighs employment, SME participation, and spread of ownership (incl. by historically disadvantaged persons) in merger review.
- **Private enforcement:** US treble damages; SA/EU narrower (SA: follow-on civil claims after a Tribunal finding).
- **Habit:** map the same theory of harm to each regime's statutory test *early*.

# Module 1.2 — Research design & the evidence triad

## The evidence triad — the spine of the course

- Every matter is built from three kinds of evidence; good work integrates all three.
  - **Empirical** — quantitative estimation (prices, shares, diversion, damages).
  - **Documentary** — internal memos, board decks, emails ("hot docs"), discovery.
  - **Expert judgment** — translating the first two into the legal standard.
- We return to this triad in *every* session. Of any claim, ask: **which leg supports it?**

## The investigation workflow: intake → remedy

- **Intake** — complaint, leniency application, or merger notification; triage "hot docs" + quick analytics.
- **Scoping** — inventory data systems, prioritise custodians, run early screens (margins, shares/HHI, diversion).
- **Analysis** — test theories: market definition, unilateral/coordinated effects, DiD / IV / simulation, married to documents.
- **Decision & remedy** — statement of issues, then forward-looking remedy within institutional constraints.
- At each hand-off: does the next dollar buy **more data** or **better documents/testimony**?

## Mapping economics onto legal elements

:::: {.columns}
::: {.column width="50%"}
**Economic question**

- Relevant market
- Market power / dominance
- Theory of harm
- Efficiencies / justification
- Remedy
:::
::: {.column width="50%"}
**Legal element it serves**

- Defines shares, the field of analysis
- An element of every abuse / §2 / Art. 102 case
- The causal mechanism the agency must prove
- The rebuttal / §4(1) or 101(3) defence
- Must be effective **and** administrable by the Tribunal
:::
::::

## Code as demo — the multi-jurisdiction timeline

Book Ch. 1, figure `antitrust-timeline`. Run live or show the output.

```{r}
library(ggplot2); source("program/R/helpers.R")
milestones <- tibble::tribble(
  ~year, ~event,                ~type,      ~jurisdiction,
  1998,  "Competition Act 89",  "Statute",  "SA",
  2019,  "SA Amendment Act",    "Statute",  "SA",
  2023,  "US Merger Guidelines","Guidance", "US")
ggplot(milestones, aes(year, 0, colour = jurisdiction)) +
  geom_point(size = 4) + theme_antitrust()
```

**Point at:** SA's 1998 Act drew on EU precedent; the 2023 US Guidelines moved toward EU-style structural presumptions — convergence over time.

# Module 1.3 — Market definition

## The hypothetical monopolist test (SSNIP)

Start with a candidate market. Could a hypothetical monopolist over *just these products* profitably impose a **SSNIP** (a Small but Significant Non-transitory Increase in Price, typically 5–10%)?

> - **YES** → the candidate market **is** a relevant market
> - **NO** → widen the market and test again

- Mechanises the question "what constrains this firm?"
- Zero-price digital markets: use **SSNDQ** (a small but significant decrease in *quality* — ad load, data extraction).

## Critical loss — and the fallacy to avoid

A SSNIP is profitable if the sales actually lost are smaller than the most you could afford to lose:

> **Critical loss   CL = t / (t + m)**  — t = SSNIP size, m = gross margin

- Example: t = 5%, m = 40% → CL = 5/(5+40) = **11.1%**.
- ACTUAL loss < CL → SSNIP profitable → market is **well defined** (do not widen).
- ACTUAL loss > CL → widen the candidate market.
- **THE FALLACY:** high margins give a *low* critical loss — but high margins often go with *low actual loss* too. Citing only the low critical loss to claim a narrow market is the classic error (O'Brien–Wickelgren; Katz–Shapiro).

## Two traps the tribunal will test

- **The Cellophane fallacy.** At the already-monopolised price, other products look like substitutes — but only because the price is already supra-competitive. Define the market at the **competitive** price. Central wherever you start from an abuse/§8/Art. 102 allegation.
- **Brown Shoe practical indicia.** Beyond SSNIP arithmetic: peculiar uses, distinct customers, distinct prices, specialised vendors, industry recognition. Useful when the data are thin.

## Measuring concentration: HHI

> **HHI = Σ (share_i)²        ΔHHI = 2 × s₁ × s₂**

- Shares as percentages (monopoly = 10,000). Merger short-cut: ΔHHI = 2 × s₁ × s₂.
- **2023 US Merger Guidelines** structural presumption: post-merger HHI > **1,800** *and* ΔHHI > **100** (or combined share ≥ 30% with ΔHHI > 100).
- The 2010 Guidelines' 1,500/2,500 bands still appear in older case law — know which standard a cited case used.

## Code as demo — HHI in two lines

```{r}
source("program/R/helpers.R")
shares <- data.frame(firm = c("A","B","C","D"), share = c(0.40, 0.30, 0.20, 0.10))
calc_hhi_change(shares, merging_firms = c("C", "D"))
#> pre_hhi 3000 ; post_hhi 3400 ; delta_hhi 400  (= 2*20*10)
```

**Point at:** Pre 3,000; post 3,400; Δ 400. Both triggers met (>1,800 and >100) → structural presumption. The class warm-up.

## Case — Grocery Retail Market Inquiry (SA)

- Inquiry into national supermarket chains (Shoprite, Pick n Pay, Spar, Woolworths) and the long-term **exclusive lease agreements** in shopping centres.
- Exclusivity foreclosed entry by rivals and informal traders — a market-definition and barriers question, not price-fixing.
- Brings together SSNIP/critical-loss reasoning and the **public-interest** lens (SME / township-trader participation).

**Why it matters:** market definition is where the public-interest story and the competition story first meet in SA practice.

# Module 1.4 — The IO toolkit

## Elasticity and the Lerner index

Market power shows up in the price–marginal-cost gap. For a profit-maximising firm:

> **L = (P − MC) / P = − 1 / ε**

- L = Lerner index (the margin); ε = own-price elasticity of demand.
- Inelastic demand (small |ε|) → high margin → market power.
- Caveat: a high margin alone is not proof — it can reflect fixed-cost recovery or quality. Pair it with other evidence.

## Diversion ratios — the unilateral-effects engine

When product 1's price rises, where do its lost sales go?

> **D₁₂ = (sales gained by 2) / (sales lost by 1)**

- If 1 and 2 are owned by merging firms, sales "lost" to 2 are recaptured → less reason to compete on price (Day 3).
- Estimated from switching data, win/loss data, surveys, or a demand model. Feeds **UPP** and **GUPPI**.

## Pass-through — and a trap

- Pass-through = how much of a cost change reaches price. Converts an overcharge or cost-saving into a price effect (Days 2–3).
- **Estimate in changes / log-differences with lags — NOT on price/cost index LEVELS.** Regressing one trending index on another (CPI on PPI in levels) is the spurious-regression trap.
- Rule of thumb: a merger's first-order price effect ≈ pass-through × UPP (≈ UPP/2). State the assumption.

## Day 1 live-demo toolbox (book helpers)

- `calc_hhi_change(shares, merging_firms)` — HHI, post-HHI, ΔHHI.
- `run_logit_sim(products, merging_firms)` — static logit merger simulation (Day 3).
- `theme_antitrust()` / `scale_colour_antitrust()` — consistent figures.
- `fetch_fred()` — pull a real concentration/price series (needs key).
- All in `program/R/helpers.R`; each concept ≈ one function call.

## Day 1 — takeaways

1. Map any conduct to §4 / §5 / §8 / §9 / §12 and to its US and EU homes.
2. State the SA §7 dominance thresholds from memory and explain why the share does legal work.
3. Run the hypothetical-monopolist test, compute critical loss, avoid the Cellophane and critical-loss fallacies.
4. Compute HHI / ΔHHI and apply the 2023 structural presumption.
5. Read off the four IO parameters every later method consumes: elasticity, margin/Lerner, diversion, pass-through.

## Day 1 — exercises

1. **HHI & presumption** [checkable]: shares 40/30/20/10; two smallest merge.
2. **Critical loss** [checkable]: t = 5%, m = 35%, predicted loss 9%.
3. **Mapping memo** [discussion]: exclusive dealing by a 48%-share firm, SA/US/EU.
4. **Public interest** [discussion]: one §12A(3) example that flips a clearance.

*Worked answers in `teaching/exercises/day1_exercises.md`.*
