---
title: "Day 1 — Foundations"
subtitle: "Institutions, evidence, market definition, and market power"
author: "Antitrust Methods: Research and Practice · Practitioner short course"
format:
  pptx:
    slide-level: 2
execute:
  echo: true
  eval: false
  warning: false
---

# Module 1.1 — The competition-law landscape

## Why institutions come before methods

- Economic analysis answers to **legal standards, agency priorities, and procedural rules** — all three vary by jurisdiction.
- Get the frame right and your analysis speaks to the questions that *decide* a matter, not the ones that are merely interesting.
- Three questions to ask of any matter:
  - Who **enforces** the law here?
  - What must they **prove** (the legal elements)?
  - Where does the **economics** enter (which element does my number go to)?
- Same number, different jobs: a 48% share is *presumed dominance* in SA, *probably dominant* in the EU, *not yet monopoly power* in the US.
- The economist who skips this step produces beautiful estimates that answer the wrong legal question.

## South Africa: the institutions

- **Competition Act 89 of 1998** (as amended, incl. the 2019 Amendment Act).
- **Competition Commission** — investigates, screens mergers, refers cases, runs market inquiries.
- **Competition Tribunal** — adjudicates (Pretoria); decides large mergers and prohibited-practice referrals.
- **Competition Appeal Court (CAC)** — hears appeals; several headline economic findings have been *set aside* on appeal (Sasol, Media24).
- Distinctive: a statutory **public-interest test** in merger review, and **market inquiries** as a standing tool.
- Regional reach: conduct affecting **SADC** markets is in scope.

## The Competition Act, by conduct type

| Conduct | South Africa | United States | EU |
|:--|:--|:--|:--|
| Cartels / horizontal | §4 (per se: §4(1)(b)) | Sherman §1 | Art. 101 |
| Abuse of dominance | §8 | Sherman §2 | Art. 102 |
| Vertical restraints | §5; §8(1)(d) if dominant | Sherman §1 (RoR) | Art. 101 + VBER |
| Mergers | §12, §12A (+ public interest §12A(3)) | Clayton §7 | EUMR 139/2004 |
| Price discrimination | §9 (by dominant firm) | Robinson-Patman | Art. 102(c) |

::: notes
Vertical restraints fall under §5 (and §8(1)(d) for a dominant firm), **not** the per se horizontal provision §4(1)(b). This is a common slip.
:::

## Dominance thresholds are statutory in SA (§7)

Unlike the US "monopoly power" inference or EU practice, **SA fixes the thresholds in the statute**:

> - **≥ 45%** → presumed dominant
> - **35–45%** → dominant *unless* it shows it lacks market power
> - **< 35%** → dominant only if market power is *proven*

- The thresholds shift the **burden of proof**, not just the label.
- At ≥45% the firm must disprove dominance; in the 35–45% band it must disprove market power; below 35% the *complainant* must prove it.
- **Why it matters:** in SA the share number does legal work directly — the economist's market-share estimate can decide who carries the burden.

## Dominance abroad: a comparison

:::: {.columns}
::: {.column width="50%"}
**United States (Sherman §2)**

- "Monopoly power" — the power to control price or exclude competition.
- No statutory number; courts infer it, in practice above **~70%** share.
- *Alcoa* (90% = monopoly; 60–64% doubtful; 33% no).
- Always paired with durable entry barriers.
:::
::: {.column width="50%"}
**EU (Art. 102)**

- "Dominance" — to behave independently of competitors and customers.
- In practice from **~40%** with entry barriers.
- **AKZO**: a 50% share is a *rebuttable presumption* of dominance.
- Structure plus barriers, not share alone.
:::
::::

## Object vs. effects — the deep structure

- **EU Art. 101:** "object" restrictions (price-fixing, market-sharing) are presumptively restrictive — no effects proof needed, though 101(3) exemption is formally available.
- **US:** the **per se** vs. **rule of reason** split; naked horizontal restraints are per se, most else gets full effects analysis.
- **SA §4(1)(b):** a closed list of per se horizontal practices (price-fixing, market division, collusive tendering) — no efficiency defence at all.
- The economist's role **disappears on liability** for per se / object conduct and reappears only for **remedy and damages**. Knowing which regime you are in tells you whether your model matters.

## Public interest — the SA distinctive

- SA merger review weighs **competition AND public interest** as a co-equal limb (§12A(3)).
- The statutory limbs: **employment**, **SME participation**, **spread of ownership** (esp. by historically disadvantaged persons), effect on a **particular industrial sector or region**, ability of national industry to compete in export markets.
- A merger can be cleared on pure competition grounds and still be **conditioned or blocked** on public interest (Walmart/Massmart conditions; Burger King/Grand Parade prohibition).
- This widens the evidence the economist must marshal — beyond price and shares to employment and ownership effects.

## Market inquiries — the SA power tool

- A **market inquiry** is a forward-looking, industry-wide study with the power to impose remedies — not a litigated abuse finding.
- It suits **fast-moving or structurally troubled** markets where case-by-case litigation is too slow.
- Three landmark SA inquiries:
  - **Grocery Retail** — exclusive leases foreclosing entry by rivals and informal traders.
  - **Health Market Inquiry** — concentrated funders and facilities; recommended a supply-side regulator.
  - **Online Intermediation Platforms (OIPMI)** — ranking, self-preferencing, commissions on digital platforms.
- **Why it matters:** SA reaches harms via inquiry + remedy where the US would litigate — a different toolkit for the same theories.

## Burden and standard of proof

- **Standard:** competition matters are **civil** — proof on the **balance of probabilities** (not "beyond reasonable doubt"), except SA criminalised cartel conduct for individuals (§73A).
- **Burden shifts with structure:** SA §7 thresholds (above) move the burden; per se / object conduct removes the effects burden entirely.
- **Prima facie then rebuttal:** the agency makes out the elements; the firm raises efficiency / objective-justification defences.
- For the economist: know **which element your number serves** and **who must prove it** — that determines whether you are building the case or the rebuttal.

# Module 1.2 — Research design & the evidence triad

## The evidence triad — the spine of the course

- Every matter is built from three kinds of evidence; good work integrates all three.
  - **Empirical** — quantitative estimation (prices, shares, diversion, damages).
  - **Documentary** — internal memos, board decks, emails ("hot docs"), discovery.
  - **Expert judgment** — translating the first two into the legal standard.
- The three corroborate each other: a screen flags conduct, a document explains it, the expert ties it to the statutory element.
- We return to this triad in *every* session. Of any claim, ask: **which leg supports it?**

## The investigation workflow: intake → remedy

- **Intake** — complaint, leniency application, or merger notification; triage "hot docs" + quick analytics.
- **Scoping** — inventory data systems, prioritise custodians, run early screens (margins, shares/HHI, diversion).
- **Analysis** — test theories: market definition, unilateral/coordinated effects, DiD / IV / simulation, married to documents.
- **Decision & remedy** — statement of issues, then forward-looking remedy within institutional constraints.
- At each hand-off: does the next dollar buy **more data** or **better documents/testimony**?

## Mapping economics onto legal elements

:::: {.columns}
::: {.column width="50%"}
**Economic question**

- Relevant market
- Market power / dominance
- Theory of harm
- Efficiencies / justification
- Remedy
:::
::: {.column width="50%"}
**Legal element it serves**

- Defines shares, the field of analysis
- An element of every abuse / §2 / Art. 102 case
- The causal mechanism the agency must prove
- The rebuttal / §4(1) or 101(3) defence
- Must be effective **and** administrable by the Tribunal
:::
::::

## How to read an expert report (1): question & data

- **Research question.** What exactly is being estimated, and how does it link to the **legal element** and to the prior **literature**? A report with no clear question is advocacy in a lab coat.
- **The data.** Where do they come from (collection process, selection)? Beware sample selection.
- **Variables.** Distinguish the **outcome** (dependent) variable from the **explanatory** variables — and ask whether the explanatory ones are themselves caused by the outcome.
- **Unit of observation.** Firm? store? transaction? market-week? The unit determines what variation identifies the effect and how to cluster standard errors.

## How to read an expert report (2): the empirical strategy

- **Ideal data vs. available data.** What experiment *would* answer the question, and how far does the actual data fall short?
- **Identification.** What variation is being used to separate the causal effect from confounders? Name the source.
- **Sources of exogenous variation.** A policy change, a weather shock, a discontinuity, an instrument — something that moves the treatment but not the outcome directly.
- **How much identification comes from the model?** The more the answer leans on functional-form and distributional **assumptions** rather than on data variation, the more fragile it is — and the more you probe the assumptions.

## How to read an expert report (3): technique & conclusions

- **The main equation.** Write it down. What is the coefficient of interest, and in what units?
- **Exogenous vs. endogenous regressors.** Which right-hand-side variables are plausibly random, which are choices?
- **What is in the error term?** Anything correlated with the regressor and the outcome biases the estimate — name the omitted variables.
- **Instruments & validity.** Relevance (correlated with the endogenous variable) **and** exclusion (affects the outcome *only* through it). Exclusion is an assumption, not a test.
- **Conclusions & alternatives.** State the most plausible **alternative interpretation** of the same result — then ask whether the report rules it out.

## "Ideal data" thinking and the counterfactual

- Every causal claim is a comparison to a **counterfactual**: what would the outcome have been *but for* the conduct/merger?
- The counterfactual is never observed — we **estimate** it.
- The gold standard for building it is **random assignment**: treatment and control differ only by chance, so differences in outcomes are caused by treatment.
- Most competition data is **observational**: firms and prices are not randomly assigned, so we must engineer a credible counterfactual from the variation we do have.
- The whole methods course is: *how do we build a defensible counterfactual when we cannot randomise?*

## The credibility revolution

- A shift (1990s onward) from elaborate structural assumptions toward **research designs** that isolate exogenous variation.
- **RCTs** — randomise treatment directly (rare in antitrust, common in policy).
- **Natural experiments** — nature/policy randomises for you. *Mariel boatlift* (1980): a sudden labour-supply shock to Miami used to estimate immigration's wage effect.
- **Quasi-experiments** — a discontinuity or staggered rollout approximates randomisation.
- **Leamer's critique:** "Let's take the *con* out of econometrics" — fragile results that flip with the specification. The fix is **design, not post-hoc** robustness patches: a good design survives reasonable specifications.

## Observational data structures → methods

| Data structure | Typical method | What it controls |
|:--|:--|:--|
| Cross-section | Multiple regression w/ controls | Observed confounders only |
| Time series | Event study | Pre/post around a date |
| Panel (firm × time) | Fixed effects; **DiD** | Time-invariant unobservables + common shocks |
| Panel + comparison | Matching / PSM | Observable selection |
| Sharp threshold | **RDD** | Local randomisation at the cutoff |
| Instrument available | **IV** | Endogeneity of the regressor |

::: notes
Map the available data structure to the method *before* estimating. The design dictates the assumption you must defend.
:::

## Threats to validity, by design

- **RCT** — check **balance**: are treatment and control groups similar on pre-treatment covariates? Attrition can re-introduce selection.
- **IV** — the **exclusion restriction**: the instrument must affect the outcome *only* through the endogenous variable. Untestable; argue it institutionally.
- **DiD** — **parallel trends**: absent treatment, treated and control would have moved together. Inspect pre-trends; worry about differential shocks.
- **RDD** — **no bunching/manipulation** at the cutoff (McCrary density test); the effect is **local** to the threshold, not global.
- Naming the threat *for your own design first* is the mark of a credible report.

## Structural IO vs. reduced form

:::: {.columns}
::: {.column width="50%"}
**Structural (BLP-style)**

- Estimate **demand** (elasticities, diversion) + a **conduct** model, then **simulate** the merger.
- Predicts effects of an event that *hasn't happened*.
- Heavy on assumptions: functional form, equilibrium, instruments for prices.
- Used when no clean experiment exists (most prospective mergers).
:::
::: {.column width="50%"}
**Reduced form (clean design)**

- e.g. **Hastings (2004)** gasoline: a vertical change used as a natural experiment on retail prices.
- Transparent, few assumptions, credible *for the event observed*.
- Limited external validity; can't simulate counterfactual mergers.
- Used when good exogenous variation exists.
:::
::::

## Institutional details and the Scheffman best practice

- **Institutional details matter.** Contracts, regulation, procurement rules, and how prices are actually set often determine the right model — and sometimes settle the question without econometrics.
- **Scheffman (FTC) best practice:** the most persuasive evidence is frequently **descriptive quantitative** — prices, shares, margins, shipments, bid data — clearly presented.
- **Make assumptions least favourable to your own case.** If the conclusion holds under conservative assumptions, it is robust; if it needs the favourable ones, the tribunal will notice.
- Fancy methods do not rescue a weak descriptive picture; they amplify a strong one.

## Code as demo — the multi-jurisdiction timeline

Book Ch. 1, figure `antitrust-timeline`. Run live or show the output.

```{r}
library(ggplot2); source("program/R/helpers.R")
milestones <- tibble::tribble(
  ~year, ~event,                ~type,      ~jurisdiction,
  1998,  "Competition Act 89",  "Statute",  "SA",
  2019,  "SA Amendment Act",    "Statute",  "SA",
  2023,  "US Merger Guidelines","Guidance", "US")
ggplot(milestones, aes(year, 0, colour = jurisdiction)) +
  geom_point(size = 4) + theme_antitrust()
```

**Point at:** SA's 1998 Act drew on EU precedent; the 2023 US Guidelines moved toward EU-style structural presumptions — convergence over time.

# Module 1.3 — Market definition

## Why market definition comes first

- The relevant market **fixes the shares**, the field for HHI, the dominance inference, and the set of constraints on the firm.
- It is **contested**, not given: defendants widen it (more competitors, lower shares); agencies narrow it (fewer, higher).
- Two dimensions, tested the same way: **product** (which products compete?) and **geographic** (where?).
- A wrong market definition can sink an otherwise sound theory of harm — get it right *before* you estimate effects.

## The hypothetical monopolist test (SSNIP)

Start with a candidate market. Could a hypothetical monopolist over *just these products* profitably impose a **SSNIP** (a Small but Significant Non-transitory Increase in Price, typically 5–10%)?

> - **YES** → the candidate market **is** a relevant market
> - **NO** → widen the market and test again

- Mechanises the question "what constrains this firm?"
- Iterate: keep adding the next-closest substitute until a SSNIP would be profitable.
- Zero-price digital markets: use **SSNDQ** (a small but significant decrease in *quality* — ad load, data extraction).

## Critical loss — the arithmetic

A SSNIP is profitable if the sales actually lost are smaller than the most you could afford to lose:

> **Critical loss   CL = t / (t + m)**  — t = SSNIP size, m = gross margin

- Example: t = 5%, m = 40% → CL = 5/(5+40) = **11.1%**.
- ACTUAL loss < CL → SSNIP profitable → market is **well defined** (do not widen).
- ACTUAL loss > CL → widen the candidate market.
- The two ingredients: an **accounting margin** and a **demand estimate** of the actual loss.

## The critical-loss fallacy

- High margins give a **low** critical loss — which *looks* like it favours a narrow market.
- **But** high margins typically go with **low actual loss** too (a firm with pricing power faces inelastic demand).
- Citing only the low critical loss to claim a narrow market — without checking the actual loss — is the classic error (**O'Brien–Wickelgren; Katz–Shapiro**).
- The discipline: **always compare actual loss to critical loss**; never quote one in isolation.

## The Cellophane fallacy

- From *du Pont* (cellophane): at the **already-monopolised** price, other products look like substitutes — but only because the price is *already supra-competitive*.
- Define the market at the **competitive** price, not the prevailing one.
- Central wherever you start from an **abuse / §8 / Art. 102** allegation, because the conduct under challenge has already moved the price.
- The mirror image of the SSNIP: SSNIP starts from a competitive price and pushes up; the Cellophane trap starts from a monopoly price and is fooled by the substitution it induced.

## Brown Shoe practical indicia

- When the SSNIP arithmetic is data-starved, courts look to **qualitative indicia** (*Brown Shoe*, 1962):
  - Industry or public **recognition** of the market as separate.
  - The product's **peculiar characteristics and uses**.
  - **Distinct customers** and **distinct prices**.
  - **Sensitivity to price changes** (cross-elasticity).
  - **Specialised vendors** and distinct production facilities.
- These corroborate (or undercut) the econometrics — the documentary leg of the triad applied to market definition.

## Measuring concentration: HHI

> **HHI = Σ (share_i)²        ΔHHI = 2 × s₁ × s₂**

- Shares as percentages (monopoly = 10,000). Merger short-cut: ΔHHI = 2 × s₁ × s₂.
- **2023 US Merger Guidelines** structural presumption: post-merger HHI > **1,800** *and* ΔHHI > **100** (or combined share ≥ 30% with ΔHHI > 100).
- The 2010 Guidelines' 1,500/2,500 bands still appear in older case law — know which standard a cited case used.
- HHI is a **screen**, not a verdict: it flags matters for scrutiny; it does not by itself prove harm.

## Code as demo — HHI in two lines

```{r}
source("program/R/helpers.R")
shares <- data.frame(firm = c("A","B","C","D"), share = c(0.40, 0.30, 0.20, 0.10))
calc_hhi_change(shares, merging_firms = c("C", "D"))
#> pre_hhi 3000 ; post_hhi 3400 ; delta_hhi 400  (= 2*20*10)
```

**Point at:** Pre 3,000; post 3,400; Δ 400. Both triggers met (>1,800 and >100) → structural presumption. The class warm-up.

## Deep dive: FTC v. Staples / Office Depot (1997)

- The **model market-definition case** — the template for unilateral-effects-plus-econometrics in merger review.
- The **office-superstore (OSS) format**: one-stop shopping, **5,000–7,500 SKUs**, deep discounts (**30–70%** off list).
- The industry consolidated from ~**23 chains to 3**: Staples, Office Depot, OfficeMax. The merger would have left **two**.
- The fight was entirely about the **market**: the FTC said "office supplies sold *through superstores*"; the parties said "all office-supply retailing" (Walmart, Best Buy, mail order) — which would have made the merger trivial.

## Staples (1): market by retail format

- The relevant market was defined as **"consumable office supplies sold through office superstores"** — by **retail format**, not by product.
- A pen is a pen wherever sold — but the *competitive constraint* on an OSS comes overwhelmingly from **other OSS**, not from Walmart or a corner stationer.
- This was the conceptual leap: markets can be drawn around a **channel of distribution** when that is where the competitive interaction lives.
- It reframed "what is the product?" as "**what constrains the price the customer actually pays?**"

## Staples (2): four pillars of evidence

- **Distinct products/services** — the one-stop superstore offering is a different shopping experience from mass merchants and mail order.
- **Competitive recognition** — internal documents labelled markets "**competitive**" (another OSS present) vs. "**non-competitive**" (OSS-monopoly) and *priced accordingly*.
- **Pricing constraints** — non-OSS retailers (Walmart, Best Buy, warehouse clubs) did **not** discipline OSS prices.
- **Econometric evidence** — the hypothetical-monopolist regression predicted a merged firm could raise prices ~**8.49%**, exceeding the 5% SSNIP threshold (**18 months** of data, **400+ stores**, **40+ cities**).

## Staples (3): the price-comparison smoking gun

- Where **Staples faced another OSS**, prices were dramatically lower than where it had a local OSS monopoly — the firm's own data, from its own scanners.
- OSS-vs-OSS prices were **nearly identical** (Detroit basket: Staples **$100**, Office Depot **$100.4**, OfficeMax **$102**).
- A non-OSS like **Best Buy** ran **~18–19% higher** on the same basket → it was **not** in the same market.
- This is the **descriptive quantitative** evidence Scheffman prizes: the parties' own prices, plainly arrayed, did most of the work.

## Staples (4): holding & legacy

- **Judge Hogan granted a preliminary injunction** (D.D.C. 1997); the merger was abandoned.
- Memorable craft: **"store visits are worth a thousand affidavits"** — get out of the spreadsheet and into the aisle.
- Precedent for combining **unilateral-effects theory + econometric and internal-document evidence** — later echoed in **Oracle/PeopleSoft**, **Whole Foods/Wild Oats**, **T-Mobile/Sprint**.

**Why it matters:** it legitimised the modern merger toolkit — define by the constraint that actually binds, prove it with the parties' own data, and let documents corroborate the numbers.

## Tying it back to South Africa

- The **same SSNIP / critical-loss logic** travels — but SA adds the **public-interest lens**.
- In the **Grocery Retail Market Inquiry**, the issue was not price-fixing but **long-term exclusive leases** foreclosing entry by rivals and informal/township traders.
- Market definition met barriers-to-entry and the spread-of-participation limb in one matter.

**Why it matters:** market definition is where the public-interest story and the competition story first meet in SA practice.

# Module 1.4 — The IO toolkit

## What is market power?

- **Market power** = the ability to **profitably**:
  - raise **price** above marginal cost, or
  - restrict **output**, or
  - degrade **quality / innovation** (the digital-market form).
- It is a **matter of degree**, not a switch — every firm has some; the question is *how much* and whether it is *durable*.
- **Dominance / monopoly power** is market power large and durable enough to clear the legal threshold (SA §7; EU AKZO; US §2).
- Everything in this module is a way to **measure** that gap or its sources.

## The Lerner index

Market power shows up in the price–marginal-cost gap. For a profit-maximising firm:

> **L = (P − MC) / P = − 1 / ε**

- L = Lerner index (the margin); ε = own-price elasticity of demand.
- Worked example: |ε| = 2 → L = 0.5 → price is **double** marginal cost. |ε| = 5 → L = 0.2.
- Inelastic demand (small |ε|) → high margin → market power.
- Caveat: a high margin alone is not proof — it can reflect **fixed-cost recovery** or quality. Pair it with other evidence.

## Measuring margins in practice

- **Price-cost margin (PCM):** accounting proxy for the Lerner index from financial data — but accounting cost ≠ marginal cost.
- **Tobin's Q:** market value / replacement cost of assets; Q ≫ 1 hints at rents or intangibles, but is noisy and not market-specific.
- **Production-function markups** (De Loecker–Warzynski): μ = **output elasticity / cost share** of a flexible input — estimable without observing marginal cost directly.
- Each measure trades **availability** against **fidelity to the true marginal cost**; triangulate, don't rely on one.

## The rise-of-markups debate

- **De Loecker–Eeckhout–Unger:** estimated average US markups rose from **~21% over cost (1980)** to **~61% (2016)**, concentrated in the right tail.
- A flashpoint: does it show rising **market power**, or something more benign?
- **Alternative explanations:**
  - Rising **fixed/intangible** costs (software, R&D, branding) → high markups *over marginal cost* but normal profits.
  - **Quality** improvements and changing **product mix**.
  - **Globalisation** and superstar-firm efficiency.
  - **Measurement**: using cost of goods sold vs. total cost changes the answer.
- Lesson: a striking aggregate trend still needs a **mechanism** before it becomes a finding.

## Sources of market power

- **Scale economies** — high fixed costs, declining average cost (utilities, networks).
- **Product differentiation** — brand, location, taste; softens price competition.
- **Network effects** — value rises with users (platforms, payment systems).
- **Switching costs / lock-in** — data, learning, contracts.
- **Intellectual property** — patents, copyrights, trade secrets.
- **Geography & regulation** — the only hospital in a region; licensing limits on entry.
- **Data** and **strategic conduct** — accumulated data as a barrier; exclusionary tactics that *create* power.

## Oligopoly: Cournot vs. Bertrand

:::: {.columns}
::: {.column width="50%"}
**Cournot (quantity)**

- Firms choose **output**; price clears the market.
- Prices fall between monopoly and competitive as N rises.
- Margins ∝ HHI / elasticity — links **structure to conduct**.
- Fits homogeneous goods with capacity choices.
:::
::: {.column width="50%"}
**Bertrand (price)**

- Firms choose **price**; lowest price wins.
- **Bertrand paradox:** with identical goods, two firms → price = MC.
- **Differentiation** breaks the paradox → positive margins.
- Fits branded consumer goods (most merger work).
:::
::::

- **The strategic variable matters:** the same structure implies very different power depending on whether firms compete on price or quantity. State which model your method assumes.

## Concentration measures

- **CR4 / CRn** — combined share of the top *n* firms; simple, but ignores the distribution among them.
- **HHI** — sum of squared shares; weights large firms more, so it captures both number and inequality of firms.
- Both are **structural screens**: high concentration **raises the question**, it doesn't answer it.
- They are inputs to (Cournot) conduct predictions, not substitutes for an effects analysis — and they depend entirely on the **market definition** (Module 1.3).

## The four parameters every later method consumes

- **Elasticity (ε)** — how quantity responds to price; the foundation of demand and the Lerner index.
- **Margin / Lerner (L)** — the price–cost gap; sizes the incentive to raise price.
- **Diversion ratio (D₁₂)** — where lost sales go; the engine of unilateral effects (UPP/GUPPI, Day 3).
- **Pass-through** — how much of a cost change reaches price; converts overcharges and savings into price effects.
- Estimate **pass-through in CHANGES, not levels** — regressing one trending index on another (CPI on PPI in levels) is the **spurious-regression trap**.

## Diversion and pass-through, defined

> - **D₁₂ = (sales gained by 2) / (sales lost by 1)**
> - **Pass-through ρ = Δprice / Δcost** (estimate in changes / log-differences)

- If 1 and 2 are owned by **merging firms**, sales "lost" to 2 are **recaptured** → less reason to compete on price (Day 3).
- Diversion is estimated from switching data, win/loss records, surveys, or a demand model; it feeds **UPP** and **GUPPI**.
- Rule of thumb (linear demand): a merger's first-order price effect ≈ **pass-through × UPP** (≈ UPP/2). Always state the assumption.

## Day 1 live-demo toolbox (book helpers)

- `calc_hhi_change(shares, merging_firms)` — HHI, post-HHI, ΔHHI (pass a data.frame).
- `run_logit_sim(products, merging_firms)` — static logit merger simulation (Day 3).
- `theme_antitrust()` / `scale_colour_antitrust()` — consistent figures.
- `fetch_fred()` — pull a real concentration/price series (needs key).
- All in `program/R/helpers.R`; each concept ≈ one function call.

# Close

## Day 1 — takeaways

1. Map any conduct to §4 / §5 / §8 / §9 / §12 and to its US and EU homes; know who enforces and what they must prove.
2. State the SA §7 dominance thresholds from memory and explain why the share does legal work (it shifts the burden).
3. Read an expert report by its design: question, data, identification, technique, and the plausible alternative interpretation.
4. Build a counterfactual — randomisation ideal, natural/quasi-experiments in practice; name the threat to validity for your own design.
5. Run the hypothetical-monopolist test, compute critical loss, and avoid the Cellophane and critical-loss fallacies (Staples as the model case).
6. Compute HHI / ΔHHI, apply the 2023 structural presumption, and read off the four IO parameters: elasticity, margin/Lerner, diversion, pass-through.

## Day 1 — exercises

1. **HHI & presumption** [checkable]: shares 40/30/20/10; two smallest merge.
2. **Critical loss** [checkable]: t = 5%, m = 35%, predicted loss 9%.
3. **Mapping memo** [discussion]: exclusive dealing by a 48%-share firm, SA/US/EU.
4. **Public interest** [discussion]: one §12A(3) example that flips a clearance.
5. **Optional R lab** [discussion]: read a real concentration series; a rising ratio is a flag, not a finding.

*Worked answers in `teaching/exercises/day1_exercises.md`.*
