---
title: "Day 2 — Conduct"
subtitle: "Cartels, abuse of dominance, labour markets, and digital platforms"
author: "Antitrust Methods: Research and Practice · Practitioner short course"
format:
  pptx:
    slide-level: 2
execute:
  echo: true
  eval: false
  warning: false
---

# Module 2.1 — Cartels and collusion

## What a cartel case has to establish

- An **agreement** (or concerted practice) among competitors to fix price, allocate markets/customers, rig bids, or restrict output — §4(1)(b) in SA, per se under Sherman §1, "by object" under Art. 101.
- Per se / by object: no need to prove market power or effects — the agreement *is* the violation.
- The hard part is proving the agreement. **Leniency** turns insiders into witnesses; economics corroborates.
- Economist's two jobs: **screens** that flag conduct, and **quantifying the overcharge** for damages.

## Cartel screens — before you have a confession

- **Variance / structural-break screens.** Cartels stabilise prices, then prices jump when the cartel breaks. Bai–Perron and CUSUM tests date the breaks.
- **Bid-rotation / allocation screens.** Who wins in turn, suspiciously similar losing bids, tight spreads. Needs **bid-level** data (every bid), not just who won.
- **Margin / parallel-pricing screens.** Co-moving prices beyond what common costs explain.
- A screen is an investigative aid, never proof — it survives cross-examination only when tied to documents and testimony.

## Quantifying the overcharge (before/after)

> - **Overcharge $ = (P_cartel − P_benchmark) × volume**
> - **Damages = pass-through × overcharge rate × commerce**

- Benchmark from a clean before/after window (validate with a break test) or a yardstick market.
- Pass-through matters when buyers are intermediaries (Day 1 trap: estimate in *changes*).
- US: **treble** damages. SA: follow-on civil claims after a Tribunal finding.

## Standing & the passing-on defence — the split

- **United States.** *Illinois Brick*: only **direct** purchasers may claim federal damages. *Hanover Shoe*: defendant may **not** reduce damages by arguing the overcharge was passed on. (Many states differ.)
- **EU and South Africa.** The passing-on defence **is** available, and indirect purchasers can in principle claim — so pass-through is central, not barred.
- First thing a litigator checks — get the forum's rule right before you model.

## Case — the bread cartel (SA)

- Premier Foods applied for **leniency**, implicating Tiger Brands, Pioneer and others in fixing the price of bread (2007).
- Documentary + testimonial evidence established the agreement; pricing analysis sized the overcharge on a staple food.
- Anchored SA follow-on damages and the public salience of cartels (Children's Resource Centre class action).

**Why it matters:** the template SA cartel matter — leniency cracks it, economics sizes the harm, and the harm lands on poor households.

## Case — construction "fast-track" settlement (SA)

- The Commission's fast-track process (settled 2013) covered **bid rigging** on major projects, incl. 2010 World Cup stadiums.
- 15 firms settled for penalties totalling about **R1.46 billion**.
- Shows the bid-rotation screen logic at industry scale, resolved by a structured settlement.

## Code as demo — a before/after overcharge

```{r}
p_cartel    <- mean(price[period == "cartel"])
p_benchmark <- mean(price[period == "after"])
overcharge_rate <- (p_cartel - p_benchmark) / p_benchmark
damages_single  <- overcharge_rate * commerce
damages_treble  <- 3 * damages_single   # US only
```

**Point at:** validate the benchmark window with a Bai–Perron break test before you trust the gap.

# Module 2.2 — Abuse of dominance

## The structure of an abuse case

- **Step 1 — Dominance.** SA §7 thresholds (Day 1); EU ~40%+ with barriers, AKZO at 50%; US §2 ~70%+.
- **Step 2 — Abusive conduct.** Exclusionary (foreclosing rivals) or exploitative (**excessive pricing** — live in SA/EU, largely not US).
- **Step 3 — Effects & justification.** Anticompetitive effect vs. objective justification / efficiency.
- The economist supplies the foreclosure measure, the cost benchmark, or the price-vs-competitive-level comparison.

## Predatory pricing: the as-efficient-competitor logic

Pricing below cost to exclude, expecting to recoup. Two legs:

> - **Leg 1:** Price < cost benchmark (AVC / AAC / LRAIC)
> - **Leg 2 (US):** recoupment must be plausible (*Brooke Group*)

- EU *AKZO*: below AVC presumed predatory; AVC–ATC predatory if part of an exclusion plan.
- US adds a **recoupment** screen; SA looks to cost benchmarks without a strict recoupment gate.
- **AEC test:** would a rival as efficient as the dominant firm survive these prices?

## Margin squeeze (vertically integrated firm)

A firm dominant in an input squeezes downstream rivals who must buy that input from it:

> **Squeeze if:  P_retail − P_wholesale < downstream cost**
> (the firm's own retail arm couldn't profit at its own wholesale price)

- The AEC test applied to two levels. Classic in telecoms (local-loop access).
- SA: the **Telkom** wholesale-broadband matters.
- **ECPR** is the benchmark for the access price (Day 3, remedies).

## Exclusion: dealing, tying, refusal to supply

- **Exclusive dealing / loyalty rebates.** Foreclose rivals' access; assessed on foreclosed share + AEC for rebates.
- **Tying / bundling.** Leverage power in one product to a second.
- **Refusal to deal / essential facilities.** Narrow in the US after *Trinko*; recognised in the EU (*Bronner / IMS*: indispensability, elimination of competition, no objective justification); SA uses §8 essential-facility provisions more readily than US law.

## Case — Sasol, excessive pricing (SA)

- Sasol Chemical Industries found by the Tribunal to have charged excessive prices for polymers to local converters.
- On appeal the **CAC SET ASIDE the finding (2015)**, faulting the economic-value / cost benchmark.
- Excessive pricing (§8(1)(a)) requires a price bearing no reasonable relation to economic value — notoriously hard to prove.

**Why it matters:** teach it as good law on the **test**, not as standing liability — SA courts have set findings aside on exactly that point.

## Case — Media24, exclusionary pricing (SA)

- Media24's *Forum* community newspaper alleged to have priced to drive out rival *Gold–Net News* in Welkom, then closed *Forum*.
- Tribunal found abuse (2015) on a recoupment-flavoured theory; the **CAC overturned the finding (2019)**.
- Refined the SA cost benchmark for predation (average avoidable cost vs. average total cost).

**Why it matters:** a second SA appellate reversal — carry the appellate history; the audience will know it.

## Comparative — the AEC framework is contested abroad

- **EU Intel:** the 2009 rebates decision and its AEC analysis were **annulled** (GC 2022, affirmed CJEU Oct 2024). Don't cite "the AEC framework established in *Intel*" as settled.
- **US Microsoft (2001):** the D.C. Circuit **vacated** the tying ruling; the liability that stood was **§2 monopoly maintenance**. Cite it for maintenance, not tying.
- Lesson: exclusion tests are live and jurisdiction-specific — anchor your method in the forum's *current* law.

# Module 2.3 — Antitrust and labour markets

## Monopsony and the wage markdown

A dominant employer faces an upward-sloping labour supply curve, so it holds wages below the value workers add:

> **wage / MRPL = η / (1 + η)**   — η = firm-level labour-supply elasticity

- MRPL = marginal revenue product of labour.
- η = 2 → wage = 2/3 of MRPL → a **33% markdown**. As η → ∞ (competitive), wage → MRPL.
- A firm can be tiny in its product market yet dominant as a **buyer** of local labour (the only large rural hospital).

## Labour-market concentration (labour HHI)

Define the market as **occupation × commuting zone**, then apply the same HHI:

> **Labour HHI = Σ (employment share_i)²**

- 2023 guidelines use the **1,800** threshold (2010 used 2,500).
- Many US local labour markets exceed 3,000 (Azar–Marinescu–Steinbaum).
- Same arithmetic as product markets; the work is defining the geography (commute) and occupation.

## Conduct: no-poach, wage-fixing, non-competes

- **No-poach** between employers (incl. across franchisees) — US DOJ treats naked no-poach and wage-fixing as **per se criminal**.
- **Non-competes** bind workers post-employment (~18% of US workers); reduce mobility and shave wages. The FTC's 2024 ban was **set aside** (*Ryan v. FTC*, 2024); appeal dropped (2025) → case-by-case enforcement.
- Reality check: *Jindal* (physical therapists) — jury **acquitted** on Sherman counts; *DaVita* — defendants **acquitted** (2022). Hard to win criminally, but reshaped compliance.
- **SA hook:** the Healthcare Market Inquiry documented a few hospital groups (>80% of beds) — a structural **monopsony** over nurses in many regions.

## Code as demo — labour HHI and markdown

```{r}
shares   <- c(50, 30, 20)          # hiring shares, %
hhi      <- sum(shares^2)          # 3800 -> concentrated
eta      <- 2                      # labour-supply elasticity
markdown <- 1 - eta / (1 + eta)    # 0.333 -> 33% below MRPL
```

**Point at:** HHI 3,800 (> 1,800) and a 33% markdown — concentration and the wage gap tell the same story.

# Module 2.4 — Digital markets and platforms

## What makes platforms different (and what doesn't)

- **Multi-sided:** value depends on the other side; network effects + switching costs feed power.
- **Zero price** on one side breaks the SSNIP — use **SSNDQ** (ad load, data extraction) and watch the other side's price.
- But the theories of harm are old friends: **self-preferencing**, exclusion via **defaults**, tying, refusal to interoperate.
- Defaults move behaviour even when switching is nominally free — the central empirical fact of the search cases.

## Measuring self-preferencing and defaults

> - **Ranking harm:  Δclicks = (CTR_top − CTR_demoted) × queries**
> - **Default effect: event-study / DiD around a choice-screen change**

- Top organic result captures ~25–35% of clicks; demotion to ~position 4 is a large traffic loss even at equal quality.
- Self-preferencing test: regress rank/visibility on quality **and** an own-product dummy — the dummy is the favouritism.

## The record is now 2024–2026 — keep it current

- **US v. Google (search):** liability Aug 2024; remedies Sept 2025 — default-exclusivity bans + data-sharing, **no Chrome divestiture**.
- **US v. Google (ad tech):** §2 liability 2025.
- **EU:** Google Shopping affirmed (CJEU 2024); DMA fines — Apple €500M, Meta €200M (Apr 2025); ex ante gatekeeper rules vs. US case-by-case.
- **Epic v. Apple:** 2025 contempt ruling barred the 27% off-app surcharge and anti-steering "scare screens".

## Case — OIPMI (SA)

- Final report (2023): e-commerce (Takealot ranking vs. its own retail arm), food delivery (commissions ~25–30%), online travel, Google's >90% search share.
- Remedies: fair-ranking reporting, anti-steering relief for restaurants, data/visibility transparency.
- A **market-inquiry remedy package**, not a litigated abuse finding — the SA tool of choice for fast-moving digital markets.

**Why it matters:** SA reaches platform harms via market inquiries with forward-looking remedies — in parallel with the EU DMA, in contrast to US litigation.

## Day 2 — takeaways

1. Lay out a cartel case: per se status, a screen, a before/after overcharge — and whether passing-on is available.
2. Structure an abuse case (dominance → conduct → justification) and apply the AEC test.
3. Carry the SA appellate history: Sasol set aside (2015), Media24 overturned (2019); treat Intel's AEC analysis as unwound.
4. Compute a labour HHI and a wage markdown.
5. Apply old theories of harm to platforms with SSNDQ and ranking/default measurement; place SA inquiries against the DMA and US litigation.

## Day 2 — exercises

1. **Overcharge & damages** [checkable]: commerce R200m, rate 15%.
2. **Wage markdown** [checkable]: hiring shares 50/30/20, η = 2.
3. **Ranking harm** [checkable]: CTR 35%→8%, demotion 1→4 on 1,000,000 queries.
4. **Margin-squeeze strategy** [discussion]: an SA telco accused of squeeze.

*Worked answers in `teaching/exercises/day2_exercises.md`.*
