---
title: "Day 3 — Mergers, Remedies & Litigation"
subtitle: "Predicting effects, fixing harm, IP, and proving it in a tribunal"
author: "Antitrust Methods: Research and Practice · Practitioner short course"
format:
  pptx:
    slide-level: 2
execute:
  echo: true
  eval: false
  warning: false
---

# Module 3.1 — Merger analysis and simulation

## Mergers are prospective — the job is prediction

- Conduct cases look backwards at what a firm *did*; merger review looks **forward** at what a combined firm *will do*. The evidence is largely predictive, not historical.
- Three economic theories of harm, plus — in SA — a **fourth, non-competition question** that can override them.
- The agency carries an initial burden (a **prima facie** case); the parties then **rebut** with efficiencies, entry, or failing-firm arguments.
- Outputs are not just "block / clear": the realistic menu is **clear, condition (remedy), or prohibit**.
- Screen cheap, quantify dear: GUPPI and diversion screens come first; full simulation only where the screen flags concern.

## The three theories of merger harm

- **Unilateral effects.** The merged firm raises price on its own because sales lost to a price rise are recaptured by the partner (diversion). No coordination needed.
- **Coordinated effects.** The merger makes tacit collusion among the remaining firms easier — fewer players, more symmetry, more transparency, or removal of a maverick.
- **Vertical / conglomerate effects.** Input foreclosure, customer foreclosure, or leverage/bundling across products.
- Plus the SA fourth question that can override all three: the **public-interest test (§12A(3))** — employment, SMEs, spread of ownership, ability of national industry to compete.
- A clean matter names *which* theory it is running early — the data you collect differ for each.

## Unilateral effects: UPP and GUPPI

> - **UPP_A = D_AB × (P_B − C_B) − E_A**
> - **GUPPI_A = D_AB × M_B × (P_B / P_A)**

- D_AB = diversion A→B; (P_B − C_B) = the **partner's** dollar margin; E_A = efficiency credit. Use the partner's margin, not A's own.
- GUPPI expresses the same pressure as a % of A's price (M_B = partner's % margin; (P_B/P_A) rescales).
- Rule of thumb: GUPPI above ~5% warrants a closer look. First-order price rise ≈ pass-through × UPP (≈ UPP/2).
- The economics: each lost sale that *returns* to the family is no longer a cost of raising price — the merged firm internalises the externality.

## Diversion ratios — where the lost sales go

> **D₁₂ = (sales gained by 2) / (sales lost by 1 when P₁ rises)**

- Diversion is the **direction and strength of competition** between two products — the single most load-bearing number in unilateral-effects work.
- Sources: switching/win-loss data, customer surveys (the "second-choice" question), a demand model, or — weakest — **share-proportional** diversion (D ≈ s_B / (1 − s_A)).
- Patient-flow, geographic catchment, and second-choice data usually beat share-proportional; when supplied diversion exceeds the proportional benchmark, the parties are **closer** rivals than shares imply.
- Aggregate diversion (sum to all merging products) gauges total recapture and the overall upward pricing pressure.

## Code as demo — a logit merger simulation

```{r}
source("program/R/helpers.R")
products <- data.frame(
  product = c("A","B","C"), firm = c("A","B","C"),
  price = c(100,95,90), share = c(.4,.35,.25), mc = c(60,58,55))
run_logit_sim(products, merging_firms = c("A","B"))
#> pre/post price, % change, ownership matrix
```

**Point at:** simulation needs a demand model + margins; GUPPI needs only diversion + margin. **Screen with GUPPI, quantify with simulation.**

## Coordinated effects — the maverick and the conditions

- **Maverick-firm theory:** harm often comes from the merger **removing or absorbing a maverick** — the disruptive price-cutter that destabilises tacit coordination.
- Facilitating factors (the checklist): **transparency** of prices/terms, product **homogeneity**, **symmetric** firms and costs, stable demand, and **high entry barriers**.
- A workable tacit scheme needs three things: reach a common understanding, **detect** deviation, and **punish** it credibly.
- Evidence: count the firms, measure symmetry, look for price-announcement / information-exchange practices, and ask whether the target *is* the maverick.
- Coordinated-effects theories are harder to prove than unilateral ones — courts want a plausible mechanism, not just "few firms".

## The merger review process — burden and outcomes

- **Notification:** turnover/asset thresholds trigger mandatory filing (SA: large/intermediate mergers to the Commission; US: HSR; EU: EUMR turnover tests).
- **Second phase:** a "second request" (US) / Phase II (EU) / referral to the Tribunal (SA large mergers) — deeper economic analysis on the flagged theory.
- **Burden:** agency makes a **prima facie** case (often via the structural presumption); the parties then **rebut**.
- **Standard:** US Clayton §7 "substantially lessen competition"; EU **SIEC** (significant impediment to effective competition); SA §12A + the §12A(3) public-interest overlay.
- **Outcome menu:** clear unconditionally / clear with **conditions** (structural or behavioural) / **prohibit**.

## Efficiencies and entry — the rebuttal

- Efficiencies can offset price pressure, but must be **merger-specific, verifiable, and passed through** — burden on the parties.
- **Marginal-cost** savings discipline price; **fixed-cost** savings usually do not (they don't change the pricing incentive).
- Entry rebuts harm only if **timely, likely, and sufficient** to deter or counteract the price rise.
- The **failing-firm** defence is narrow: the firm must be failing, with no less-anticompetitive buyer and assets otherwise exiting.
- In SA, efficiencies and entry sit alongside — and can be outweighed by — the **public-interest** assessment.

## Deep dive — the Staples / Office Depot model (US)

- The FTC's challenge rested on a **large-scale price regression**: ~400+ stores × ~18 months, regressing prices on the **number of office-superstore competitors** in each local market.
- Result: prices were materially higher where Staples faced no superstore rival — the model predicted roughly an **~8.49% price increase** going from three superstore firms to one (effective monopoly).
- An **event study** on the merger announcement / FTC challenge dates corroborated the direction: stock-price moves were consistent with the market expecting the merger to *raise* prices.
- The court enjoined the merger — an early, influential case of an **econometric model carrying the day** on unilateral effects.
- **Why it matters:** a transparent, replicable price regression on real transaction data can beat a market-definition argument — example-led merger economics at its strongest.

## SA public interest deep dive — Walmart / Massmart

- Acquisition (2011, case **73/LM/Nov10**) raised few horizontal competition concerns.
- The battle was **public interest**: local suppliers and jobs. Approved with conditions — a **supplier-development fund**, honouring labour agreements, and **limits on merger-specific retrenchments** (with some reinstatements).
- The conditions were litigated up to the **CAC**, which refined how public-interest conditions are framed and supervised over time.
- **Why it matters:** a merger can pass the competition test and still be reshaped — or saved — by public-interest conditions.

## SA public interest deep dive — Burger King / Grand Parade

- A foreign private-equity buyer sought the SA Burger King franchise from Grand Parade.
- The Commission **prohibited it (2021)** because it cut shareholding by **historically disadvantaged persons** from a high level toward zero — the first SA merger blocked **solely** on the spread-of-ownership limb of public interest.
- Later **approved** once the parties offered an **employee/HDP ownership and worker commitment**.
- **Why it matters:** public interest is not a tie-breaker bolted on — in SA it can be the whole case.

# Module 3.2 — Merger retrospectives and natural experiments

## Why look back at cleared mergers

- Merger review is prediction; **retrospectives** check the predictions against what actually happened — the only feedback loop the system has.
- The point is institutional learning: were the agency's price-effect forecasts right, were efficiencies realised, did remedies work?
- Methods: **difference-in-differences**, **event studies**, **synthetic controls**, and re-estimated **structural models** on post-merger data.
- The credible designs lean on a clean **control** — an unaffected product, region, or comparable transaction that did not experience the merger.
- Retrospectives are also a litigation asset: an expert who pre-commits a retrospective design looks far more credible than one who reverse-engineers a result.

## What the retrospective literature finds

- **Efficiency claims are often overstated** and frequently **not passed through** to consumers — the rebuttal that wins approval often does not materialise.
- **Behavioural remedies often underperform** structural ones — conduct rules get gamed, monitoring lapses, and the fix expires.
- **Some cleared mergers raised prices** — evidence that enforcement was, in places, **too lenient** at the margin.
- The asymmetry: blocked mergers can't be studied counterfactually, so the literature is built mostly from **cleared** deals — a selection caveat to state out loud.
- Practitioner upshot: discount unverified efficiency promises and favour structural relief where it is feasible.

## Natural experiments in competition policy

- **Deregulation episodes** are policy's natural experiments: **airline deregulation (1978)**, the **AT&T breakup (1984)**, banking-branching liberalisation, and **electricity restructuring**.
- **Antitrust-law changes** also serve: the **US corporate leniency program (1993)** revision sharply changed cartel-detection incentives — a clean before/after.
- Each gives a sharp, often discontinuous, change in the competitive environment that is **plausibly exogenous** to the firms studied.
- Design discipline: identify the treated and control groups, check pre-trends, and worry about confounding shocks that coincide with the policy date.
- These episodes are the empirical backbone of much of what we *know* about how concentration maps to prices.

## Worked example — Liebersohn (2017), banking

- Setting: US merger rules require **divestiture** where a merger pushes a market's HHI **up by ≥ 200 to above 1,800** — a sharp, rule-based threshold.
- Design: **difference-in-differences** comparing markets just inside vs. just outside a **500-point band around the 1,800 cutoff** — markets similar in everything but whether the rule bound.
- Result: where the rule forced more competition, banks did **more lending to larger, safer borrowers** — concentration had been constraining credit supply.
- Clean-design check: **no bunching** of mergers just below the cutoff — firms were not gaming the threshold, so the running variable is credible.
- **Why it matters:** a regulatory threshold is a gift to the empiricist — it manufactures the discontinuity a credible causal design needs.

## Worked example — Whirlpool / Maytag appliances

- Post-merger evaluation using **scanner (point-of-sale) data** across appliance categories after the Whirlpool–Maytag combination.
- Design: use **freezers as a control** category (little affected by the merger) against the treated categories to net out common shocks.
- Finding: prices for **dishwashers and dryers rose** post-merger, while the control category did not move the same way — consistent with exercised market power.
- A model retrospective: real transaction data, a defensible control, and a transparent before/after that anyone can replicate.
- **Why it matters:** ex-post evidence like this is exactly what should discipline the *next* appliance-merger prediction.

# Module 3.3 — Remedies and regulation

## Structural vs. behavioural remedies

:::: {.columns}
::: {.column width="50%"}
**Structural**

- Divestiture of a business / assets
- Clean break; little ongoing monitoring
- Preferred where feasible (US/EU default)
- Risk: divestiture package too weak to compete
:::
::: {.column width="50%"}
**Behavioural**

- Conduct rules: access, pricing, firewalls, non-discrimination
- Needs ongoing supervision
- Common in SA (Tribunal-supervised conditions)
- Risk: gaming, monitoring burden, expiry
:::
::::

## Remedy design principles

- A remedy must be **effective** (it actually cures the identified harm) **and administrable** (the Tribunal can monitor and enforce it).
- Match the remedy to the **theory of harm**: structural for horizontal overlap; access/firewall conduct rules for vertical foreclosure.
- Prefer remedies with **clear, measurable** compliance triggers over open-ended "behave well" obligations.
- Build in **monitoring**: reporting duties, a compliance trustee/monitor, and a sunset or review date.
- Recall the retrospective lesson: **behavioural remedies underperform** — use them where structure is impractical, and supervise hard.

## Access pricing: ECPR and the revenue requirement

> - **ECPR access price = direct cost + opportunity cost**
> - **Revenue requirement = opex + depreciation + (return × rate base)**

- **ECPR** (efficient component pricing rule) lets an efficient entrant in while compensating the incumbent — the benchmark in margin-squeeze remedies.
- Rate regulation builds **allowed revenue** from costs plus a fair **return on capital**; the contested number is almost always the allowed return.
- **Rate base** = the regulated asset value on which the return is earned; gold-plating it is the classic incumbent strategy.
- **Benchmarking / yardstick:** compare the regulated firm to comparable firms or regions to discipline its cost claims.

## Benchmarking and yardstick regulation

- Yardstick regulation ties a firm's allowed prices/costs to the **performance of comparators**, breaking the regulator's information dependence on the firm.
- Works best with **many comparable units** (regional utilities, water boards, port terminals) facing similar conditions.
- Methods range from simple peer **ratio benchmarks** to frontier techniques (DEA, stochastic frontier analysis) that estimate an efficient cost frontier.
- The trap: comparators must be **genuinely comparable** — adjust for scale, density, and input prices or the firm will (rightly) attack the benchmark.
- In SA, sector regulators (energy, ports/rail, telecoms) increasingly lean on benchmarking to test cost claims.

## Case — Evanston Northwestern (US)

- A consummated hospital merger found to have raised prices.
- Rather than unwind it (structural), the FTC imposed a **behavioural** remedy: **separate, independent negotiating teams** for the two hospitals when bargaining with insurers.
- The canonical behavioural hospital-merger remedy — an effective conduct fix where divestiture of an integrated hospital was impractical, at the cost of ongoing supervision.
- **Why it matters:** when structural relief is impractical, a well-designed conduct remedy can restore competition — if someone watches it.

## Tying it to SA — Tribunal-supervised conditions

- Behavioural and public-interest **conditions are the SA norm**, not the exception — the Tribunal routinely supervises conditions over years.
- Typical machinery: reporting obligations, a **compliance monitor/trustee**, defined remedies for breach, and periodic review by the Tribunal.
- The CAC's refinements (Walmart/Massmart) shape how conditions are drafted so they are enforceable, not aspirational.
- The design lesson and the SA institutional reality coincide: **a remedy is only as good as the body that can supervise it.**

# Module 3.4 — Innovation, IP, and antitrust

## SEPs and FRAND: hold-up vs hold-out

Once a standard is set, an SEP holder can hold up implementers; FRAND commitments are the check.

> **Cumulative royalty = Σ (royalty rate_i)**

- **Hold-up:** threaten an injunction post-adoption to extract supra-FRAND rates. **Hold-out:** implementers delay/underpay, free-riding on the standard.
- FRAND ("fair, reasonable and non-discriminatory") is the licensing promise made to the standard body — the legal hinge of every SEP dispute.
- *Huawei v. ZTE* (EU 2015): a **negotiation framework** — the SEP holder must offer FRAND terms before seeking an injunction; the implementer must respond in good faith.
- *Unwired Planet* (UK 2017): courts will set a **global FRAND rate**, not just a national one — a major jurisdictional escalation.

## Royalty stacking — the arithmetic

> **Cumulative royalty = Σ (royalty rate_i)**

- The problem: many SEP holders each charging a "reasonable" rate that **stacks** to an unviable total.
- Worked figure: **10 holders × 5% each = 50%** of the device price in royalties — manufacturing becomes unprofitable on any normal margin.
- Each individual demand can look defensible; the **aggregate** is what kills the product. FRAND and ND ("non-discriminatory") exist to contain the stack.
- Remedy logic: aggregate royalty caps, top-down rate-setting from a total reasonable royalty, and patent-pool licensing.

## Pay-for-delay — FTC v. Actavis

- **Reverse-payment ("pay-for-delay") settlements:** a brand pays a generic to **stay out** of the market — the brand keeps the monopoly, both split the rents.
- *FTC v. Actavis* (US 2013): **no "scope of the patent" shield.** Even within the patent's nominal scope, a reverse payment is judged under the **rule of reason**.
- The key signal: a **large, unexplained reverse payment** is itself evidence of anticompetitive intent — its size proxies the value of the avoided competition.
- Damages logic: harm ≈ (P_brand − P_generic) × volume × delay — the consumer overcharge from postponed generic entry.
- **Why it matters:** the payment's *size*, not the patent's nominal validity, became the analytical centre of gravity.

## Killer acquisitions

- **Killer acquisition:** an incumbent buys a **nascent rival to shut its pipeline**, not for synergies — pre-emptive elimination of future competition.
- Cunningham, Ederer & Ma (2021) estimate roughly **5–7% of pharma deals** are killer acquisitions — small in count, large in lost innovation.
- The test: compare **project discontinuation rates**, acquired vs. independent — acquired overlapping projects are dropped more often.
- The enforcement gap: many such deals fall **below notification thresholds** (the target has little revenue) — hence the move toward share-of-supply and call-in powers.
- Live in digital and pharma policy debates (the "**string of nascent-rival acquisitions**" theory in platform cases).

## Appellate caution — Qualcomm, both sides

- *FTC v. Qualcomm*: the district court found liability, but the **9th Circuit reversed (2021)** — no duty to deal, and the licensing practices were not shown to harm competition under §2. **Cite the reversal, not the headline.**
- EU: the Commission's **€997m exclusivity-payment fine on Qualcomm was annulled** by the General Court **(2022)** on procedural and substantive grounds.
- Lesson: the most-cited "abuse" stories were **undone on appeal** in both jurisdictions — carry the outcome, not the press release.
- This is the IP-and-dominance counterpart to the Day-1/Day-2 habit: **cite outcomes, not headlines** (Sasol, Media24, Intel, Qualcomm).

## Case — ARV / antiretroviral access (SA)

- SA competition complaints over **antiretroviral** pricing pushed originators toward **voluntary licences** expanding generic supply.
- Method mix: **medicine price data**, the **patent landscape**, and **SAHPRA registration timelines** to estimate the harm from delayed generic entry.
- Sits at the intersection of **§8 abuse, IP rights, and a pressing public-health interest** — excessive pricing and refusal-to-licence theories under SA dominance law.
- **Why it matters:** in SA the IP-vs-competition question has been about **access to life-saving medicines**, not abstract royalty rates.

# Module 3.5 — Research integrity and credibility

## The research-credibility crisis

- A wave of evidence shows much published empirical work does **not replicate** — directly threatening expert testimony built on it.
- Four interacting culprits: **publication bias** (only significant results get out), **specification searching** (p-hacking), **low statistical power**, and outright **replication failures**.
- For an expert, this is not academic hygiene — it is **cross-examination risk**: opposing counsel will probe every degree of freedom you took.
- The defence is **transparency**: pre-commit the design, disclose the specification space, and make the work reproducible end to end.
- This module is the methodological conscience of the whole course — it governs *how much to believe* any number, including your own.

## The Ioannidis PPV model

> **PPV = (1 − β) R / [ (1 − β) R + α ]**

- PPV = **positive predictive value**: the probability a "significant" finding is **true**. β = false-negative rate (so 1 − β = power); R = prior odds a tested hypothesis is true; α = significance level.
- Plug in good practice — high power, plausible prior — and an ideal field can hit **~89%** true positives.
- Plug in a **hot field** — many teams testing many long-shot hypotheses, low power — and PPV can collapse toward **~20%**: most "discoveries" are false.
- The drivers are exactly the credibility crisis: **low power + bias + a crowded field** drive PPV down.
- Practitioner reading: a single significant result from a low-powered study in a hot literature is **weak evidence**, statistics notwithstanding.

## Publication bias and the file drawer

- The **file-drawer problem:** null results sit unpublished in drawers; significant results get out — so the **published** literature over-represents positives.
- Consequence: published **effect sizes are inflated**, because only the larger, "significant" estimates clear the journal filter.
- Diagnostic: the **funnel plot** — plot effect size vs. precision; symmetry suggests no bias, a missing corner (small studies with null results) signals it.
- Tools: trim-and-fill, p-curve, and (best) **pre-registration** so the result exists regardless of significance.
- For experts: a literature that *looks* unanimous may be an artefact of the filter — weight it accordingly.

## Multiple testing — the garden of forking paths

> **P(≥1 false positive) = 1 − (1 − α)ⁿ**

- Run enough tests and a "significant" result is guaranteed by chance: at α = 0.05, **10 tests → ~40%** chance of at least one false positive; **20 tests → ~64%**.
- The **garden of forking paths**: even without formal multiple testing, flexible choices (controls, subsamples, transformations) implicitly search the space.
- Corrections: **Bonferroni** (divide α by n — conservative), Holm, or FDR control for many tests.
- The real fix: a **pre-analysis plan** that fixes the hypotheses and specifications *before* seeing outcomes, converting exploratory work into confirmatory.
- On the stand, "how many specifications did you run?" is a standard, deadly question — have the honest answer ready.

## The replication crisis in economics

- Replication efforts in economics find roughly **half** of studies fail to reproduce the original result — comparable to the better-known psychology numbers.
- Causes mirror the rest of the module: small samples, researcher degrees of freedom, weak data, and selective reporting.
- Distinguish **reproducibility** (same data + code → same numbers) from **replicability** (new data → same finding) — experts owe at least the first.
- The journals' response: mandatory **data-and-code archives** and replication packages — increasingly expected of expert reports too.
- The lesson for advocacy: build your analysis so a hostile third party can **re-run it and get your numbers** — that is the strongest credibility signal there is.

## The FTC's nine characteristics of useful analysis

- A practical checklist for credible econometric work: (1) addresses a **relevant** issue; (2) **consistent with economic theory**; (3) reflects **institutional facts**; (4) uses **appropriate data**; (5) uses **suitable techniques**; ...
- ... (6) **realistic about its limits**; (7) tests **robustness**; (8) is **replicable**; (9) **communicates clearly** to a non-technical decision-maker.
- The most common **deficiencies** in practice: **institutional disconnect** (the model ignores how the industry actually works), **data-quality** problems, and **poor communication**.
- Use it as a self-audit before filing — and as a cross-examination map of the *other* side's report.
- It rhymes with the credibility module: relevance, robustness, and replicability are the recurring themes.

## Pre-analysis plans and open science

- **Pre-analysis plans (PAPs):** register hypotheses, specifications, and the analysis path *before* seeing the data — the antidote to forking paths and p-hacking.
- **Open data and code:** post the cleaning and analysis scripts (GitHub) and the data (Dataverse) so the result is independently checkable.
- **Registration**: clinical-trial-style registries (AEA RCT Registry, OSF) timestamp the design and deter selective reporting.
- A reproducible **lockfile** (renv / packages pinned) means the analysis runs identically years later — when the matter is finally heard.
- These practices are slowly migrating from journals into **agency and tribunal** expectations of expert work.

## Reproducibility IS advocacy

- A clean **/data /scripts /reports** bundle with a **lockfile** is not just good science — it is your **cross-examination defence**.
- If the whole pipeline runs from raw data to final exhibit with one command, you can answer "did you cherry-pick?" with a demonstration, not a denial.
- **Run the sensitivity analysis before opposing counsel does** — disclose the specifications that don't change the answer, and explain the ones that do.
- A reproducible report lets you **update fast** when new data arrives mid-matter — re-run, don't rebuild.
- The single highest-leverage habit in expert work: make your conclusion **mechanically reproducible** from the inputs.

# Module 3.6 — Litigation practice and expert evidence

## Getting expert evidence admitted — Daubert

- **US — the Daubert gate.** The judge is a gatekeeper; four (non-exclusive) factors: **testability/falsifiability**, known/potential **error rate**, **peer review/publication**, and **general acceptance** in the field.
- Build the report to satisfy each: state the method's error properties, cite the peer-reviewed basis, and use accepted techniques.
- *Kumho Tire* extended Daubert beyond "scientific" to **all expert** testimony, including economics.
- The gate excludes **ipse dixit** — "trust me, I'm an expert" — so every step from data to conclusion must be defensible on its own.
- Contrast with the more flexible **weight-not-admissibility** approach of the SA Tribunal and UK/Australian courts (next slide).

## Class certification — Dukes and Comcast

- *Wal-Mart v. Dukes* (2011): **commonality** requires a common *question* whose answer **drives the resolution** of the class's claims — not just a shared characteristic.
- *Comcast v. Behrend* (2013): the **damages model must match the theory of liability** and be capable of measurement on a **class-wide** basis.
- Implication for experts: a **common-impact** regression must show the violation affected the class **in common**, and the damages methodology must be **common** to the class — not a patchwork of individual inquiries.
- A mismatch between the liability theory and the damages model is now a **certification-killer** — align them from the start.
- This is where statistics meets procedure: certification can decide a case before merits are ever reached.

## SA & concurrent expert evidence ("hot-tubbing")

- **Concurrent expert evidence** ("hot-tubbing", from Australia, used in the UK and SA): opposing experts are **sworn together** and questioned side by side.
- Before the hearing they file a **joint statement** itemising points of **agreement and disagreement** — narrowing the real dispute for the tribunal.
- The **SA Tribunal's flexible procedure** favours this over a rigid Daubert-style admissibility gate: weight, not exclusion, and direct expert-to-expert testing.
- Advantage: the tribunal hears the genuine disagreement in real time, not filtered through advocates — and experts are disciplined by a peer in the box.
- Prepare differently: you must be able to **defend every choice live**, against a knowledgeable opponent, on the record.

## Three damages models — and how to defend them

- **Before/after:** compare the violation period to a clean **benchmark window** for the same market — validate that the break is the violation, not a shock.
- **Yardstick:** compare the affected market to an **unaffected comparator** market — defend the comparability of the yardstick.
- **Difference-in-differences:** changes over time, treated vs. untreated — the **workhorse**, with **clustered SEs** (or randomisation inference for few clusters).
- Each needs an explicit **counterfactual** ("but-for" world); the fight is always over which benchmark is clean.
- Pair the point estimate with its **confidence interval** — the tribunal is buying the uncertainty, not just the number.

## The expert report — structure walkthrough

- Real reports follow a stable spine (cf. **Farber** in *Moussouris v. Microsoft*; **Card** in the Harvard admissions case):
- **Qualifications → Assignment → Executive summary → Institutional background & data → Methodology & variable selection → Analyses → Damages → Rebuttal.**
- **Variable selection** is where reports live or die: avoid **"tainted" controls** — variables that are themselves products of the challenged conduct (don't control away your own effect).
- The **executive summary** must be readable by a generalist judge; the **methodology** must be replicable by a hostile economist — write to both audiences.
- The **rebuttal** section pre-empts the other side: address the obvious robustness attacks before they are made.

## Significance and treble damages — the checkable close

> - **t = coefficient / standard error**
> - **US exposure = 3 × (overcharge rate × commerce)**

- t = 2.5 / 0.5 = **5.0** → beyond the 1% critical value (~2.58) → common impact is statistically significant.
- Commerce $200m at a 15% overcharge → $30m single → **$90m trebled** (US, automatic treble for Sherman Act violations).
- In **SA**, follow-on civil damages recover actual loss (≈ single damages) **without** the US multiplier — so the same overcharge implies very different exposure by forum.
- Always pair the point estimate with its uncertainty — the tribunal is buying the **confidence interval**, not just the number.

# Capstone — a hospital merger, end to end

## Capstone data (supplied)

| Hospital | Discharges | Share | Price | Margin |
|:--|:--|:--|:--|:--|
| A | 30,000 | 30% | \$12,000 | 40% |
| B | 20,000 | 20% | \$11,000 | 35% |
| C | 25,000 | 25% | — | — |
| D | 15,000 | 15% | — | — |
| E | 6,000 | 6% | — | — |
| F | 4,000 | 4% | — | — |

Diversion: A→B = 0.32, A→C = 0.30, B→A = 0.45. Total = 100,000. **A and B propose to merge.**

## Capstone — the seven steps

1. Market definition — patient-flow / critical-loss reasoning.
2. Concentration — HHI, post-HHI, ΔHHI.
3. Diversion — patient-flow vs. share-proportional.
4. Unilateral effects — GUPPI for A and for B.
5. Bargaining — Nash-in-Nash: how the merger shifts the disagreement payoff (qualitative).
6. Retrospective design — pre-commit a DiD with leads/lags.
7. Recommendation — translate to the legal standard (US Clayton §7 / EU SIEC / SA §12 + §12A(3)).

## Capstone — checkable answers (steps 2–4)

> - **HHI: pre 2,202 → post 3,402,  ΔHHI = 1,200**
> - **GUPPI_A = 0.32 × 0.35 × (11/12) = 10.3%**
> - **GUPPI_B = 0.45 × 0.40 × (12/11) = 19.6%**

- ΔHHI = 2 × 30 × 20 = 1,200; post-HHI well above 1,800 → structural presumption (either guidelines vintage).
- Both GUPPIs exceed the ~5% screen; B's is larger because A is the bigger, higher-margin recapture target.
- Share-proportional diversion A→B = 20/70 = 0.286 < supplied 0.32 → A and B are **closer** substitutes than shares imply.

## Capstone — verify the HHI in R

```{r}
source("program/R/helpers.R")
hosp <- data.frame(
  firm  = c("A","B","C","D","E","F"),
  share = c(.30,.20,.25,.15,.06,.04))
calc_hhi_change(hosp, merging_firms = c("A","B"))
#> pre_hhi 2202 ; post_hhi 3402 ; delta_hhi 1200
```

**Point at:** the helper reproduces the hand arithmetic exactly — pre 2,202, post 3,402, Δ 1,200. **Always tie the code back to the checkable number.**

## Capstone deliverable — the memo

- Write a **2–3 page recommendation memo** on the A/B hospital merger.
- **Required computations:** HHI / ΔHHI, both GUPPIs (show the arithmetic).
- **Required reasoning:** define the market, classify the unilateral-effects concern, address one efficiency/entry rebuttal.
- **Required framing:** choose ONE jurisdiction; state its legal test; recommend **clear / condition / prohibit**. For SA, address **§12A(3)** explicitly.
- *Worked answers and grading guide in `teaching/exercises/day3_exercises.md`.*

## Course close — the habits to keep

1. **Institutions first:** know the forum's test before you pick a method.
2. **The evidence triad:** never rest a conclusion on one leg — empirics, documents, judgment together.
3. **Screen cheap, quantify dear:** GUPPI before simulation, a variance screen before a full damages model.
4. **Cite outcomes, not headlines:** carry the SA and EU appellate history (Sasol, Media24, Intel, Qualcomm).
5. **Reproducibility is advocacy** — a one-command pipeline from raw data to exhibit is your best cross-examination defence.
6. In South Africa, **public interest can be the whole case** — and a remedy is only as good as the body that can supervise it.
